---
title: Spec-files for differential analysis
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spec-files for differential analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
author: Gavin Kelly
---

## Analysis Philosophy

This is a wrapper round DESeq2 which standardises the process of
generating `DESeq2::results` objects for a specifiable set of
comparisons. Each comparison is done in the context of a model (a
specification of what covariates need to be accounted for in
predicting the expression of any transcript). And each model is
formulated in a sample-set (which is usually all the samples, but can
be changed to drop samples, or look at a stratum in an isolated way).

There are three classes of input to the analysis pipeline: the raw
count files; the experiment table; and an analysis specification. The
count files should be stored in a folder
`extdata/alignmentID/genes.results/` relative to the analysis home
directory, where `alignmentID` is a freeform name that identifies the
alignment settings and choices (so could just be 'nfcore', or a
species name, or a version number of a reference ...).

The experiment table is called `extdata/metadata_alignmentID.csv` (to
match with the choice of the folder name above) and the first column
should be called `ID` and identifies the sample that a row is
describing, so if `extadata/nfcore/genes.results/A1.genes.results` is
sample A1 that is a control, then `extdata/metadata_nfcore.csv` should
start

```
ID,treatment,...
A1,control,...
```

Other columns should be used to entirely describe the experimental
state of the sample that might have an influence on expression. There
is a convention that the first column _after_ `ID` which takes on
unduplicated values throughout the rows is to be used as the sample
label in plots (e.g. a `samplename` column 2.) If there is no such
column, the `ID` column will be used.  In either case, the label will be
shorn of any common prefix, so if the IDs are SAMP9999A1
... SAMP999A10, and no other unduplicated column is stored, then the
samples will be labelled 1...10).


The plan as to what analyses are to be carried out is stored in
a `.spec` file, and there can be as many of these as you want - so you
could have an `overview.spec` file that tried out a handful of
approaches, and a `publication.spec` that recorded the chosen
approach. Each `.spec` file is in fact an R script (and so should be
treated with as much caution, security-wise, as any other script), but
is expected to be of a certain form, the first criterion is that it
should start and end as follows:

```{r, eval = FALSE}
specification(
...
)
```

### Sample sets
The highest level of any standardised analysis is a choice as to what
samples are to be analysed together. Often this is trivial, and the
only sensible analysis is to look at all the samples together. But
even in the simplest experiment, it might be observed for example that
one sample is an outlier and we'd like to examine the effect of
omitting or retaining this sample. We can accomplish this easily:

```{r, eval = FALSE}
specification(
  sample_sets =list( 
    all = sample_set(
      subset = TRUE,
      name="All Samples",# an optional short name
      description="All samples included in the analysis"# an optional longer description
    ),
    no_outlier=sample_set(
    subset = sample_id != "id_of_bad_sample"
    )
  )
)
```

So a collection of samples is called a `sample_set`, and we have two
(a `list`) of them here. The `subset` expression of a sample\_set is
evaluated in the context of the `colData`, so the first one always
evaluates to `TRUE`, denoted that we include all samples; the second
one evaluates to `FALSE` for a particular sample (assuming we have a
sample\_id column in our `colData`), dropping that sample.

Other reasons we may want to subset our samples into different
configurations is: if there are two very different cohorts of samples
where transcripts are likely to have substantially different noise
profiles and we want to avoid contaminating one set of samples with
the estimates of dispersion that hold for the other set. Of course it
then becomes impossible to _statistically test_ between the two
cohorts, but we can have a third set that still combines them:

```{r, eval = FALSE}
specification(
  sample_sets =list( 
    all = sample_set(
      subset = TRUE
    ),
    tumour=sample_set(
      subset = condition == "tumour"
    ),
    normal=sample_set(
      subset = condition == "normal"
    )

  )
)
```

It's possible for `sample_set`s to inherit subsets from the
`specification` level.  We discuss this concept [later on](#cascade), but a simple
example should suffice to illustrate how we can refine a 'global' subset:

```{r, eval = FALSE}
specification(
  subset = ! sample_id %in% c("bad1","bad2"), # get rid of two samples that failed qc
  sample_sets =list( 
    all_good = sample_set(
      ... # No need for a subset here
    ),
    tumour=sample_set(
      subset = condition == "tumour" # this will get 'AND'ed with the top-level `subset`
    ),
    normal=sample_set(
      subset = condition == "normal" # as will this.
    )

  )
)
```

There's also a `influential_samples` field that takes the same form.
This is to allow situations where you want to focus the statistical
testing to a subset of samples but not limit the visualisations of the
differential results to those samples. So in the above example, the
differential heatmaps of the tumour samples wouldn't show the
behaviour in the normals. Setting `subset = TRUE` and supplying the
restriction of the test as `influential_samples = condition ==
"tumour"` would visualise just the differential genes in the tumour
subset (same number of genes/rows), but show their expression in all
samples (larger number of samples/columns).  Samples excluded by an
`influential_samples` predicate will __not__ play a role in any of the
following:

 - Contributing to the calculation of gene variance when selecting
   variable genes for exploratory heatmaps and cluster centroids
 - Calculating the gene:gene distances for the heatmaps
 - Calculating the direction of the principal components
 - Any part of the DESeq2 processing

The conditions on which we `subset` must be defined in the experiment
table. It is possible to transform some of the predictors (columns of
`colData`) using the other main component of a `sample_set`, the
`transform`:

#### Transform

Each specification can have a single `transform` which is a
`dplyr::mutate` clause that will operate on the sample metadata (but
with no need for the initial data object):

```{r, eval = FALSE}
specification(
    transform=mutate(
               treatement=relevel(factor(treatment), "control"),
               patient_id=factor(patient_id),
           number_of_visits=table(patient_id)[as.character(patient_id)],
           .include=number_of_visits>1
    ),...
```

Notice that we change/create various covariates within a single
`mutate` - having multiple `transform`s is a mistake as only one will
be applied.  As mentioned above, `subset` only operates on covariates
prior to the transform being applied - if we want to utilise
`transform`ed covariates in the subset, the special covariate
`.include` should be generated in the `transform`, and the colData
will be post-processed to only include samples for which that variable
is `TRUE`.

Also, each `sample_set` can have it's own `transform`. In this case,
the individual `sample_set`'s version will take complete precedence:
the global `transform` will not be applied at all to `sample_set`s
with their own `transform`, and it will simply be used in
`sample_set`s which don't have their own `transform`.


#### Extra Assays

In some situations it is necessary to add extra assays that represent
additional pre-processing steps - for instance in strongly paired data
(for example where each biological sample is split into, say,
cytoplasmic and nuclear subsamples) it might be preferable to look at
one of the pair normalised against it's pair-mate. This is purely for
visualisation purposes: for differential analysis it is always worth
sticking with the primary assay (counts). This can be achieved by a
setting `extra_assays` in a dataset as a list of `extra_assay`
calls:

```
  subset = ...,
  extra_assays = list(
      to_nuclear=extra_assay(
      method="normalise", from="vst",
      compartment="nuclear",
      design=~compartment:treatment:replicate,
      recentre=FALSE, rescale=FALSE)
  ),
  visualise = compartment != "nuclear",
  ...
```

Currently there are only two methods implemented: `normalise` and
`identity`. `normalise` takes:

 - a `design` argument that identifies the groups within which to
normalise
 - an assay `from` to use as the base assay
 - options to `recentre` and `rescale` the data (these default to `TRUE`
   for our visualisations)
 - other arguments set the baseline value against which everything in
   a group (defined by the `design`) will be normalised to

This clearly leaves, in this case, the _nuclear_ samples normalised to
themselves (so identically zero across all genes). To avoid this
disrupting the visual picture, we have a third dataset-specific
predicate `visualise` (alongside `subset` and `influential_samples`)
that will remove these samples from plots (this is optional, but if
you dislike seeing a primarily white set of columns in your heatmaps,
this is the way to eradicate them.)

The `method="identity"` allows us to create assays identical to the
`from` assay, but allow us to switch off the centring and/or scaling.

We'll see how to use the extra assays in the [plotting
section](#profile-formulae). But a reminder: these extra assays are
ignored in the differential testing, as the inferences and estimations
are always done on the primary assay. You'll need to model the pairing
explicitly.

The extra assays accumulate, so you can have a later `to_T0` that
could be generated `from="to_nulcear"`, for instance.

### Models

DESeq2 uses R's standard one-sided formulas to inform the analysis as
to which predictors to use in fitting a negative-binomial model of
expression counts. The `colData` slot should contain all the potential
predictors, and the `design` slot of a DESeq2 object encodes a
considered choice of which are relevant to the biological
question. Obviously, if the question is to find which genes are
differentially expressed between two treatments, then a predictor
annotating which treatment a sample is subject to is necessary. But
quite often there will be other covariates need to be accounted for,
such as batch; or perhaps there's a subtler question of finding genes
that respond differently to treatment depending on the genotype of the
sample.

We may want to consider alternative models, for instance whether or
not to include a batch effect: if the batch effect is negligible we
won't want to use up degrees of freedom (ie power) estimating it, but
if it's substantial then we want to make our estimates more precise by
accounting for it. So we allow, in any given sample_set, for there to
be multiple models:

```{r, eval = FALSE}
specification(
  sample_sets =list( 
    all = sample_set(
      subset = TRUE,
      models = list(
        accurate = model(
          design = ~treatment + batch
        ),
        powerful = model(
          design = ~treatment,
      name = "Just Treatment",
      description = "Account for treatment, ignore batch - if there is a batch effect it'll increase the noise, though"
        )
      )
    )
  )
)
```

So once again the plural 'models' is a list of individual `model`
objects, which for the moment we've just specified by their `design`,
exactly as we would for a `DESeq` object. Once we have a model, we are
then in a position to test the significance of contrasts (e.g.
individual log fold-changes between conditions) and terms in the model
(e.g whether batch effect is necessary or not):

#### Universal models

Sometimes all/some models should apply to all `sample_sets` in a spec
file. To save typing, it's possible to elevate the models to the
top-level within a `specification`, so for example:

```{r, eval = FALSE}
specification(
  models = list(
          powerful = model(
          design = ~treatment
        )
  ),
  sample_sets =list( 
    all = sample_set(
      subset = TRUE,
      models = list(
        accurate = model(
          design = ~treatment + batch
        ),
      )
    ),
    batch1 = sample_set(
      subset = batch=="B1"
      )
    )    
  )
)
```

will result in the `all` dataset having two models (the universal one
from the top of the spec file, and a specific one), whereas the
`batch` dataset will have only the universal one.

There are other similar patterns for other parts of the spec file,
that certain properties 'cascade' down to other parts of the hierarchy
- see [here](#cascade) for a list of behaviours.

### Comparisons

DESeq2 has multiple ways of specifying a contrast. All of these are
supported, and again as 'comparisons' is plural, we can have a list of
them, and have the pipeline loop through all of them for the parent
model:

```{r, eval = FALSE}
specification(
  sample_sets =list( 
    all = sample_set(
      subset = TRUE,
      models = list(
        accurate = model(
          design = ~treatment + batch,
          comparisons = list(
            comp1 = "treat_vs_ctrl",
            comp2 = c("treatment", "treat", "ctrl"),
            comp3 = list(c("treat_vs_ctrl"), c("vehicle_vs_ctrl"), listValues=c(1,-1))
            comp4 = c(1,0,-0.5,-0.5),
            comp5 = ~batch,
        comp6 = comparison(c("treatment", "treat", "ctrl"), name="Treat-control", description="Treatment vs Control")
          )
        )
      )
    )
  )
)
```

So all of the traditional ways of specifying a contrast are
represented, respectively:

- by a single character, traditionally done in DESeq2 with
  `results(..., name="treat_vs_ctrl")`
  
- A character triplet

- A two-component list of characters indicating the numerator and
  denominator (with possible `listValues` set)
  
- A numeric vector.

- A formula specifying the `reduced` term for carrying out Anova via
  a likelihood ratio test.

You are referred to the DESeq2 reference manual for further
details. They can also be wrapped in a `comparison` call, to add extra
information about the comparison, that might get used in the report. But one
advantage of our wrapper is that it allows a simple enumeration of
complex comparisons using machinery from the `emmeans` package:

#### Automatic comparison enumeration

We often find ourselves copying a contrast multiple times, with slight
changes to elicit comparisons between different conditions. For
example if our samples have been subject to three different treatment
regimes, and a control, so treatment={_control_, _vehicle_,
_standard_, _novel\_treatment_}, and we wanted to compare everything to
_control_, then we'd need to hand-write three comparisons. Instead we
can now write

```{r, eval = FALSE}
specification(
  sample_sets =list( 
    all = sample_set(
      subset = TRUE,
      models = list(
        accurate = model(
          design = ~treatment + batch,
          comparisons = list(
            mult_comp(trt.vs.ctrl ~ treatment, ref="control")
          )
        )
      )
    )
  )
)
```

where the `trt.vs.ctrl` is a 'keyword' instructing as to which comparisons to
carry out - in this case every value of `treatment` against the ones
labelled "control". There are a number of other helpful keywords:

- `revpairwise` is the most 'verbose', in that it will generate every
  pair of comparisons, in this case there are six combinations, and it
  can go up rapidly. Due to a slightly annoying convention, `pairwise`
  choses an unintuitive ordering that would result in comparisons such
  as 'control vs novel\_treatment' (and a positive fold change would mean
  higher expression in the control than the novel treatment), and
  `revpairwise` has the more conventional ordering of
  'novel\_treatment vs control' and it's more intuitive positive log
  fold-change indicating higher expression in the treated to the
  control. This assumes the levels of the factor are given in the
  conventional order of the previous paragraph.
  
- `consec` takes adjacent pairs along the levels, so again quite a
  natural 'vehicle vs control', 'standard vs vehicle' and finally
  'novel\_treatment vs standard'
  
- `trt.vs.ctrl` chooses one baseline (which we can select with an
  additional `ref` argument, rather than having to `relevel` the data)
  and compares everything else against it.
  
- `mean_chg` looks to see if splitting the levels at any particlar
  point reveals a change between levels before vs after that
  breakpoint, so amongst other things would test if the average of the
  _vehicle_ and _control_ was different from the average of the
  _standard_ and _novel_ treatment.
  
- `eff` tests whether each individual level is different from the
  average of the remaining levels.

These latter two still estimate the _standard error_ in the individual
experimental groups, so the significance is still assessed relative to
replicate variability. This is different from 'relabelling' the
samples (via a `transform`) so that e.g. there was one group called
"both controls" and another called "both treatments", where the
standard error would include not only the 'true' replicate variability
but also the discrepancy between the former experimental groups that
have now been pooled.

When we don't have an interaction term in our model, this is
everything we'd ever need, as for example the difference between
'novel\_treatment' and 'control' is by construction (of the model
formula) independent of which batch we're in.

The _keywords_ are inherited from the [emmeans
package](https://rvlenth.github.io/emmeans/articles/AQuickStart.html),
so that is a good place for extra background material on their
meaning.

One thing to note is that `emmeans` has a different way of handling
continuous covariates, so to specifiy a test of difference of slopes,
it's necessary to request that functionality by providing a `var`
argument to the `mult_comp` call, e.g.:

```{r, eval = FALSE}
...
#design= ~ Infection * time + Participant_in_infection:infection
mult_comp(revpairwise ~ Infection, var="time")
```

is the correct formulation within the `mult_comp` framework to
estimate the per-participant rate of change of expression over `time`,
and then see if that differs between pairs of groups of patients that
are distinguished by their `Infection`.

#### Interaction terms

When our model includes an interaction between two main effects, say
between treatment and genotype, then the magnitude of difference
between two treatments can depend on the genotype. This leads to two
types of follow-up question: stratifying the data by one factor, and
then examining the effect of the other factor in each individual
stratum, or; examining the interaction itself, so looking to see if
the magnitude of effect is different in one stratum than another.

To make this more concrete for the genotype x treatment case, we could
ask: which genes are responding to treatment in the KO; which genes
are responding to treatment in the WT; which genes are differential
between WT and KO in the untreated. These are all questions of the
first type, and we can automatically generate their contrasts by:

```{r, eval = FALSE}
specification(
  sample_sets =list( 
    all = sample_set(
      subset = TRUE,
      models = list(
        accurate = model(
          design = ~treatment *  genotype,
          comparisons = list(
            mult_comp(revpairwise ~ treatment | genotype)
            mult_comp(revpairwise ~ genotype  | treatment)
          )
        )
      )
    )
  )
)
```

so the `|` symbol is to be read as "stratified by" (or "conditioned
on"). The first `mult_comp` is stratifying the genotype, and will
generate the pairwise comparisons within treatment, separately for
each different genotype. The second is the dual question: For each
treatment, find the genes that are 'markers of genotype' within that
treatment group.

The second type of question, where we want to investigate a
'difference of differences', is achieved by the following grammar:

```{r, eval = FALSE}
mult_comp(revpairwise ~ treatment + genotype, interaction=TRUE)
```

The terms on the right hand side of the formula enumerate the
predictors we're concerned in, and the left hand side specifies the
contrasts that are going to be considered (so any of the keywords from
the previous section) - there needs to be just one, applied to all
factors. If we omit the `interaction=TRUE` argument then it will
enumerate all possible combinations of treatment and genotype, and
then contrast every combination against every other combination -
probably not what is wanted (e.g. treated KO vs untreated WT); keeping
the `interaction=TRUE` term is more usual, and builds a contrast that
represents the differences of differences, e.g. treatment effect in KO
- treatment effect in WT (where treatment effect in KO is treated KO -
untreated KO.)

The above is probably quite intimidating, so please ask for help on
this - I'll try to write something a bit more friendly.

#### Caveat

With the ease of testing so many hypotheses comes a risk of fishing
the data, resulting in spurious statistical associations. Strictly
speaking, anything beyond one comparison on the data requires a more
conservative approach than is achieved even with the traditional
control for FDR to account for the multiplicity of genes.

The ideal situation is the tests are prespecified. If that is the
case, then we should really run an anova to confirm that a change
exists somewhere, and then do at most n-1 comparisons
(where n is the number of levels in the factor of interest, so at most
three comparisons for our four-treatment example) and choose
transcripts where both the anova and contrast are significant. There's
a further technical constraint that the comparisons should be
independent but that is often ignored. If there are more comparisons
of interest, then the p-values won't be controlled at the correct
rate.

Often, the ideal situation is not achieved, and an exploratory
approach is requested. Being strict one should immediately correct for
the abundance of hypotheses in this situation. Pragmatically and
traditionally we neither adjust in this situation nor consider
possible adjusting in the pre-specificied case - this means that the
p-values aren't correctly calibrated, and one of the reasons why I
discourage presenting them in any quantitative form as part of the
results.

Also remember the "Table Two Fallacy" - any inference made about
non-randomly-allocated groups is an observational finding, and doesn't
carry the same rigour as inferences where treatment was randomly
allocated. Observations of associations can easily be spurious and
non-causal.

### Rank deficiency 

One of the most common hurdles in complex designs is the
rank-deficiency problem. The way the experiment has been described can
result in the analysis struggling to identify an estimate for a
particular experimental condition. This happens for one of three
reasons:

- The design has a fatal flaw. If all your treated samples are in one
  batch, and all your control samples in another batch, you cannot
  estimate the extent on how these individually influence expression.
  
- Some conditions have no samples. For example, in our genotype x
  treatment experiment, we might not have a 'vehicle' treatment in one
  of our genotypes.
  
- There is a nesting indicative of repeated measurements. Individuals
  might have submitted multiple samples (perhaps a time-course), and
  also be members of larger groups (their disease status, which is
  assumed constant throughout the time-course).
  
The latter two are remediable through some tweaks to the inputs of
DESeq2, the first will need the advice of a statistician to see if
anything is recoverable and at what compromise.

The 'missing condition' case is easily remedied by supplying the
option `drop_unsupported_combinations=TRUE` to any _model_ you want
this to apply to. The 'nesting' case normally requires a tweak to the
way factors are coded, and we have made it easy to achieve this
through a `transform` option to any _sample\_set_.  Both options are
illustrated below. Note, it is important that there be only **one**
`transform=mutate` component per sample_set (if you want to
introduce/transform multiple derived covariates, then separate them by
commas _within_ the `mutate` expression:

```{r, eval = FALSE}
specification(
  sample_sets =list( 
    all = sample_set(
      subset = TRUE,
      transform = mutate(unit = recode_within(unit, genotype),
                         time = as.numeric(sub("^t", "", timepoint))),
      models = list(
        nested = model(
          design = ~timepoint *  genotype + person:genotype,
          comparisons = list(
            mult_comp(revpairwise ~ timepoint + genotype, interaction=TRUE)
          ),
          drop_unsupported_combinations=TRUE
        )
      )
    )
  )
)
```

The 'missing condition' case is achieved by just the one line; the
nesting is more complex - we firstly have to have a 'unit' column in
our `colData` that identifies which biological unit (person, animal,
plate) a sample belongs to. Then the `transform` line calls a `recode_
within` function that tells us that unit is nested within genotype
here. And we alter the `design` in accordance with the DESeq2 vignette
on within- and between- group comparisons.

The `transform` statement can be used to adjust any of the `colData`
columns in-line. It's a question of personal preference whether
'mistakes' in the original experiment table should be fixed at source
(which allows the scientist to review changes manually), or via a
transformation specified in the spec file (which keeps an entirely
faithful record of what is done).

### Other functionality

#### Profile formulae

In the exploratory plots, we present PC and dendodrograms. Primarily
these are of the normalised counts without any further processing, but
there is a mechanism for removing the effect of known covariates prior
to the calculation of variable genes (for heatmaps) and rotations (for
PCs). This is useful in cases where the dominant transcriptional
variation might be driven by "nuisance" covariates, such as batch or
patient, and so the simply normalised plots focus on experimentally
uninteresting features.

This is **not** the same as looking at an unadjusted PCA plot,
noticing that the main clusters appear to relate to batch, and the
then noticing that, say, timepoint rather than treatment is
determining the subclusters within the large clusters, and from this
saying that timepoint is more "important" than treatment. That is a
fallacy, as the original dimension reduction is looking in the "wrong"
direction. Recalculating the PCs having removed the nuisance effect is
more valid.

By default, no additional plots are generated - this is a change from
previous `qc_formulae` versious, because we were starting to see more
complex designs where the default option (producing all colours x
covariate plots) was becoming unmanageably verbose. Now it is
necessary to add descriptions of the profiles to be plotted explicitly:

```{r, eval = FALSE}
    all = sample_set(
      subset = TRUE,
      models = list(
        nested = model(
          design = ~treatment * time + batch,
          profile_plots = list(
            aes(x=time, colour=batch, shape=treatment)  ~ treatment + time + timepoint * time + batch,
            aes(x=time, colour=treatment)  ~ treatment + time + timepoint * time,
            aes(y=vs_t0, x=time, colour=treatment) ~ . - batch
          )
        )
      )
    )
```

Everything on the LHS is used as an aesthetic mapping for any given
plot, and the model on the right says which covariates's effects are
kept in the 'normalised' data: anything that is omitted is effectively
normalised/regressed-out .

In the first element of the `profile_plots` list, the RHS of the
formula is algebraically identical to the `design`, so no effects will
be regressed out. But plots will be drawn so that time is on the
horizontal axis, expression will be on the vertical axes - and samples
will be coloured by batch, and shaped according to which treatment
group they were in.

The second element will produce a different set of plots, this time
we've regressed out (corrected-for) batch, so the PCA and heatmaps
will be different from those on raw vst data. Note, for the
two-dimensional PC plots, we need both the `x` and `y` aesthetics to
represent the projection, so the `time` aspect of this data would not
be represented in the visualisation: to get round this, we
automatically generate a plot where we use the `colour` aesthetic to
represent the overwritten `x`.

The third example illustrates the standard use of the `.` placeholder
in R formulae to represent the 'existing' model.  Here, the `.`
represents the design, and the `-` is the standard Wilkinson notation
to _remove_ an effect, so the third example is a different way of
phrasing the second instance: regress out the effect of batch.

The third example also illustrates how to access `extra_assay` as
defined [here](#extra-assays) - simply setting the `y` aesthetic to
map to the name of the extra assay will replace the default assay
(vst-transformed normalised counts) in the plots.


#### Clustering

We use the `bluster` package to cluster samples and features in our
heatmaps, via the `gene_clust` and `sample_clust` settings. The defaults are:

```
settings( ...
  gene_clust     = bluster::HclustParam(cut.params = list(k = 12)),   ## When we need to chose clusters of genes, how many?
  sample_clust   = bluster::HclustParam(metric="pearson"),...
  )
```

The `k` there determines the number of feature clusters that a heatmap
is broken into (which are then used in the profile-plots, so we'll be
plotting e.g. 12 centroids).  The metric can be any `dist` metric, or
`cor` method.


#### Sample swaps

If there have been a pair of samples swapped, this can be corrected in
the spec file.

```{r, eval = FALSE}
    all = sample_set(
      subset = TRUE,
      swap=list("A"="B")
```

This will swap the _metadata_ of sample with ID `A` with that of
sample ID `B`.  **Every** column of the metadata will be exchanged:
this is generally what is required (remember that `ID` is what is used
to connect the metadata to the correct quantified counts file):

```
ID Name Treatment Batch
A  T1   Treated   B1
B  C1   Control   B1
```

correctly becomes

```
ID Name Treatment Batch
A  C1   Control   B1
B  T1   Treated   B1
```

Tellings us that the contents of nfcore file `A.genes.results` are
from a control sample from batch B1. But if somehow the swap had not
affected the `Name`, and the situation that needed correcting was

```
ID Name Treatment Batch
A  C1   Treated   B1
B  T1   Control   B1
```

then you might want to consider fixing the anomaly manually (perhaps
via a `mutate`) as swapping around that table will result in the
somewhat confusingly "Named"

```
ID Name Treatment Batch
A  T1   Control   B1
B  C1   Treated   B1
```

(the correct association between A and "Control Batch 1" has been
made, but the name is now "off" - unimportant for the analysis, but
counter-intuitive).

#### Technical Replicates

Sometimes a scientist has done purely technical replicates. Ideally
these should be concatenated at the nfcore level, but sometimes they
get kept separate. We can correct for this in the spec file, for
example:

```{r, eval = FALSE}
    all = sample_set(
      subset = TRUE,
      collapse=~treatment + biological_replicate + timepoint + batch,
      ...
```

The collapse formula should identify the distinct experimental groups,
so in the above formula we might expect that there are multiple
samples (rows in the experiment_table) with the same values for each
of those four variables; so there may be three rows with
`treatment=control`, `biological_replicate=3`, `timepoint=7hr` and
`batch=1` corresponding to three technical replicates. Ensure you have
enough terms in your model to distinguish all experimental conditions
and true replicates.

Keeping technical replicates in (without a `collapse`) is almost
always bad, as the standard errors are badly miscalculated.

#### Normalisation strategies

We can fine-tune which genes are analysed and which genes are used to
calculate size factors:

```{r, eval = FALSE}
    all = sample_set(
      subset = TRUE,
      controlGenes=expression(grepl("^ercc", ID)),
      subsetGenes=expression(symbol!=""),
      ...
```

`controlGenes` and `subsetGenes` both take an R expression that should
evaluate to a boolean of the featureData of the dataset
(traditionally, `symbol` and `entrez` get set, and `ID` is my
convention for accessing the row.names (probably the ensembl ID). The
size factors are calculated on all rows that pass the controlGenes
filter, and then the subset filter is applied to all the original rows
(so one can calculate on the spike-ins, and then drop them out.)

For `SummarizedExperiment` objects (e.g. proteomics or metabolomics
analysis, but not _yet_ DESeq2 RNASeq analyses), we have the
capability to stratify the normalisation process: adding a `strata = ~
batch` to a `sample_set` definition will separately calculate the
imputation and vst transformations for each stratum defined by the formula.

#### External genelists

If there are lists of _a priori_ interesting genes, the analysis can
be informed of these via an additional csv file, the location of which
is supplied with a line in the spec-file:

```{r, eval = FALSE}
  external_list="extdata/lists.csv"
```

This can either be an element of the `settings`, or the
`specification` or any individual `sample_set`, and is expected to be of the form:


|ID  | Comparison | Cluster |
|----|------------|---------|
|S1  | Pos Control|  C1     |
|S2  | Pos Control|  C2     |
|S3  | Pos Control|  C3     |
|S4  | Public     |  C4     |
|S5  | Public     |  C4     |

`ID` needs to match up with the `row.names` of your assay,
`Comparison` will determine which genes are grouped together in a plot
(given the same colour in scatter plots; given their own plot in a
heatmap or profile-plot), and the `Cluster` is only used in the
profile plots: genes with the same Cluster within a Comparison will be
averaged, to provide a single trace - if Cluster is omitted, each
symbol will be plotted as its own singleton cluster.

#### Imputation

For `limma`-like analyses, (see [this vignette on running just the
'differential' phase of the analysis](isolated.html) ), it is possible
to define different imputation strategies:

```{r, eval = FALSE}
sample_set(...,
    normalise = "vsn",
    impute    = list(method="MinProb", q=0.01),
    ...
)
```

We will better document this shortly.


### Settings

Other than the overall modelling strategy, it's important to be able
to tune the analysis by choosing the parameters behind the
algorithms. It's possible to set any parameters that the core analysis
script has made available to you through the `settings` top level
option:

```{r, eval = FALSE}
specification(
  sample_sets =list( 
  ...
  ),
 settings=settings(     ## analysis parameters
   alpha          = 0.05,    ## p-value cutoff
   lfcThreshold   = 0,       ## abs lfc threshold
   baseMeanMin    = 0,       ## discard transcripts with average normalised counts lower than this
   top_n_variable = 500,     ## For PCA
   showCategory   = 25,      ## For enrichment analyses
   seed           = 1,       ## random seed gets set at start of script, just in case.
   filterFun      = IHW::ihw ## NULL for standard DESeq2 results, otherwise  functions,
   gene_clust     = bluster::HclustParam(cut.params = list(k=12))    )
)
```


### Cascading properties of a spec file {#cascade}

To save repeating oneself, there are various aspects of a spec file
that will propagate down the `specificiation > sample_sets > models >
comparisons` hierarchy.  We list the expected behaviour of those:

 - `models` can be set at the top `specification` level, and will get pre-pended to any sample-set specific models.
 - `subset` and `influential_samples`  can be set at the top `specification` level, and will get boolean-`AND`ed with sample-set specific subsets.
 - `transform` can be set at the top `specification` level, and will only be used when sample-set doesn't specify its own transform
 - `drop_unsupported_combinations` can be set at the `specification` and/or `sample_set` level, and will only be used when a model doesn't specify this directly.
 - `drop_incomplete` can be set at the `specification` and/or `sample_set` level, and will only be used when a model doesn't specify this directly.
 - `profile_plots` can be set at the `specification` and/or `sample_set` level, and will be added to those of any model-specific plot profiles
 - `extra_assay`s can be set at the `specification` level, and will be added to any sample_set-specific ones.

These are intended to 'do what I mean' for the most typical situations.

### Less verbose specification

Sometime the specfile can seem somewhat long-winded:

```{r, eval = FALSE}
specification(
  sample_sets =list(
    S1=sample_set(
      models=list(
       M1=model (
         comparisons=list(
	   c1 = comparison(
	   ...
```



We can cut this down in three ways:

#### Shortcuts for parent-list > child

There are a set of functions `sample_sets`, `models`, `comparisons` which replace the whole `plurals=list(singular...`, so the above can become:

```{r, eval = FALSE}
specification(
  sample_sets(
    S1=sample_set(
      models(
       M1=model (
         comparisons(
	   c1 = comparison(
```

Using R's meta-programming capabilities, these expand to the same syntax as the original above, there's often no need to have the explicit `list`.

#### Automatic wrapping of objects

A further level of abbreviation allows us to say that any singular
object (model, sample_set, comparison) that's not in a corresponding
list (models, samples_sets, comparisons) - in either of the above
syntaxes - should get wrapped in the corresponding 'plural list'
wrapper. For example, here we have a singular `model` that isn't
explicitly container in a `models` list, but the pipeline will now
automically adjust for that, so the following is valid:

```{r, eval = FALSE}
specification(
  sampleset(subset=TRUE,
    model(design = ~1,
      comparison(name="coef"
```

We can have multiple singletons, and give it/them short IDs to help identification:

```{r, eval = FALSE}
specification(
  sampleset(subset=TRUE,
    const=model(design = ~1,
      comparison(name="coef"...
    treat=model(design = ~treatment,
      comparison(name="coef"...
```



#### Automatic nesting

Sometime's we'll want, say, the same model fitted to each subset.  If
a model is defined at the top of the specification (ie alongside,
rather than within, a sample_set), then it will get added to all the
sample_sets it sits alongside.


```{r, eval = FALSE}
specification(
  sample_set(subset=TRUE),
  sample_set(subset=!ID in c("outlier1","outlier2")),
  model(design ~ treatment + batch)
)
  
```

will result in two sample_sets, each fitted with the same design.

Similarly a comparison that is not within a model will be applied
across the elements it is defined alongside:

```{r, eval = FALSE}
specification(
  sample_set(subset=TRUE),
  sample_set(subset=!ID in c("outlier1","outlier2")),
  model(design ~ treatment + batch),
  model(design ~ treatment)
  comparison(mult_comp(revpairwise ~ treatment)
)
```

will result in two sample_sets, each fitted with two models, and the
comparison between treatments being applied to these four settings
separetly. This is the same as the [cascading idea](#cascade)
mentioned previously, but works particularly well in combination with the two other tricks.


#### Availability of these helper functions

The `plural(` shortcut notation for `plural=list(` is available for the
following objects:

 - `sample_sets`
 - `models`
 - `comparisons`
 - `missingness`
 - `feature_filters`
 - `settings`
 - `strata`
 - `profile_plots`
 - `extra_assays`

basically, anywhere one used to type `plural=list(`!

The automatic wrapping of singleton objects in their corresponding
plurals is more limited to the primary specfile objects

 - `sample_set`
 - `model`
 - `comparison`
 - `extra_assay`
 - `profile_plot`
