---
title: Differential analysis
description: Visualisation and differential analysis
categories: [Differential]
author: 
  - name: Gavin Kelly
    affiliations: BABS - Francis Crick Institute
format:
  html:
    theme: cerulean
    fontsize: 1.1em
    linestretch: 1.7
code-fold: true
execute:
  output: asis
  echo: false
code-summary: "Show the code"
params:

---


# Preface

We load in all the necessary additional R packages, and set up some initial parameters.


```{r}
#| label: load-01
#| output: false
#| warning: false
#| error: false

library(tidyverse)       # Language
library(devtools)
library(openxlsx)        # IO
library(ggrepel)         # General Vis.
library(RColorBrewer)
library(ComplexHeatmap)
library(gt)
library(DESeq2)          # Assay-specific
library(PoiClaClu)
library(rtracklayer)
library(tximport)
library(clusterProfiler)
library(ReactomePA)
library(GO.db)
library(IHW)
library(equatiomatic)
library(DESdemonA)

fig_caption <- captioner()
set.seed(1)


knitr::opts_chunk$set(
  warning=FALSE, error=TRUE, message=FALSE,
  dev=c("ragg_png","pdf"), out.width="100%",
  fig.width=14, fig.height=10,
  results='asis', fig.cap=expression(fig_caption())
)
knitr::knit_hooks$set(
  error = function(x, options) {
    ERROR <<- x
    knitr::knit_exit()
  },
  chunk = function(x, options){
    if(exists('ERROR')) paste0(x,'```\n',ERROR,'```\n\n**Script stopped with error**')
    else x
  })


link2path <- function(lname) {  file.path(knitr::opts_knit$get("output.dir"),lname) }

```



```{r}
#| label: tbl-load2
#| tbl-cap: Context
if (!isTRUE(getOption('knitr.in.progress'))) {
  my_params_yml <- "resources/01_analyse.yml"
  if (file.exists(my_params_yml)) {
    params <- yaml::read_yaml(my_params_yml)
    if (is.null(names(params))) {
      params <- params[[1]]$params
    }
  } else {
    params <- list()
  }
} else {
  data.frame(
  Option=names(params),
  Value=unlist(params)
) %>%
  gt() %>%
  as_raw_html()
}



```

```{r}
#| label: read
if (is.character(params$alignment)) {
  data(list=paste0("counts_", params$alignment))
} else {
  dds <- params$alignment
}

if (!is.null(metadata(dds)$organism$org)) {
  library(metadata(dds)$organism$org, character.only=TRUE)
}
metadata(dds)$template_git <- packageDescription("DESdemonA")$git_last_commit

specs   <- DESdemonA::load_specs(file=paste0(params$spec, ".spec"), context=dds)

param <- DESdemonA::ParamList$new(defaults=specs$settings)
# Calling the setter without a value will pick up the default (from the spec), _and_ alert it in the markdown, and return the default
set.seed(param$set("seed"))
param$set("title", "{{{project}}}")
param$set("script", file.path(getwd(),"01_analyse.r"))
param$set("spec",  params$spec, "Using analysis plan '{}'.")
if (is.character(params$alignment)) {
  param$set("alignment", params$alignment, "Using alignment settings '{}'")
} else {
  param$set("alignment", params$param_call$alignment, "Using counts from '{}'")
}  
  
ddsList <- DESdemonA::build_dds_list(dds, specs)


#Which features are zero across all samples in all datasets
all_zero <- Reduce(f=`&`,
                  x=map_des(ddsList, function(x) apply(counts(x)==0,1, all)),
                  init=TRUE)
ddsList <- map_des(ddsList, function(dds) dds[!all_zero,])


## Calculate size factors, separately for each sample-set
ddsList <- mapply(
  function(samp, dds) {
    if ("subsetGenes" %in% names(samp)) {
      dds <- dds[eval(samp$subsetGenes, transform(rowData(dds), ID=row.names(dds))),]
    }
    if ("controlGenes" %in% names(samp)) {
      controlGenes <- eval(samp$controlGenes, transform(rowData(dds), ID=row.names(dds)))
      estimateSizeFactors(dds, controlGenes=controlGenes)
    } else {
      estimateSizeFactors(dds)
    }
  },
  specs$sample_sets,
  ddsList,
  SIMPLIFY=FALSE
)


param$set("baseMeanMin")
if (param$get("baseMeanMin")>0) {
  ddsList <- map_des(ddsList,
                   function(x) x[rowMeans(counts(x, normalized=TRUE)) >= param$get("baseMeanMin"),]
                   )
}
```

# Input Summary

The sample annotations are as follows:


```{r}
#| label: tbl-inputs
#| tbl-cap: Sample Annotation
table1 <- function(dds, ddsList) {
  samp <- as.data.frame(colData(dds))
  # Extra columns introduced by datasets
  extra_meta <- lapply(
    names(ddsList),
    function(x) merge(
      setNames(data.frame(ifelse(colnames(dds) %in% colnames(ddsList[[x]]), "✓", ""), row.names=row.names(colData(dds))), x),
      as.data.frame(colData(ddsList[[x]])[, setdiff(names(colData(ddsList[[x]])), names(colData(dds)))]),
      by=0, all.x=TRUE
    )[,-1,drop=FALSE]
  )
  
  extra_meta <- extra_meta[sapply(extra_meta, function(x) ncol(x)!=0 & nrow(x)!=0)]
  # Generate tab_spanner parameters for each datasets extra columns
  ts_extra <- lapply(names(extra_meta),
                    function(x) list(name=x,
                              cols=1:ncol(extra_meta[[x]]))
                    )
  # Cummulatively offset the columns to be spanned
  ts_extra <- Reduce(function(ts, x) {
    x$cols <- x$cols+max(ts$cols)
    x},
    ts_extra,
    accumulate=TRUE)
  
  df <- cbind(samp, do.call(cbind, unname(extra_meta)))
  clabel <- setNames(as.list(names(df)), 1:ncol(df))
  names(df) <- names(clabel)
  gt(data=df) %>%
    cols_label(.list=clabel) %>%
    tab_spanner(label="Metadata",
                columns=seq_along(samp)) %>%
    Reduce(f=function(gti, x) tab_spanner(gti, label=x$name,columns=x$cols + ncol(samp)), x=ts_extra, init=.)
}

table1(dds, ddsList) %>%
  as_raw_html()

```

We may examine the samples in different combinations, and leave out
certain samples.  In the above table, the columns under the 'In
Subset' group indicate these combinations are listed, and the
samples' inclusions are indicated.  Each of those combinations may be analysed in
potentially several ways, as formulated here:


```{r}
#| label: formulae
for (dataset in names(specs$sample_sets)) {
  mdls <- lapply(specs$sample_sets[[dataset]]$models, function(x) x$design)
  mdls <- mdls[sapply(specs$sample_sets[[dataset]]$model, function(x) "comparisons" %in% names(x))]
  for (mdl in names(mdls)) {
    cat("\n\n####", dataset, mdl, "{-}\n\n")
    print(extract_eq(lm(
      update(mdls[[mdl]], Expression ~ .),
      data=cbind(data.frame(colData(ddsList[[dataset]]), check.names=FALSE), Expression=1:ncol(ddsList[[dataset]])),
      wrap=TRUE
    )
    ))
  }
}

```


We also have the following defaults set.  Whenever they are used
or changed in the, analysis that will be highlighted in the text
alongside a 'wrench' icon:


```{r}
#| label: tbl-settings
#| tbl-cap: Analysis Settings

data.frame(
  Option=names(specs$settings),
  Value=sapply(specs$settings, deparse)
) %>%
  gt() %>%
  as_raw_html()

```


# QC Visualisation {.tabset}

These visualisations are carried out blind to the experimental
design. For the heatmap @fig-qc-visualisation-1, we select a subset of
genes removing the (assumed uninformative) genes with flat expression
levels across the whole sample set. We'd expect samples from the same
experimental group to cluster together, in the sense that they are on
the same branch of the tree. But this is a transcriptome-wide picture,
and even if the clustering is not perfect, there will still be genes
that are consistent with the expected expression profiles, and they
will be revealed in the differential analysis.

The second heatmap @fig-qc-visualisation-2 focuses on the pairwise
similarity of samples, using a slightly different metric than the
first, to emphasise that there's not one way of calculating how
similar two samples are, transcriptome-wide.  The colours in the first
heatmap are a gene's expression in a sample, relative to its average
expression.  The second heatmap represents the similarity of each pair
of samples.

The come a series of (pairs of) scatter plots of the first two
Principal Components - a way of summarising all the genes down to just
two 'metagenes' that attempt to encapsulate the variability across the
samples in as faithful way as possible. The first of each pair
(e.g. @fig-qc-visualisation-3) takes the PCs from the unadjusted data,
and colours them by each of your covariates/treatments in turn. They
alternate with 'focussed' PC adjustments, such as
@fig-qc-visualisation-4 which regress out ('normalise away') the
effect of all experimental factors except the one being focussed
on. This 'realigns' the principal components, and may be useful if
there's one experimental factor (e.g. strong batch effect) that
dominates the transcriptional profile and masking a less dominant
factor: we then colour by the one experimental factor that hasn't been
regressed out, hopefully allowing its effect to emerge.

We then revert to the PCA picture that is unadjusted for any
covariates. In this plot the columns explore the lower-order principal
components (which although not providing a universally strong signal
transcriptome wide, might still detect biology that is affecting a
smaller subset of genes, or a reasonably large set of genes to a
lesser extent than happens in the higher-order PCS). The rows
represent the experimental factors we have recorded, and the depth of
colour of the cell records the degree of 'association' that factor has
with the give PC (and grey says there was no statistically significant
association.) This being an automated report, we've had to make
various compromises so please don't read too much into this chart: it
primarily guides which other plots we should generate to look at how
samples associate with each other.

We use that information to derive the remaining plots in this section,
where we these components in the context of the various experimental
factors, producing 'partial residual plots' of two selected PCs, being
selected by the criteria of: 'best' being the PC most strongly
associated with the experimental factor; 'first' being the
highest-order PC associated with the experimental factor; and then
resorting to the first and second PCs in the case where we haven't
already generated two plots. 'Partial residual plots' are effectively
plots of the PC against the experimental factor, where once again we
have 'regressed out' the effect of all _other_ experimental factors at
play.


```{r}
#| label: fig-qc-visualisation


ddsList <- map_des(ddsList, function(x) DESdemonA::add_dim_reduct(x))

for (assay in c("counts", "norm", "vst")) {
  DESdemonA::write_assay(ddsList, assay=assay, path=params$res_dir)
}
                       
param$set("top_n_variable")
param$set("clustering_distance_rows")
param$set("clustering_distance_columns")

map_des(ddsList, 
        ~DESdemonA::qc_heatmap(
          .x,
          title=dataset_name(.x),
          caption=fig_caption,
          param=param$publish()
        )
        )
```


# Find differential genes

 We use DESeq2 [@pkg_DESeq2] to find differential genes using the
negative binomial distribution to model counts, with IHW [@pkg_IHW]
for multiple-testing correction with greater power than
Benjamini-Hochberg, and ashr [@pkg_ashr] for effect-size shrinkage
to ensure reported fold-changes are more robust.


```{r}
#| label: differential

param$set("alpha")
param$set("lfcThreshold")
param$set("filterFun")

## For each dataset, fit all its models
dds_model_comp <- map_des(
  ddsList,
  function(x) DESdemonA::fit_models(x,
                             minReplicatesForReplace = Inf)
)


## Now put results in each 3rd level mcols(dds)$results
dds_model_comp <- map_des(
  dds_model_comp,
  DESdemonA::get_result,
  filterFun=param$get("filterFun"),
  alpha=param$get("alpha"),
  lfcThreshold=param$get("lfcThreshold")
)



dds_name <- paste0(basename(params$spec),"_x_", params$alignment)
save(dds_model_comp,
     file=file.path("data", paste0(dds_name, ".rda")),
     eval.promises=FALSE
     )
saveRDS(dds_model_comp, file=file.path("data", paste0(dds_name, ".rds")))


xl_files <- DESdemonA::write_results(dds_model_comp, param, dir=params$res_dir)
iwalk(xl_files,
     ~cat("\n\n<a class=\"download-excel btn btn-primary\" href=\"", sub(paste0("^", params$res_dir, "/?"), "", .x), "\"> Open Spreadsheet '", .y,"'</a>", sep="")
     )

#DESdemonA::write_all_results(dds_model_comp, dir=params$res_dir)
```


## Summary Tables {.tabset}

Here we summarise the results of the differential testing. Each
sample-set gets its own table (flick between tabs to choose which),
within which the different models, and their null hypotheses, are
enumerated.

In the 'Significant' column we tally the number of significant
genes (for pairwise comparisons, separated into up or down, where
A-B being labelled up means expression is higher in A; for omnibus
comparisons, a broad categorisation of the most extreme groups, as
described above.) We also tally the total number of genes
exhibiting that behaviour (ie not necessarily statistically
signficant) - these might not always add up to the same value, as
there is an independent filter of low-signal genes whose effect
varies from comparison to comparison.

Our naming convention for a comparison often looks like `A-B|M` -
again, because this is an semi-automated report, this mightn't be
optimal, but the rationale is `A-B` refers to the two experimental
conditions being compared (a 'upward' fold change will be reported
if the expression is higher in `A` than it is in `B`). The `|M` part
refers to the fact that we may have looked at that `A-B` comparison
conditional upon some other variable (which in this example takes the
value `M`):  anything before the `|` symbol identifies the
experimental conditions that are being contrasted, and the part after
the `|` identifies any stratification/conditioning of the data.

In @fig-summary we present some diagnostic plots of the observed
p-values. Theoretically we want the pink part of the plot to have a
solid peak on the left, and then make a reasonable plateau across
higher p-values: gross deviations from this can indicate an absence of
differential behaviour (lack of left-peak) or a model
mis-specification that might need investigating. The blue tracks
indicate p-values corresponding to very lowly expressed genes, which
would typically struggle to provide evidence of differential
behaviour, and so show an example of what a 'bad' pink trace might
look like.

```{r}
#| label: tbl-summary
#| tbl-cap: Summaries

summaries <- dds_model_comp %>%
  map_des(DESdemonA::summarise_results) %>%
  map_des(function(x) {bind_rows(x, .id="Comparison")}, depth="model") %>%
  map_des(function(x) {bind_rows(x, .id="Model")}, depth="dataset") %>%
  map(function(x) {
    levels(x$Group) <- sub(".*<(.+<.+<0)$", "\\1", levels(x$Group))
    levels(x$Group) <- sub("^(0<.+?<.+?)<.*$", "\\1", levels(x$Group))
    levels(x$Group) <- sub("(.*?<)(.*0.*)(<.*)", "\\10\\3", levels(x$Group))
    droplevels(subset(x, Group!="Zero Count" & Group!="Low Count"))
  }
  )

pval_frame <- dds_model_comp %>%
  map_des(~as.data.frame(mcols(.x)$results)) %>%
  map_des(function(x) {bind_rows(x, .id="Comparison")}, depth="model")
                      
                      

## for (dataset in names(summaries)) {
##   #options(htmltools.preserve.raw=TRUE)
##   #cat("### ", dataset, " \n", sep="")
##   #caption=paste0("Size of gene-lists for ", dataset),
##   summaries[[dataset]] %>%
##     gt(groupname_col="Model") %>%
##     tab_header(title="Genelist summary",
##                subtitle=dataset) %>%
##     as_raw_html()
## }


for (dataset in names(summaries)) {
  res <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      paste0("#| label: tbl-summary-",dataset),
      paste0("#| tbl-cap: ",dataset),
      "summaries[[dataset]] %>%",
      "gt(groupname_col=\"Model\") %>%",
      "tab_header(title=\"Genelist summary\", subtitle=dataset)",
      "```",
      "",
      ""
    ), envir = environment(), quiet = TRUE)
  cat(res, sep = '\n')
}


```


```{r}
#| label: fig-summary
  
for (dataset in names(summaries)) {
  for (model in names(pval_frame[[dataset]])) {
    pl <- ggplot(pval_frame[[dataset]][[model]], aes(x=pvalue, fill=baseMean<1)) +
      geom_density(alpha=0.5) +
      facet_wrap(~Comparison) +
      theme_bw()
    print(pl)
    fig_caption(caption=paste0("p-Diagnostic for", dataset, model))
  }
}
```

# Scatterplot of differential genes {.tabset}

Here we plot the fold-change across all genes on the vertical axis,
against their overall expression on the horizontal. Statistically
significant genes are highlighted in colour, and we attempt to
auto-label as many gene symbols as might be legible.

The fold-changes are regularised ('shrunk') to reflect their
reliability: genes with large variability between replicates are
moved towards the horizontal axis, to de-emphasise them and to
provide a more reliable prediction of behaviour in future
replications.

Again, select the dataset you wish to examine via the tabs - there
are sub-tabs for different experimental designs, if any
(e.g. removing/ignoring batch effects).


```{r}
#| label: fig-differential-MA

for (dataset in names(dds_model_comp)) {
  cat("## ", dataset, " {.tabset} \n", sep="") 
  for (mdl in names(dds_model_comp[[dataset]])) {
    cat("### ", mdl, "\n", sep="") 
    differential_MA(dds_model_comp[[dataset]][[mdl]], caption=fig_caption)
  }
}
```


# Differential Heatmaps {.tabset}

Here we present heatmaps of the differential genelists.  Note
that these will, by definition, divide the samples along lines
of the experimental groups, so should be interpreted differently
from the QC heatmaps which were blind to the experimental design.

```{r}
#| label: fig-differential-heatmap
for (dataset in names(dds_model_comp)) {
  cat("## ", dataset, " {.tabset} \n", sep="") 
  for (mdl in names(dds_model_comp[[dataset]])) {
    cat("### ", mdl, "\n", sep="")
    DESdemonA::differential_heatmap(
      ddsList=dds_model_comp[[dataset]][[mdl]],
      param=param$publish(),
      caption=fig_caption
    )
  }
}
```

# Terms Of Use

The Crick has a [publication
policy](https://intranet.crick.ac.uk/our-crick/library-information-services/pages/guidelines-publication)
and we expect to be included on publications, regardless of funding
arrangements. Any use of these results in publication must be
discussed with BABS regarding authorship. If not authorship then
the BABS analyst must receive a named acknowledgement. Please also
cite the following sources which have enabled the analysis to be
carried out.

# Bibliography


