---
title: Exploratory analysis 
description: >-
  Exploratory analysis 
categories: [Exploratory]
author:
  - name: Gavin Kelly
    affiliations: BABS - Francis Crick Institute
format:
  html:
    theme: cerulean
    fontsize: 1.1em
    linestretch: 1.7
code-fold: true
execute:
  output: asis
  echo: false
code-summary: Show the code
params:
  section: Exploratory analysis
---


# Preface

We load in all the necessary additional R packages, and set up some initial parameters.


```{r}
#| label: load-exp1
#| output: false
#| warning: false
#| error: false

library(tidyverse)       # Language
library(devtools)
library(openxlsx)        # IO
library(ggrepel)         # General Vis.
library(RColorBrewer)
library(ComplexHeatmap)
library(gt)
library(DESeq2)          # Assay-specific
library(PoiClaClu)
library(rtracklayer)
library(tximport)
library(clusterProfiler)
library(ReactomePA)
library(GO.db)
library(IHW)

source("R/plot.r", local = knitr::knit_global())
source("R/markdown.r", local = knitr::knit_global())
source("R/rnaseq.r", local = knitr::knit_global())
source("R/partial.r", local = knitr::knit_global())
source("R/helpers.r", local = knitr::knit_global())
source("R/output.r", local = knitr::knit_global())

set.seed(1)


if (!isTRUE(getOption('knitr.in.progress'))) {
  params <- load_params(choose_staging("01_exploratory"))
}

do_plot <- plot_tracker(params)

knitr::opts_chunk$set(
  warning=FALSE, error=TRUE, message=FALSE,
  dev=c("ragg_png","pdf"), out.width="100%",
  dev.args=list(ragg_png=list(scaling=1.5)),
  fig.width=14, fig.height=10, fig.process=rename_with_tag(params),
  results='asis'
)

link2path <- function(lname) {  file.path(knitr::opts_knit$get("output.dir"),lname) }

```



```{r}
#| label: tbl-load-exp2
#| tbl-cap: Context

if (isTRUE(getOption('knitr.in.progress'))) {
  gt(data.frame(Setting=names(params),
                Value=unlist(params)))
}

```

```{r}
#| label: read-expl

data(list=paste0("counts_", params$alignment))

if (!is.null(metadata(dds)$organism$org)) {
  library(metadata(dds)$organism$org, character.only=TRUE)
}

specs   <- load_specs(file=paste0(params$spec, ".spec"), context=dds)
param <- ParamList$new(defaults=specs$settings)


```

We also have the following defaults set.  Whenever they are used
or changed in the analysis, that will be highlighted in the text
in a table identified with an info icon.


```{r}
#| label: tbl-settings
#| tbl-cap: Analysis Settings

data.frame(
  Option=names(specs$settings),
  Value=sapply(specs$settings, deparse)
) %>%
  gt() %>%
  as_raw_html()

```

```{r}
#| label: read-expl2

# Calling the setter without a value will pick up the default (from the spec), _and_ alert it in the markdown, and return the default
set.seed(param$set("seed", postpone=TRUE))
param$set("spec", params$spec, "Using analysis plan '{}'.", postpone=TRUE)
param$set("alignment", params$alignment, "Using alignment settings '{}'", postpone=TRUE)
param$set("baseline_heuristic", postpone=TRUE)
param$set("LRT_effect", postpone=TRUE)
param$set("baseMeanMin")

ddsList <- build_dds_list(dds, specs)
dtag <- dmc_factory(ddsList, type="dataset")
mtag <- dmc_factory(ddsList, type="model")


#Which features are zero across all samples in all datasets
all_zero <- Reduce(f=`&`,
                  x=lapply(ddsList, function(x) apply(counts(x)==0,1, all)),
                  init=TRUE)
ddsList <- lapply(ddsList, function(dds) dds[!all_zero,])


## Calculate size factors, separately for each sample-set
ddsList <- mapply(
  function(samp, dds) {
    if ("controlGenes" %in% names(samp)) {
      controlGenes <- eval(samp$controlGenes, transform(rowData(dds), ID=row.names(dds)))
      dds <- estimateSizeFactors(dds, controlGenes=controlGenes)
    } else {
      dds <- estimateSizeFactors(dds)
    }
    if ("subsetGenes" %in% names(samp)) {
      dds <- dds[eval(samp$subsetGenes, transform(rowData(dds), ID=row.names(dds))),]
    }
    return(dds)
  },
  specs$sample_sets,
  ddsList,
  SIMPLIFY=FALSE
)


if (param$get("baseMeanMin")>0) {
  ddsList <- lapply(ddsList,
                   function(x) x[rowMeans(counts(x, normalized=TRUE)) >= param$get("baseMeanMin"),]
                   )
}
ddsList <- lapply(ddsList, function(x) add_dim_reduct(x))


```

# Input Summary

The sample annotations are as follows:


```{r}
#| label: tbl-inputs
#| tbl-cap: Sample Annotation
#| column: page

table1 <- function(dds, ddsList) {
  samp <- as.data.frame(colData(dds))
  # Extra columns introduced by datasets
  extra_meta <- lapply(
    setNames(names(ddsList), names(ddsList)),
    function(x) {
      m <- merge(
        setNames(data.frame(ifelse(colnames(dds) %in% colnames(ddsList[[x]]), "✓", ""), row.names=row.names(colData(dds))), x),
        as.data.frame(colData(ddsList[[x]])[, setdiff(names(colData(ddsList[[x]])), c(".PCA", names(colData(dds))))]),
        by=0, all.x=TRUE
      )
      transform(m, row.names=Row.names, Row.names=NULL)[row.names(samp),,drop=FALSE]
    }
  )

  extra_meta <- extra_meta[sapply(extra_meta, function(x) ncol(x)!=0 & nrow(x)!=0)]
  # Generate tab_spanner parameters for each datasets extra columns
  ts_extra <- lapply(names(extra_meta),
                    function(x) list(name=x,
                              cols=1:ncol(extra_meta[[x]]))
                    )
  # Cummulatively offset the columns to be spanned
  ts_extra <- Reduce(function(ts, x) {
    x$cols <- x$cols+max(ts$cols)
    x},
    ts_extra,
    accumulate=TRUE)

  df <- cbind(samp, do.call(cbind, unname(extra_meta)))
  clabel <- setNames(as.list(names(df)), 1:ncol(df))
  names(df) <- names(clabel)
  gt(data=df) %>%
    cols_label(.list=clabel) %>%
    tab_spanner(label="Metadata",
                columns=seq_along(samp)) %>%
    Reduce(f=function(gti, x) tab_spanner(gti, label=x$name,columns=x$cols + ncol(samp)), x=ts_extra, init=.)
}

table1(dds, ddsList) %>%
  as_raw_html()


assay_files <- list()
for (assay in c("counts", "norm", "vst")) {
  fname <- write_assay(ddsList,
                       assay=assay,
                       p=params)
  assay_files[[assay]] <- attr(fname, "link")
}

save(ddsList,
     file=file.path("data",paste0("pca_", params$alignment, "_", params$spec, ".rda")))

```



# Datasets and Models

We may examine the samples in different combinations, and leave out
certain samples.  In the above table, the columns under the 'In
Subset' group indicate these combinations are listed, and the
samples' inclusions are indicated.

Each of those combinations may be analysed in
potentially several ways, as formulated here:

```{r}
#| label: formulae

for (dataset in names(ddsList)) {
  dmc_heading(ddsList, dataset=dataset, model=NULL, depth="##")
  met <- metadata(ddsList[[dataset]])
  mdls <- lapply(met$models, function(x) x$design)
  mdls <- mdls[sapply(met$models, function(x) "comparisons" %in% names(x))]
  for (mdl in names(mdls)) {
    dmc_heading(ddsList, dataset=dataset, model=mdl, depth="###", numbered=FALSE)
    cat("`",
        Reduce(paste, sub("~", "Expression ~ ", deparse(mdls[[mdl]]))),
        "`\n\n"
        )
  }
}

```


# Description of Visualisations

## Overall Heatmaps

These visualisations are carried out blind to the experimental
design. For the heatmap @fig-Heatmap-1, we select a subset of
genes removing the (assumed uninformative) genes with flat expression
levels across the whole sample set. We'd expect samples from the same
experimental group to cluster together, in the sense that they are on
the same branch of the tree. But this is a transcriptome-wide picture,
and even if the clustering is not perfect, there will still be genes
that are consistent with the expected expression profiles, and they
will be revealed in the differential analysis.

In case that heatmap doesn't cluster exactly according to the
experimental design, in @fig-SuperV-Heatmap-1 we provide the
samples ordered according to one particular interpretation of the
experimental design.  In simple (one-way) designs where there is only
one covariate, this is a fairly straightforward way of gathering
samples in a set of columns, but in more complex designs it isn't
obvious whether to sort first by genotype and then treatment, say, or
perhaps the reverse, so for this 'supervised' clustering we chose the
(arbitrary) order the 'model' was specified, above.

The third heatmap @fig-sampDist-1 focuses on the pairwise
similarity of samples, using a slightly different metric than the
first, to emphasise that there's not one way of calculating how
similar two samples are, transcriptome-wide.  The colours in the
previous heatmaps are a gene's expression in a sample, relative to its
average expression. In this heatmap, colour represents the similarity
of each pair of samples.

### Focussed Heatmaps

Heatmap @fig-Heatmap-correct-1 is the first in a series of
'focussed' plots. By focussed, we mean that in turn we remove the main
effect of some of the covariates, and then also optionally force the
tree-structure to respect one of the remaining covariate (samples
within the given enforced covariate group are allowed to cluster
empirically, and the covariate groups themselves are also clustered at
the aggregate level.)  Sometimes this allows structures that are
masked by one 'dominant' effect to emerge, or to remove annoyance
effects such as that due to batch which might be preventing the real
structure from being apparent. The interpretation of these is very
specific to the experimental design. We can alter which covariates
split the heatmap, and also which covariates are regressed out before
the heatmap is calculated, so please consult with the analyst if you
want to change this: the default is that we remove the effect of
_everything_ other than the factor being used to colour a specific
plot, but this can be changed (e.g. if you just want to see what the
result of removing a batch effect looks like, but want to preserve a
genotype x treatment composite)

## Principal Component Scatterplots

The come a series of scatter plots of the first two
Principal Components - a way of summarising all the genes down to just
two 'metagenes' that attempt to encapsulate the variability across the
samples in as faithful way as possible. Each plot
(e.g. @fig-color-1) takes the PCs from the unadjusted data,
and colours them by each of your covariates/treatments in turn. 

## Focussed Principal Component Scatterplots

Principal components, just like clusters in the heatmaps, can be
driven by a dominant effect that might mask other important
effects. So again we have the flexibility to remove systematic effects
before recalculating the PCs such as in @fig-focus-1 . which
regress out ('normalise away') the effect of a set of covariates. This
'realigns' the principal components. We then colour by one of the
experimental factors that hasn't been regressed out, hopefully
allowing its effect to emerge, or to diagnose which is causing the
clustering. Again, the default is to regress out everything other than
the effect being used to colour the plot, but that can (or maybe has
been - read the legends to double check!) be customised.


## PC-Covariate assocation

We then revert to the PCA picture that is unadjusted for any
covariates. In @fig-covar-1 the columns explore the lower-order principal
components (which although not providing a universally strong signal
transcriptome wide, might still detect biology that is affecting a
smaller subset of genes, or a reasonably large set of genes to a
lesser extent than happens in the higher-order PCS). The rows
represent the experimental factors we have recorded, and the depth of
colour of the cell records the degree of 'association' that factor has
with the give PC (and grey says there was no statistically significant
association.) This being an automated report, we've had to make
various compromises so please don't read too much into this chart: it
primarily guides which other plots we should generate to look at how
samples associate with each other.

## Partial Residuals of PCAs

We use that information to derive the remaining plots such as @fig-covariate-1,
where we these components in the context of the various experimental
factors, producing 'partial residual plots' of two selected PCs, being
selected by the criteria of: 'best' being the PC most strongly
associated with the experimental factor; 'first' being the
highest-order PC associated with the experimental factor; and then
resorting to the first and second PCs in the case where we haven't
already generated two plots. 'Partial residual plots' are effectively
plots of the PC against the experimental factor, where once again we
have 'regressed out' the effect of all _other_ experimental factors at
play.


```{r}
#| label: fig-qc-visualisation


param$set("top_n_variable", postpone=TRUE)
param$set("clustering_distance_rows", postpone=TRUE)
param$set("clustering_distance_columns", postpone=TRUE)
param$set("gene_clust", postpone=TRUE)
param$set("pc_x", 1, postpone=TRUE)
param$set("pc_y", 2)
first <- TRUE
all_clusters <- list()
cluster_ind <- counter(post_fn=to_letter)

for (dataset_id in names(ddsList)) {
  this_dds <- ddsList[[dataset_id]]
  dataset_name <- metadata(this_dds)$dmc$dataset_name %||% dataset_id
  # Only produce visualisations for models with a qc formula
  # (if none supplied in the spec file, the first model will get one corresponding to its design formula)
  dmc_heading(ddsList, dataset=dataset_id, model=NULL, depth="#")
  mdls <- sapply(metadata(this_dds)$models, function(mdl) "profile_plots" %in% names(mdl) )
  for (model_id in names(mdls[mdls])) {
    plot_formulae <- metadata(this_dds)$models[[model_id]]$profile_plots
    design <- metadata(this_dds)$models[[model_id]]$design
    model_vars <- list(
      design=all.vars(design),
      plot=unique(unlist(lapply(plot_formulae, all.vars))),
      all=unique(c(all.vars(design),
                   unique(unlist(lapply(plot_formulae, all.vars)))
                   ))
    )
    do_focus <- length(plot_formulae)>0

    if (sum(mdls)>1) {
      dmc_heading(ddsList, dataset=dataset_id, model=model_id, depth="##")
    }
    var_stab <- assay(this_dds, "vst")
    top <- order(apply(var_stab, 1, sd), decreasing = TRUE)[1:min(nrow(var_stab),param$get("top_n_variable"))]
    nice_legends <- separate_legend(ddsList[[dataset_id]], model_vars$all)
    pl <- do.call(gridExtra::arrangeGrob, c(
      nice_legends,
      ncol=ceiling(sqrt(length(nice_legends)))))
    do_plot(pl,
            label= paste0("legend-", dataset_id),
            caption=paste("Metadata colour-codings for", dtag(dataset_id))
            )
    cat(var_heading("Overall Heatmap of Variable Genes in ", dtag(dataset_id),  depth=2 + (sum(mdls)>1)))
    plotDat <- var_stab[top, ]
    plotDat <- plotDat - rowMeans(plotDat)
    colDat <- as.data.frame(colData(this_dds))
    colnames(plotDat) <- rownames(colDat)
    gene_clust <- bluster::clusterRows(plotDat, param$get("gene_clust"), full=TRUE)
    this_clust <- cluster_ind()
    levels(gene_clust$clusters) <- paste0(this_clust, levels(gene_clust$clusters))
    pl <- ComplexHeatmap::Heatmap(
      plotDat, name = "Mean Centred", 
      column_title = "Samples", row_title = "Genes", col = sym_colour(plotDat), 
      clustering_distance_columns = param$get("clustering_distance_columns"), 
      cluster_rows = gene_clust$objects$hclust,
      heatmap_legend_param = list(direction = "horizontal"), 
      top_annotation = ComplexHeatmap::HeatmapAnnotation(
        df = colDat[model_vars$all], 
        col = metadata(colData(this_dds))$palette$Heatmap[model_vars$all],
        show_legend=FALSE), 
      show_row_names = FALSE, show_column_names = TRUE)
    do_plot(pl,
            label= "Heatmap",
            caption=paste("Heatmap of variable genes in", dtag(dataset_id)),
            preview=first)

    first <- FALSE
    cat("\n\n---\n\n")


    pl <- ComplexHeatmap::Heatmap(
      plotDat, name = "Mean Centred",
      column_title = "Samples", row_title = "Genes", col = sym_colour(plotDat), 
      cluster_columns = FALSE,
      column_order=do.call(order, colDat[model_vars$all]),
      cluster_rows = gene_clust$objects$hclust,
      heatmap_legend_param = list(direction = "horizontal"),
      top_annotation = ComplexHeatmap::HeatmapAnnotation(
        df = colDat[,model_vars$all],
        col = metadata(colData(this_dds))$palette$Heatmap[model_vars$all],
        show_legend=FALSE),
      show_row_names = FALSE, show_column_names = TRUE)
    do_plot(pl,
            label= "SuperV-Heatmap",
            caption=paste("Supervised Heatmap of variable genes in", dtag(dataset_id)))
    cat("\n\n---\n\n")

    poisd <- PoiClaClu::PoissonDistance(t(counts(this_dds, normalized = TRUE)))
    samplePoisDistMatrix <- as.matrix(poisd$dd)
    if (length(model_vars$all) > 1) {
      rownames(samplePoisDistMatrix) <- Reduce(
        function(...) paste(..., sep = "_"), colDat[model_vars$all])
    } else {
      rownames(samplePoisDistMatrix) <- colDat[[model_vars$all]]
    }
      colnames(samplePoisDistMatrix) <- row.names(colDat)
    pl <- ComplexHeatmap::Heatmap(
      samplePoisDistMatrix,
      col = circlize::colorRamp2(
        c(0,  max(samplePoisDistMatrix)),
        colors = c("white", "red")), 
      name = "Poisson Distance", clustering_distance_rows = function(x) {poisd$dd},
      clustering_distance_columns = function(x) { poisd$dd},
      heatmap_legend_param = list(direction = "horizontal"), 
      top_annotation = ComplexHeatmap::HeatmapAnnotation(
        df = colDat[model_vars$all], 
        col = metadata(colData(this_dds))$palette$Heatmap[model_vars$all],
        show_legend=FALSE))
    do_plot(pl,
            label= "sampDist",
            caption=paste("Heatmap of sample distances in", dtag(dataset_id)))
    cat("\n\n---\n\n")

    if (do_focus) {
      cat(var_heading("Raw Cluster Profiles",   depth=2 + (sum(mdls)>1)))
      for (profile in plot_formulae[1]) {
        cc <- cluster_calc(plotDat, clusterID=gene_clust$clusters)
        centroid_df <- cbind(cc$centroid, colDat[model_vars$all])  %>%
          pivot_longer(cols=any_of(names(cc$centroid)), names_to="babs.name", values_to="babs.value") %>%
          mutate(babs.name=factor(babs.name, levels=names(cc$centroid)))
        pl <- ggplot(centroid_df,
                    eval(profile[[2]])) +
          facet_wrap(~ babs.name) +
          theme_bw() +
          geom_point(aes(y= babs.value)) +
          labs(y="Expression")
        if ("group" %in% names(profile[[2]])) {
          pl <- pl+geom_line(aes(y=babs.value))
        }
        if ("ScaleDiscrete" %in% class(layer_scales(pl)$x)) {
          pl <- pl + scale_x_discrete(guide=guide_axis(n.dodge=2))
        }
        capt <- paste("Cluster profile in", dtag(dataset_id))
        do_plot(pl,
                label= "clustProfile",
                caption=capt)
        all_clusters[[this_clust]] <- list(
          id=gene_clust$clusters,
          description=capt
        )

      }}
    
    is_vary <- sapply(colDat[model_vars$all], function(v) length(unique(v)) != 1)
    
    if (do_focus) {
      design(this_dds) <- design
      partials <- partialise(this_dds)
      cat(var_heading("Focussed Heatmaps",   depth=2 + (sum(mdls)>1)))
      for (fml in plot_formulae) {
        splitters <- c(NA, fml[[2]]$column_split)#NA denotes 'no split'.
        for (split in splitters) {
          removed <- paste(
            translate_terms(dropped_terms(partials, update(fml, NULL ~ .)), this_dds),
            sep=", ", collapse=", ")
          if (length(removed)==0 || removed=="") removed <- "nothing"
          part_res <- assemble_partialised(partials, update(fml, NULL ~ .))
          group_mat <- assemble_partialised(partials, update(fml, NULL ~ .), resids=FALSE)
          if (is.na(split)) {
            my_caption <- paste("In",  dtag(dataset_id), "having removed effect of", removed)
          } else {
            my_caption <- paste("Split on", paste(all.vars(split), collapse=":"), "in",  dtag(dataset_id), "having removed effect of", removed) # TODO
          }
          top <- order(apply(part_res, 1, sd), decreasing = TRUE)[1:min(nrow(part_res),param$get("top_n_variable"))]
          plotDat <- part_res[top, ]
          group_mat <- group_mat[top,]
          gene_clust <- bluster::clusterRows(plotDat, param$get("gene_clust"), full=TRUE)
          this_clust <- cluster_ind()
          levels(gene_clust$clusters) <- paste0(this_clust, levels(gene_clust$clusters))
          colnames(plotDat) <- rownames(colDat)
          pl <- ComplexHeatmap::Heatmap(
            plotDat, name = "Residual", 
            column_title = "Samples", row_title = "Genes", col = sym_colour(plotDat),
            column_split = if (is.na(split)) NULL else colDat[all.vars(split)],
            clustering_distance_columns = param$get("clustering_distance_columns"), 
            cluster_rows = gene_clust$objects$hclust,
            heatmap_legend_param = list(direction = "horizontal"), 
            top_annotation = ComplexHeatmap::HeatmapAnnotation(
              df = colDat[model_vars$all], 
              col = metadata(colData(this_dds))$palette$Heatmap[model_vars$all],
              show_legend=FALSE), 
            show_row_names = FALSE, show_column_names = TRUE)
          do_plot(pl,
                  label= "Heatmap-correct",
                  caption=my_caption
                  )
          cat("\n\n---\n\n")
          cc <- cluster_calc(plotDat, clusterID=gene_clust$clusters, group_mat=group_mat)
          centroid_df <- cbind(cc$centroid, colDat[model_vars$all]) %>%
            pivot_longer(cols=any_of(names(cc$centroid)), names_to="babs.name", values_to="babs.value") %>%
            mutate(babs.name=factor(babs.name, levels=names(cc$centroid)))
          pl <- ggplot(centroid_df ,
                      eval(profile[[2]])) +
            facet_wrap(~ babs.name) +
            theme_bw() +
            geom_point(aes(y= babs.value)) +
            labs(y="Expression")
          if ("group" %in% names(profile[[2]])) {
            group_df <- cbind(cc$group_centroid, colDat[model_vars$all])  %>%
              pivot_longer(cols=any_of(names(cc$centroid)), names_to="babs.name", values_to="babs.value") %>%
              mutate(babs.name=factor(babs.name, levels=names(cc$centroid)))
            pl <- pl+geom_line(data=group_df, aes(y=babs.value))
          }
          if ("ScaleDiscrete" %in% class(layer_scales(pl)$x)) {
            pl <- pl + scale_x_discrete(guide=guide_axis(n.dodge=2))
          }
          capt <- paste("Cluster profile in", dtag(dataset_id), "having removed", removed)
          do_plot(pl,
                  label= "clustProfile",
                  caption=capt)
          all_clusters[[this_clust]] <- list(
            id=gene_clust$clusters,
            description=capt
          )
          
        } # next split
      }   # next formula
    }

    ################################################################
    
    cat(var_heading("PC-Covariate Associations", depth=2 + (sum(mdls)>1)))
    pc <- as.matrix(colData(this_dds)$.PCA)
    pc_frame <- expand.grid(sample = rownames(colDat), PC = 1:ncol(pc))
    pc_frame$coord <- as.vector(pc)
    pc_frame <- cbind(pc_frame, colDat)
    fitFrame <- colDat
    yvar <- make.unique(c(colnames(fitFrame), "y", sep = ""))[ncol(fitFrame) + 1]
    fml <- update(design, 
                 paste(yvar, "~ ."))
    covar_x_pc <- expand.grid(
      Covariate = attr(terms(fml), "term.labels"),
      PC = 1:ncol(pc))
    covar_x_pc$Assoc <- NA
    npc <- ncol(pc)
    fit_selected <- list()
    for (ipc in 1:npc) {
      fitFrame[[yvar]] <- pc[, ipc]
      fit1 <- lm(fml, data = fitFrame)
      ind_complete <- intersect(row.names(fitFrame), names(residuals(fit1)))
      if (length(ind_complete) < nrow(fitFrame)) {
        fit1 <- lm(fml, data = fitFrame[ind_complete,])
      }
      fit_selected[[ipc]] <- MASS::stepAIC(fit1, trace = 0)
      ind <- attr(terms(fit_selected[[ipc]]), "term.labels")
      if (length(ind)) {
        fit0 <- as.data.frame(anova(fit_selected[[ipc]]))
        rss <- fit0[ind, "Sum Sq"]/sum(fit0[, "Sum Sq"])
        covar_x_pc$Assoc[covar_x_pc$PC == ipc][match(ind,  covar_x_pc$Covariate[covar_x_pc$PC == ipc])] <- rss
      }
    }
    covar_x_pc$wrap <- (covar_x_pc$PC - 1)%/%20
    covar_x_pc$wrap <- paste0("PCs ", covar_x_pc$wrap * 20 + 
                                        1, "-", min((covar_x_pc$wrap + 1) * 20, npc))
    covar_x_pc$PC <- sprintf("%02d", covar_x_pc$PC)
    pl <- ggplot(covar_x_pc, aes(x = PC, y = Covariate, fill = Assoc)) + 
      geom_raster() +
      facet_wrap(~wrap, scales = "free_x",ncol = 1) +
      scale_fill_gradient2(low = "#4575b4",mid = "grey90", high = "#d73027") +
      theme_bw() + 
      theme(aspect.ratio = length(unique(covar_x_pc$Covariate))/min(20, npc))
    do_plot(pl,
            label="covar",
            caption=paste("Covariate association in", dtag(dataset_id))
            )

    ################################################################
    
    cat(var_heading("Principal Component Scatterplots of PCs ", param$get("pc_x"), " and ", param$get("pc_y"), 
                    depth=2 + (sum(mdls)>1)))
    do_labels <- nrow(colDat) < 10

    percentVar <- metadata(colData(this_dds)$.PCA)$percentVar
    done_aes <- list()
    for (fml in plot_formulae) {
      if (as.character(fml)[2] %in% done_aes) {
        next
      } else {
        done_aes <- c(done_aes, as.character(fml)[2])
      }
      aes_list <- substitute_x_aes(fml[[2]])
      for (this_aes in aes_list) {
        ## For PCA plots, we do an unregressed PCA, but using the aes from the plot_formulae
        pc.df <- cbind(colDat,
                      data.frame(
                        PC1 = pc[, param$get("pc_x")],
                        PC2 = pc[, param$get("pc_y")], 
                        sample = rownames(colDat)))
        pl <- ggplot(
          pc.df,
          modifyList(eval(this_aes), aes(x=PC1, y=PC2))) +
          geom_point(size = 5) +
          xlab(paste0("PC ", param$get("pc_x"), ": ",  percentVar[param$get("pc_x")], "% variance")) +
          ylab(paste0("PC ",param$get("pc_y"), ": ", percentVar[param$get("pc_y")], "% variance")) +
          theme_bw()
        colour_by <- all.vars(eval(this_aes)$colour)
        if (length(colour_by)>1) colour_by <- NA
        if (is.numeric(colDat[[colour_by]])) {
          pal_col <- metadata(colData(this_dds))$palette$ggplot[[colour_by]]
          pl <- pl + scale_colour_gradient(low = pal_col[1], high = pal_col[2])
        }  else {
          pl <- pl + scale_colour_manual(values = metadata(colData(this_dds))$palette$ggplot[[colour_by]])
        }
        if (do_labels) {
          pl <- pl + geom_text_repel(aes(label = sample))
        }
        do_plot(pl,
                label="color",
                caption=paste("Annotated by", aes_caption(eval(this_aes)), "in", dtag(dataset_id))
                )
        cat("\n\n---\n\n")
      } # Repeat if need to annotate by an overwritten 'x'
    } # next formula

    ################################################################
    
    if (do_focus) {
      cat(var_heading("Focussed Principal Component Scatterplots of PCs ", param$get("pc_x"), " and ", param$get("pc_y"),
                      depth=2 + (sum(mdls)>1)))
      for (fml in plot_formulae) {
        aes_list <- substitute_x_aes(fml[[2]])
        my_pc <- prcomp(
          t(assemble_partialised(partials, update(fml, NULL ~ .))),
          scale = FALSE)
        my_percent <- round(100 * my_pc$sdev^2/sum(my_pc$sdev^2))
        pc.df <- cbind(colDat,
                      data.frame(
                        PC1 = my_pc$x[, param$get("pc_x")],
                        PC2 = my_pc$x[, param$get("pc_y")], 
                        sample = rownames(colDat)))
        for (this_aes in aes_list) {
          pl <- ggplot(pc.df,
                      modifyList(eval(this_aes), aes(x=PC1, y=PC2))) + 
            geom_point(size = 5) + 
            xlab(paste0("PC ", param$get("pc_x"), ": ", my_percent[param$get("pc_x")], "% variance")) +
            ylab(paste0("PC ", param$get("pc_y"), ": ", my_percent[param$get("pc_y")], "% variance")) + 
            theme_bw()
          colour_by <- all.vars(eval(this_aes)$colour)
          if (length(colour_by)>1) colour_by <- NA
          if (is.numeric(colDat[[colour_by]])) {
            pal_col <- metadata(colData(this_dds))$palette$ggplot[[colour_by]]
            pl <- pl + scale_colour_gradient(low = pal_col[1], high = pal_col[2])
          } else {
            pl <- pl + scale_colour_manual(values = metadata(colData(this_dds))$palette$ggplot[[colour_by]])
          }
          if (do_labels) {
            pl <- pl + geom_text_repel(aes(label = sample))
          }
          removed <- paste(
            translate_terms(dropped_terms(partials, update(fml, NULL ~ .)),this_dds),
            sep=", ", collapse=", ")          
          if (length(removed)==0 || removed=="") removed <- "nothing"
          do_plot(pl,
                  label="focus",
                  caption=paste("Annotated by", aes_caption(eval(this_aes)), "in", dtag(dataset_id), "having removed effect of", removed)
                  )
          cat("\n\n---\n\n")
        } # Repeat if need to annotate by an overwritten 'x'
      } # Next formula
    }
    
    ################################################################
    
    if (do_focus) {
      cat(var_heading("Partial Residuals of Associated PCs ", depth=2 + (sum(mdls)>1)))
      pc_partials <- partialise.matrix(t(pc), colDat, design)
      for (fml in plot_formulae) {
        if (length(all.vars(fml[[3]]))==0) next
        mat <- assemble_partialised(pc_partials, update(fml, NULL ~ .))
        pc_hits <- data.frame(covar=all.vars(fml[[3]]), strongest=NA_character_, first=NA_character_, row.names=all.vars(fml[[3]]))
        for (covar in all.vars(fml[[3]])) {
          this_covar <- subset(covar_x_pc,
                              Covariate == covar & 
                                !is.na(Assoc) & PC != PC[nrow(covar_x_pc)])
          if (nrow(this_covar) == 0) 
            next
          pc_hits[covar, "first"] <- as.character(this_covar$PC[1])
          this_covar <- this_covar[order(this_covar$Assoc, 
                                        decreasing = TRUE), ]
          pc_hits[covar, "strongest"] <- as.character(this_covar$PC[1])
        }
        hits <- paste0("PC", sort(unique(as.integer(c(pc_hits$strongest, pc_hits$first)))))
        wide_dat <- cbind(colDat[model_vars$all], t(mat)[,hits,drop=FALSE])
        long_dat <- pivot_longer(wide_dat, cols=all_of(hits), names_to="PC")
        pl <- ggplot(long_dat, modifyList(eval(fml[[2]]), aes(y=value))) + geom_point() + facet_wrap(~PC)
        if ("group" %in% names(fml[[2]])) pl <- pl+geom_line()
        if ("ScaleDiscrete" %in% class(layer_scales(pl)$x)) {
          pl <- pl + scale_x_discrete(guide=guide_axis(n.dodge=2))
        }
        removed <- paste(
          translate_terms(dropped_terms(partials, update(fml, NULL ~ .)),this_dds),
          sep=", ", collapse=", ")
        if (length(removed)==0 || removed=="") removed <- "nothing"
        do_plot(pl,
                label= "covariate",
                caption=paste("Annotated by",aes_caption(eval(fml[[2]])), "in", dtag(dataset_id), "having removed effect of", removed) 
                )
        cat("\n\n---\n\n")
      }
    }
    
    ## factors_1 <- attr(terms(fit1), "factors")
    ## many_levels <- sapply(colDat, function(x) length(unique(x)))
    ## for (covar in unique(covar_x_pc$Covariate)) {
    ##   this_covar <- subset(covar_x_pc,
    ##                       Covariate == covar & 
    ##                         !is.na(Assoc) & PC != PC[nrow(covar_x_pc)])
    ##   if (nrow(this_covar) == 0) 
    ##     next
    ##   pc_first <- as.integer(as.character(this_covar$PC[1]))
    ##   this_covar <- this_covar[order(this_covar$Assoc, 
    ##                                 decreasing = TRUE), ]
    ##     pc_strong <- as.integer(as.character(this_covar$PC[1]))
    ##     pc.df <- subset(pc_frame, PC %in% c(pc_first, pc_strong))
    ##     pc.df$label <- as.character(pc.df$PC)
    ##     pc.df$label[pc.df$PC == pc_first] <- paste0(pc.df$label[pc.df$PC == 
    ##                                                             pc_first], "+First")
    ##     pc.df$label[pc.df$PC == pc_strong] <- paste0(pc.df$label[pc.df$PC == 
    ##                                                              pc_strong], "+Best")
    ##     pc.df$label <- paste0(pc.df$label, " (", percentVar[pc.df$PC], 
    ##                          "%)")
    ##     inter_vars <- rownames(factors_1)[factors_1[, covar] !=  0]
    ##     inter_vars <- inter_vars[order(many_levels[inter_vars], decreasing = TRUE)]
    ##     pc.df$X <- colDat[, inter_vars[1]]
    ##     for (ipc in unique(c(pc_first, pc_strong))) {
    ##       pc.df$coord[pc.df$PC == ipc] <- part.resid(fit_selected[[ipc]])[, covar]
    ##     }
    ##     if (length(inter_vars) > 1) {
    ##       pc.df$col <- colDat[, inter_vars[2]]
    ##       if (length(inter_vars) == 2) {
    ##         pl <- ggplot(pc.df, aes(x = X, y = coord, colour = col, group = col)) +
    ##           stat_summary(fun = mean,geom = "line") +
    ##           labs(x = inter_vars[1], y = "Residual",
    ##                colour = paste(inter_vars[-1], collapse = "x")) + theme_bw()
    ##         if (is.numeric(pc.df$col)) {
    ##           pal_col <- metadata(colData(this_dds))$palette$ggplot[[inter_vars[2]]]
    ##           pl <- pl + scale_colour_gradient(low = pal_col[1], high = pal_col[2])
    ##         }
    ##         else {
    ##           pl <- pl + scale_colour_manual(values = metadata(colData(this_dds))$palette$ggplot[[inter_vars[2]]])
    ##         }
    ##       }
    ##       else {
    ##         pc.df$grp <- Reduce(interaction, colDat[, inter_vars[-1]])
    ##         pl <- ggplot(pc.df, aes(x = X, y = coord, colour = col,  group = grp)) +
    ##           stat_summary(fun = mean, geom = "line") +
    ##           labs(x = inter_vars[1], y="Residual",
    ##                colour = paste(inter_vars[-1], collapse = "x")) +
    ##           theme_bw()
    ##         if (is.numeric(pc.df$col)) {
    ##           pal_col <- metadata(colData(this_dds))$palette$ggplot[[inter_vars[2]]]
    ##           pl <- pl + scale_colour_gradient(low = pal_col[1], high = pal_col[2])
    ##         }
    ##         else {
    ##           pl <- pl + scale_colour_manual(values = metadata(colData(this_dds))$palette$ggplot[[inter_vars[2]]])
    ##         }
    ##       }
    ##     }
    ##     else {
    ##       pl <- ggplot(pc.df, aes(x = X, y = coord, colour = X, group = 1)) +
    ##         stat_summary(fun = mean, geom = "line", colour = "black") +
    ##         labs( x = inter_vars[1], y="Residual", colour = covar) +
    ##         theme_bw()
    ##       if (is.numeric(pc.df$X)) {
    ##         pal_col <- metadata(colData(this_dds))$palette$ggplot[[inter_vars[1]]]
    ##         pl <- pl + scale_colour_gradient(low = pal_col[1], high = pal_col[2])
    ##       }
    ##       else {
    ##         pl <- pl + scale_colour_manual(values = metadata(colData(this_dds))$palette$ggplot[[inter_vars[1]]])
    ##       }
    ##     }
    ##     pl <- pl + geom_point(size = 2) +
    ##       facet_wrap(~label, scales = "free_y")
    ##     if (do_labels) {
    ##       pl <- pl + geom_text_repel(aes(label = sample))
    ##     }
    ##     do_plot(pl,
    ##             label= "covariate",
    ##             caption=paste("Against", covar, "in", dtag(dataset_id))
    ##             )
    ##     cat("\n\n---\n\n")
    ##   }
    ## #}
  }
}




```

# Downloads


```{r}
#| label: downloads

iwalk(assay_files,
     ~cat("\n\n<a class=\"download-excel btn btn-primary\" href=\"", attr(.x, "link"), "\">", ifelse(.y=="counts", "raw", .y) ," count matrix</a>", sep="")
     )

fname <- write_clusters(all_clusters,
                       mcols(ddsList[[1]]),
                       p=params,
                       prefix="exploratory-clusters")


cat('\n\n<a class="download-excel btn btn-primary" href="',
    attr(fname,"link"),
    '">Gene-Cluster mappings</a>', sep="")
     

```


# Terms Of Use

The Crick has a [publication
policy](https://intranet.crick.ac.uk/our-crick/library-information-services/pages/guidelines-publication)
and we expect to be included on publications, regardless of funding
arrangements. Any use of these results in publication must be
discussed with BABS regarding authorship. If not authorship then
the BABS analyst must receive a named acknowledgement. Please also
cite the following sources which have enabled the analysis to be
carried out.



