---
title: Exploratory analysis 
description: >-
  Exploratory analysis 
categories: [Exploratory]
author:
  - name: Gavin Kelly
    affiliations: BABS - Francis Crick Institute
format:
  html:
    theme: cerulean
    fontsize: 1.1em
    linestretch: 1.7
code-fold: true
execute:
  output: asis
  echo: false
code-summary: Show the code
params:
---


# Preface

We load in all the necessary additional R packages, and set up some initial parameters.


```{r}
#| label: load-01
#| output: false
#| warning: false
#| error: false

library(tidyverse)       # Language
library(devtools)
library(openxlsx)        # IO
library(ggrepel)         # General Vis.
library(RColorBrewer)
library(ComplexHeatmap)
library(gt)
library(DESeq2)          # Assay-specific
library(PoiClaClu)
library(rtracklayer)
library(tximport)
library(clusterProfiler)
library(ReactomePA)
library(GO.db)
library(IHW)
library(equatiomatic)
library(DESdemonA)

source("R/plot.r")
fig_caption <- captioner()
set.seed(1)


knitr::opts_chunk$set(
  warning=FALSE, error=TRUE, message=FALSE,
  dev=c("ragg_png","pdf"), out.width="100%",
  fig.width=14, fig.height=10,
  results='asis', fig.cap=expression(fig_caption())
)
if (do_knit_error <- FALSE) {
knitr::knit_hooks$set(
  error = function(x, options) {
    ERROR <<- x
    knitr::knit_exit()
  },
  chunk = function(x, options){
    if(exists('ERROR')) paste0(x,'```\n',ERROR,'```\n\n**Script stopped with error**')
    else x
  })
}

link2path <- function(lname) {  file.path(knitr::opts_knit$get("output.dir"),lname) }

```



```{r}
#| label: tbl-load2
#| tbl-cap: Context

if (!isTRUE(getOption('knitr.in.progress'))) {
  my_params_yml <- "resource/01_exploratory.yml"
  if (file.exists(my_params_yml)) {
    params <- yaml::read_yaml(my_params_yml)
    if (is.null(names(params))) {
      params <- params[[1]]$params
    }
  } else {
    params <- list()
  }
} else {
  data.frame(
  Option=names(params),
  Value=unlist(params)
) %>%
  gt() %>%
  as_raw_html()
}

```

```{r}
#| label: read

data(list=paste0("counts_", params$alignment))

if (!is.null(metadata(dds)$organism$org)) {
  library(metadata(dds)$organism$org, character.only=TRUE)
}
metadata(dds)$template_git <- packageDescription("DESdemonA")$git_last_commit

specs   <- DESdemonA::load_specs(file=paste0(params$spec, ".spec"), context=dds)

param <- DESdemonA::ParamList$new(defaults=specs$settings)
# Calling the setter without a value will pick up the default (from the spec), _and_ alert it in the markdown, and return the default
set.seed(param$set("seed"))
param$set("title", "{{{project}}}")
param$set("script", file.path(getwd(),"01_analyse.r"))
param$set("spec",  params$spec, "Using analysis plan '{}'.")
if (is.character(params$alignment)) {
  param$set("alignment", params$alignment, "Using alignment settings '{}'")
} else {
  param$set("alignment", params$param_call$alignment, "Using counts from '{}'")
}  

ddsList <- DESdemonA::build_dds_list(dds, specs)


#Which features are zero across all samples in all datasets
all_zero <- Reduce(f=`&`,
                  x=map_des(ddsList, function(x) apply(counts(x)==0,1, all)),
                  init=TRUE)
ddsList <- map_des(ddsList, function(dds) dds[!all_zero,])


## Calculate size factors, separately for each sample-set
ddsList <- mapply(
  function(samp, dds) {
    if ("subsetGenes" %in% names(samp)) {
      dds <- dds[eval(samp$subsetGenes, transform(rowData(dds), ID=row.names(dds))),]
    }
    if ("controlGenes" %in% names(samp)) {
      controlGenes <- eval(samp$controlGenes, transform(rowData(dds), ID=row.names(dds)))
      estimateSizeFactors(dds, controlGenes=controlGenes)
    } else {
      estimateSizeFactors(dds)
    }
  },
  specs$sample_sets,
  ddsList,
  SIMPLIFY=FALSE
)


param$set("baseMeanMin")
if (param$get("baseMeanMin")>0) {
  ddsList <- map_des(ddsList,
                   function(x) x[rowMeans(counts(x, normalized=TRUE)) >= param$get("baseMeanMin"),]
                   )
}

```

# Input Summary

The sample annotations are as follows:


```{r}
#| label: tbl-inputs
#| tbl-cap: Sample Annotation

table1 <- function(dds, ddsList) {
  samp <- as.data.frame(colData(dds))
  # Extra columns introduced by datasets
  extra_meta <- lapply(
    names(ddsList),
    function(x) merge(
      setNames(data.frame(ifelse(colnames(dds) %in% colnames(ddsList[[x]]), "✓", ""), row.names=row.names(colData(dds))), x),
      as.data.frame(colData(ddsList[[x]])[, setdiff(names(colData(ddsList[[x]])), names(colData(dds)))]),
      by=0, all.x=TRUE
    )[,-1,drop=FALSE]
  )
  
  extra_meta <- extra_meta[sapply(extra_meta, function(x) ncol(x)!=0 & nrow(x)!=0)]
  # Generate tab_spanner parameters for each datasets extra columns
  ts_extra <- lapply(names(extra_meta),
                    function(x) list(name=x,
                              cols=1:ncol(extra_meta[[x]]))
                    )
  # Cummulatively offset the columns to be spanned
  ts_extra <- Reduce(function(ts, x) {
    x$cols <- x$cols+max(ts$cols)
    x},
    ts_extra,
    accumulate=TRUE)
  
  df <- cbind(samp, do.call(cbind, unname(extra_meta)))
  clabel <- setNames(as.list(names(df)), 1:ncol(df))
  names(df) <- names(clabel)
  gt(data=df) %>%
    cols_label(.list=clabel) %>%
    tab_spanner(label="Metadata",
                columns=seq_along(samp)) %>%
    Reduce(f=function(gti, x) tab_spanner(gti, label=x$name,columns=x$cols + ncol(samp)), x=ts_extra, init=.)
}

table1(dds, ddsList) %>%
  as_raw_html()

ddsList <- map_des(ddsList, function(x) DESdemonA::add_dim_reduct(x))

for (assay in c("counts", "norm", "vst")) {
  DESdemonA::write_assay(ddsList, assay=assay, path=params$res_dir)
}

save(ddsList, file=file.path("data",paste0("pca_", params$alignment, "_", params$spec, ".rda")))

```

We may examine the samples in different combinations, and leave out
certain samples.  In the above table, the columns under the 'In
Subset' group indicate these combinations are listed, and the
samples' inclusions are indicated.  Each of those combinations may be analysed in
potentially several ways, as formulated here:


```{r}
#| label: formulae

for (dataset in names(specs$sample_sets)) {
  mdls <- lapply(specs$sample_sets[[dataset]]$models, function(x) x$design)
  mdls <- mdls[sapply(specs$sample_sets[[dataset]]$model, function(x) "comparisons" %in% names(x))]
  for (mdl in names(mdls)) {
    cat("\n\n####", dataset, mdl, "{-}\n\n")
    print(extract_eq(lm(
      update(mdls[[mdl]], Expression ~ .),
      data=cbind(data.frame(colData(ddsList[[dataset]]), check.names=FALSE), Expression=1:ncol(ddsList[[dataset]])),
      wrap=TRUE
    )
    ))
  }
}

```


We also have the following defaults set.  Whenever they are used
or changed in the, analysis that will be highlighted in the text
alongside a 'wrench' icon:


```{r}
#| label: tbl-settings
#| tbl-cap: Analysis Settings

data.frame(
  Option=names(specs$settings),
  Value=sapply(specs$settings, deparse)
) %>%
  gt() %>%
  as_raw_html()

```


# Description of Visualisations

## Overall Heatmaps

These visualisations are carried out blind to the experimental
design. For the heatmap @fig-qc-1-Heatmap, we select a subset of
genes removing the (assumed uninformative) genes with flat expression
levels across the whole sample set. We'd expect samples from the same
experimental group to cluster together, in the sense that they are on
the same branch of the tree. But this is a transcriptome-wide picture,
and even if the clustering is not perfect, there will still be genes
that are consistent with the expected expression profiles, and they
will be revealed in the differential analysis.

The second heatmap @fig-qc-1-sampDist focuses on the pairwise
similarity of samples, using a slightly different metric than the
first, to emphasise that there's not one way of calculating how
similar two samples are, transcriptome-wide.  The colours in the first
heatmap are a gene's expression in a sample, relative to its average
expression.  The second heatmap represents the similarity of each pair
of samples.

## Principal Component Scatterplots

The come a series of (pairs of) scatter plots of the first two
Principal Components - a way of summarising all the genes down to just
two 'metagenes' that attempt to encapsulate the variability across the
samples in as faithful way as possible. The first of each pair
(e.g. @fig-qc-1-color-1) takes the PCs from the unadjusted data,
and colours them by each of your covariates/treatments in turn. They
alternate with 'focussed' PC adjustments, such as
@fig-qc-1-focus-1 which regress out ('normalise away') the
effect of all experimental factors except the one being focussed
on. This 'realigns' the principal components, and may be useful if
there's one experimental factor (e.g. strong batch effect) that
dominates the transcriptional profile and masking a less dominant
factor: we then colour by the one experimental factor that hasn't been
regressed out, hopefully allowing its effect to emerge.

## PC-Covariate assocation

We then revert to the PCA picture that is unadjusted for any
covariates. In this plot the columns explore the lower-order principal
components (which although not providing a universally strong signal
transcriptome wide, might still detect biology that is affecting a
smaller subset of genes, or a reasonably large set of genes to a
lesser extent than happens in the higher-order PCS). The rows
represent the experimental factors we have recorded, and the depth of
colour of the cell records the degree of 'association' that factor has
with the give PC (and grey says there was no statistically significant
association.) This being an automated report, we've had to make
various compromises so please don't read too much into this chart: it
primarily guides which other plots we should generate to look at how
samples associate with each other.

## Partial Residuals of PCAs

We use that information to derive the remaining plots in this section,
where we these components in the context of the various experimental
factors, producing 'partial residual plots' of two selected PCs, being
selected by the criteria of: 'best' being the PC most strongly
associated with the experimental factor; 'first' being the
highest-order PC associated with the experimental factor; and then
resorting to the first and second PCs in the case where we haven't
already generated two plots. 'Partial residual plots' are effectively
plots of the PC against the experimental factor, where once again we
have 'regressed out' the effect of all _other_ experimental factors at
play.


```{r}
#| label: fig-qc-visualisation

                       
param$set("top_n_variable")
param$set("clustering_distance_rows")
param$set("clustering_distance_columns")
param$set("pc_x", 1)
param$set("pc_y", 2)

for (dname_i in seq_along(ddsList)) {
  dname <- names(ddsList)[dname_i]
  this_dds <- ddsList[[dname]]
  cat("\n# ", dname, "\n", sep = "")
  var_stab <- assay(this_dds, "vst")
  top <- order(apply(var_stab, 1, sd), decreasing = TRUE)[1:param$get("top_n_variable")]
  models_for_qc <- metadata(this_dds)$models[sapply(metadata(this_dds)$models, 
                                               "[[", "plot_qc")]
  cat("\n## Overall Heatmap of Variable Genes in ", dname, "\n", sep = "")
  plotDat <- var_stab[top, ]
  plotDat <- plotDat - rowMeans(plotDat)
  vars <- DESdemonA:::get_terms(this_dds)
  colDat <- as.data.frame(colData(this_dds))
  colnames(plotDat) <- rownames(colDat)
  pl <- ComplexHeatmap::Heatmap(
    plotDat, name = "Mean Centred", 
    column_title = "Samples", row_title = "Genes", col = DESdemonA:::sym_colour(plotDat), 
    clustering_distance_columns = param$get("clustering_distance_columns"), 
    clustering_distance_rows = param$get("clustering_distance_rows"), 
    heatmap_legend_param = list(direction = "horizontal"), 
    top_annotation = ComplexHeatmap::HeatmapAnnotation(
      df = colDat[vars$fixed], 
      col = metadata(colData(this_dds))$palette$Heatmap[vars$fixed]), 
    show_row_names = FALSE, show_column_names = TRUE)
  do_plot(pl,
          label= paste("qc", dname_i, "Heatmap"),
          caption=paste("Heatmap of variable genes,", dname))
  cat("\n\n---\n\n")

  poisd <- PoiClaClu::PoissonDistance(t(counts(this_dds, normalized = TRUE)))
  samplePoisDistMatrix <- as.matrix(poisd$dd)
  if (length(vars$fixed) > 1) {
    rownames(samplePoisDistMatrix) <- Reduce(
      function(...) paste(..., sep = "_"), colDat[vars$fixed])
  } else {
    rownames(samplePoisDistMatrix) <- colDat[[vars$fixed]]
  }
  if (is.null(vars$groups)) {
    colnames(samplePoisDistMatrix) <- row.names(colDat)
  } else {
    if (length(vars$groups) > 1) {
      rownames(samplePoisDistMatrix) <- Reduce(function(...) paste(..., sep = "_"), colDat[vars$groups])
    } else {
      rownames(samplePoisDistMatrix) <- colDat[[vars$groups]]
    }
  }
  pl <- ComplexHeatmap::Heatmap(
    samplePoisDistMatrix,
    col = circlize::colorRamp2(
      c(0,  max(samplePoisDistMatrix)),
      colors = c("white", "red")), 
    name = "Poisson Distance", clustering_distance_rows = function(x) {poisd$dd},
    clustering_distance_columns = function(x) { poisd$dd},
    heatmap_legend_param = list(direction = "horizontal"), 
    top_annotation = ComplexHeatmap::HeatmapAnnotation(df = colDat[vars$fixed], 
                                                       col = metadata(colData(this_dds))$palette$Heatmap[vars$fixed]))
  do_plot(pl,
          label= paste("qc", dname_i, "sampDist"),
          caption=paste("Heatmap of sample distances,", dname))
  cat("\n\n---\n\n")

  pc <- as.matrix(colData(this_dds)$.PCA)
  percentVar <- metadata(colData(this_dds)$.PCA)$percentVar
  is_vary <- sapply(colDat[vars$fixed], function(v) length(unique(v)) != 1)
  do_focus <- length(vars$fixed) > 1
  do_batch <- FALSE
  if (do_focus) {
    sample_gene_factor <- DESdemonA:::residual_heatmap_transform(
      assay(this_dds, 
            "vst"), colData(this_dds), models_for_qc[[1]]$design)
    if ("batch" %in% names(models_for_qc[[1]])) {
      batch_vars <- all.vars(models_for_qc[[1]]$batch)
      pc_batch <- prcomp(
        t(assay(this_dds, "vst")) -
          apply(sample_gene_factor$terms[, , batch_vars, drop = FALSE], 1:2, sum),
        scale = FALSE)
      pc_resid <- list(
        pc = pc_batch$x,
        percent = round(100 * pc_batch$sdev^2/sum(pc_batch$sdev^2)))
      do_focus <- FALSE
      do_batch <- TRUE
    }
    else {
      pc_resid <- lapply(
        dimnames(sample_gene_factor$terms)[[3]], 
        function(fac) {
          pc <- prcomp(
            sample_gene_factor$terms[, , fac, drop = TRUE] + sample_gene_factor$resid, 
            scale = FALSE)
          list(pc = pc$x, percent = round(100 * pc$sdev^2/sum(pc$sdev^2)))
        })
      names(pc_resid) <- dimnames(sample_gene_factor$terms)[[3]]
    }
  }
  cat("\n## Principal Component Scatterplots of PCs ", param$get("pc_x"), " and ", param$get("pc_y"), 
      " coloured by covariate", "\n\n", sep = "")
  do_labels <- nrow(colDat) < 10
  for (covar in vars$fixed[is_vary]) {
    covar_i <- match(covar, vars$fixed[is_vary])
    pc.df <- data.frame(
      PC1 = pc[, param$get("pc_x")],
      PC2 = pc[, param$get("pc_y")], 
      col = colDat[[covar]], sample = rownames(colDat))
    pl <- ggplot(pc.df, aes(x = PC1, y = PC2, colour = col)) + 
      geom_point(size = 5) +
      xlab(paste0("PC ", param$get("pc_x"), ": ",  percentVar[param$get("pc_x")], "% variance")) +
      ylab(paste0("PC ",param$get("pc_y"), ": ", percentVar[param$get("pc_y")], "% variance")) +
      labs(colour = covar) + 
      theme_bw()
    if (is.numeric(colDat[[covar]])) {
      pal_col <- metadata(colData(this_dds))$palette$ggplot[[covar]]
      pl <- pl + scale_colour_gradient(low = pal_col[1], high = pal_col[2])
    }  else {
      pl <- pl + scale_colour_manual(values = metadata(colData(this_dds))$palette$ggplot[[covar]])
    }
    if (do_labels) {
      pl <- pl + geom_text_repel(aes(label = sample))
    }
    do_plot(pl,
            label=paste("qc", dname_i, "color", covar_i),
            caption=paste("Coloured by ", covar, "in", dname)
            )
    cat("\n\n---\n\n")
    if (do_batch) {
      pc.df <- data.frame(
        PC1 = pc_resid$pc[, param$get("pc_x")],
        PC2 = pc_resid$pc[, param$get("pc_y")],
        col = colDat[[covar]],
        sample = rownames(colDat))
      pl <- ggplot(pc.df, aes(x = PC1, y = PC2, colour = col)) + 
        geom_point(size = 5) +
        xlab(paste0("PC ", param$get("pc_x"), ": ", pc_resid$percent[param$get("pc_x")], "% variance")) + 
        ylab(paste0("PC ", param$get("pc_y"), ": ", pc_resid$percent[param$get("pc_y")],  "% variance")) +
        labs(colour = covar) +
        theme_bw()
      if (is.numeric(colDat[[covar]])) {
        pal_col <- metadata(colData(this_dds))$palette$ggplot[[covar]]
        pl <- pl + scale_colour_gradient(low = pal_col[1], high = pal_col[2])
      }  else {
        pl <- pl + scale_colour_manual(values = metadata(colData(this_dds))$palette$ggplot[[covar]])
      }
      if (do_labels) {
        pl <- pl + geom_text_repel(aes(label = sample))
      }
      do_plot(pl,
              label= paste("qc", dname_i, "correct", covar_i),
              caption=paste0("Coloured by ", covar, ", correcting for ", 
                             paste(batch_vars, collapse = ", "), "in", dname)
              )
      cat("\n\n---\n\n")
    }
    if (do_focus && covar %in% names(pc_resid)) {
      pc.df$PC1 <- pc_resid[[covar]]$pc[, param$get("pc_x")]
      pc.df$PC2 <- pc_resid[[covar]]$pc[, param$get("pc_y")]
      pl <- ggplot(pc.df, aes(x = PC1, y = PC2, colour = col)) + 
        annotate("text", x = Inf, y = -Inf, label = "PROOF ONLY", 
                 hjust = 1.1, vjust = -1.1, col = "white", cex = 6, 
                 fontface = "bold", alpha = 0.8) + geom_point(size = 5) + 
        xlab(paste0("PC ", param$get("pc_x"), ": ", pc_resid[[covar]]$percent[param$get("pc_x")], "% variance")) +
        ylab(paste0("PC ", param$get("pc_y"), ": ", pc_resid[[covar]]$percent[param$get("pc_y")], "% variance")) + 
        labs(colour = covar) + theme_bw()
      if (is.numeric(colDat[[covar]])) {
        pal_col <- metadata(colData(this_dds))$palette$ggplot[[covar]]
        pl <- pl + scale_colour_gradient(low = pal_col[1], high = pal_col[2])
      } else {
        pl <- pl + scale_colour_manual(values = metadata(colData(this_dds))$palette$ggplot[[covar]])
      }
      if (do_labels) {
        pl <- pl + geom_text_repel(aes(label = sample))
      }
      do_plot(pl,
              label=paste("qc", dname_i, "focus", covar_i),
              caption=paste("Focussed on ", covar, "in", dname)
              )
      cat("\n\n---\n\n")
    }
  } # next covariate

  cat("\n## PC-Covariate Associations\n\n")
  pc_frame <- expand.grid(sample = rownames(colDat), PC = 1:ncol(pc))
  pc_frame$coord <- as.vector(pc)
  pc_frame <- cbind(pc_frame, colDat)
  fitFrame <- colDat
  yvar <- make.unique(c(colnames(fitFrame), "y", sep = ""))[ncol(fitFrame) + 1]
  for (model_name in names(models_for_qc)) {
    fml <- update(metadata(this_dds)$models[[model_name]]$design, 
                  paste(yvar, "~ ."))
    plotFrame <- expand.grid(
      Covariate = attr(terms(fml), "term.labels"),
      PC = 1:ncol(pc))
    plotFrame$Assoc <- NA
    npc <- ncol(pc)
    fit_selected <- list()
    for (ipc in 1:npc) {
      fitFrame[[yvar]] <- pc[, ipc]
      fit1 <- lm(fml, data = fitFrame)
      fit_selected[[ipc]] <- MASS::stepAIC(fit1, trace = 0)
      ind <- attr(terms(fit_selected[[ipc]]), "term.labels")
      if (length(ind)) {
        fit0 <- as.data.frame(anova(fit_selected[[ipc]]))
        rss <- fit0[ind, "Sum Sq"]/sum(fit0[, "Sum Sq"])
        ind_pc <- plotFrame$PC == ipc
        ind_fit <- match(ind, plotFrame$Covariate[ind_pc])
        plotFrame$Assoc[ind_pc][ind_fit] <- rss
      }
    }
    plotFrame$wrap <- (plotFrame$PC - 1)%/%20
    plotFrame$wrap <- paste0("PCs ", plotFrame$wrap * 20 + 
                                       1, "-", min((plotFrame$wrap + 1) * 20, npc))
    plotFrame$PC <- sprintf("%02d", plotFrame$PC)
    pl <- ggplot(plotFrame, aes(x = PC, y = Covariate, fill = Assoc)) + 
      geom_raster() +
      facet_wrap(~wrap, scales = "free_x",ncol = 1) +
      scale_fill_gradient2(low = "#4575b4",mid = "grey90", high = "#d73027") +
      theme_bw() + 
      theme(aspect.ratio = length(unique(plotFrame$Covariate))/min(20, npc))
    do_plot(pl,
            label=paste("qc", dname_i, "covar"),
            caption=paste("Covariate association in", dname)
            )
    
    cat("\n## Partial Residuals of Associated PCs ", model_name, "\n\n", sep = "")
    factors_1 <- attr(terms(fit1), "factors")
    many_levels <- sapply(colDat, function(x) length(unique(x)))
    for (covar in unique(plotFrame$Covariate)) {
      covar_i <- match(covar, unique(plotFrame$Covariate))
      this_covar <- subset(plotFrame, Covariate == covar & 
                                        !is.na(Assoc) & PC != PC[nrow(plotFrame)])
      if (nrow(this_covar) == 0) 
        next
      pc_first <- as.integer(as.character(this_covar$PC[1]))
      this_covar <- this_covar[order(this_covar$Assoc, 
                                     decreasing = TRUE), ]
      pc_strong <- as.integer(as.character(this_covar$PC[1]))
      pc.df <- subset(pc_frame, PC %in% c(pc_first, pc_strong))
      pc.df$label <- as.character(pc.df$PC)
      pc.df$label[pc.df$PC == pc_first] <- paste0(pc.df$label[pc.df$PC == 
                                                                pc_first], "+First")
      pc.df$label[pc.df$PC == pc_strong] <- paste0(pc.df$label[pc.df$PC == 
                                                                 pc_strong], "+Best")
      pc.df$label <- paste0(pc.df$label, " (", percentVar[pc.df$PC], 
                            "%)")
      inter_vars <- rownames(factors_1)[factors_1[, covar] !=  0]
      inter_vars <- inter_vars[order(many_levels[inter_vars], decreasing = TRUE)]
      pc.df$X <- colDat[, inter_vars[1]]
      for (ipc in unique(c(pc_first, pc_strong))) {
        pc.df$coord[pc.df$PC == ipc] <- DESdemonA:::part.resid(fit_selected[[ipc]])[, covar]
      }
      if (length(inter_vars) > 1) {
        pc.df$col <- colDat[, inter_vars[2]]
        if (length(inter_vars) == 2) {
          pl <- ggplot(pc.df, aes(x = X, y = coord, colour = col, group = col)) +
            stat_summary(fun = mean,geom = "line") +
            labs(x = inter_vars[1], y = "Residual",
                 colour = paste(inter_vars[-1], collapse = "x"))
          + theme_bw()
          if (is.numeric(pc.df$col)) {
            pal_col <- metadata(colData(this_dds))$palette$ggplot[[inter_vars[2]]]
            pl <- pl + scale_colour_gradient(low = pal_col[1], high = pal_col[2])
          }
          else {
            pl <- pl + scale_colour_manual(values = metadata(colData(this_dds))$palette$ggplot[[inter_vars[2]]])
          }
        }
        else {
          pc.df$grp <- Reduce(interaction, colDat[, inter_vars[-1]])
          pl <- ggplot(pc.df, aes(x = X, y = coord, colour = col,  group = grp)) +
            stat_summary(fun = mean, geom = "line") +
            labs(x = inter_vars[1], y="Residual",
                 colour = paste(inter_vars[-1], collapse = "x")) +
            theme_bw()
          if (is.numeric(pc.df$col)) {
            pal_col <- metadata(colData(this_dds))$palette$ggplot[[inter_vars[2]]]
            pl <- pl + scale_colour_gradient(low = pal_col[1], high = pal_col[2])
          }
          else {
            pl <- pl + scale_colour_manual(values = metadata(colData(this_dds))$palette$ggplot[[inter_vars[2]]])
          }
        }
      }
      else {
        pl <- ggplot(pc.df, aes(x = X, y = coord, colour = X, group = 1)) +
          stat_summary(fun = mean, geom = "line", colour = "black") +
          labs( x = inter_vars[1], y="Residual", colour = covar) +
          theme_bw()
        if (is.numeric(pc.df$X)) {
          pal_col <- metadata(colData(this_dds))$palette$ggplot[[inter_vars[1]]]
          pl <- pl + scale_colour_gradient(low = pal_col[1], high = pal_col[2])
        }
        else {
          pl <- pl + scale_colour_manual(values = metadata(colData(this_dds))$palette$ggplot[[inter_vars[1]]])
        }
      }
      pl <- pl + geom_point(size = 2) +
        facet_wrap(~label, scales = "free_y")
      if (do_labels) {
        pl <- pl + geom_text_repel(aes(label = sample))
      }
      do_plot(pl,
              label= paste("qc", dname_i, "covar", covar_i),
              caption=paste("Against ", covar, "in", dname)
              )
      cat("\n\n---\n\n")
    }
  }
}




```



# Terms Of Use

The Crick has a [publication
policy](https://intranet.crick.ac.uk/our-crick/library-information-services/pages/guidelines-publication)
and we expect to be included on publications, regardless of funding
arrangements. Any use of these results in publication must be
discussed with BABS regarding authorship. If not authorship then
the BABS analyst must receive a named acknowledgement. Please also
cite the following sources which have enabled the analysis to be
carried out.

# Bibliography



