---
title: Differential analysis
description: Differential analysis
categories: [Differential]
author: 
  - name: Gavin Kelly
    affiliations: BABS - Francis Crick Institute
format:
  html:
    theme: cerulean
    fontsize: 1.1em
    linestretch: 1.7
code-fold: true
execute:
  output: asis
  echo: false
code-summary: "Show the code"
params:
  section: Differential analysis

---


# Preface

We load in all the necessary additional R packages, and set up some initial parameters.


```{r}
#| label: load-diff1
#| output: false
#| warning: false
#| error: false

library(tidyverse)       # Language
library(devtools)
library(openxlsx)        # IO
library(ggrepel)         # General Vis.
library(RColorBrewer)
library(ComplexHeatmap)
library(gt)
library(emmeans)
library(DESeq2)          # Assay-specific
library(PoiClaClu)
library(rtracklayer)
library(tximport)
library(clusterProfiler)
library(ReactomePA)
library(GO.db)
library(IHW)

source("R/plot.r", local = knitr::knit_global())
source("R/markdown.r", local = knitr::knit_global())
source("R/rnaseq.r", local = knitr::knit_global())
source("R/partial.r", local = knitr::knit_global())
source("R/output.r", local = knitr::knit_global())
source("R/helpers.r", local = knitr::knit_global())

set.seed(1)

if (!isTRUE(getOption('knitr.in.progress'))) {
  params <- load_params(choose_staging("02_differential"))
}

align_params <- get_align_params(params$alignment)
rowNoun <- align_params$rowNoun %||% "gene"
RowNoun <- stringr::str_to_title(rowNoun)
if ("missing" %in% names(align_params)) {
  attr(rowNoun, "readout") <- "missingness"
  attr(RowNoun, "readout") <- "Missingness"
}

do_plot <- plot_tracker(params)

knitr::opts_chunk$set(
  warning=FALSE, error=TRUE, message=FALSE,
  dev=c("ragg_png","pdf"), out.width="100%",
  dev.args=list(ragg_png=list(scaling=1.5)),
  fig.width=14, fig.height=10, fig.process=rename_with_tag(params),
  results='asis'
)



```



```{r}
#| label: tbl-load-diff2
#| tbl-cap: Context
if (isTRUE(getOption('knitr.in.progress'))) {
  gt(data.frame(Setting=names(params),
                Value=unlist(params)))
}


```

```{r}
#| label: read-diff

base_dir <- getwd()
data(list=paste0("counts_", params$alignment))
data(list=paste0("pca_", params$alignment, "_", params$spec))

ddsList <- lapply(ddsList, function(dds) {
  imp <- metadata(dds)$impute
  if (!is.null(imp) && !is.null(imp$differential) && !imp$differential) {
    assays(dds) <- setFirstAssay(dds, setdiff(assayNames(dds), "imputed")[1])
  }
  dds
})
  
if (!is.null(metadata(dds)$organism$org)) {
  library(metadata(dds)$organism$org, character.only=TRUE)
}


specs <- load_specs(file=paste0(params$spec, ".spec"), context=dds)

param <- ParamList$new(defaults=specs$settings)
# Calling the setter without a value will pick up the default (from the spec), _and_ alert it in the markdown, and return the default
set.seed(param$set("seed", postpone=TRUE))
param$set("spec",  params$spec, "Using analysis plan '{}'.", postpone=TRUE)
param$set("alignment", params$alignment, "Using alignment settings '{}'")

dtag <- dmc_factory(ddsList, type="dataset")
mtag <- dmc_factory(ddsList, type="model")

```

# Find differential `{r} rowNoun`s

 We use DESeq2 [@pkg_DESeq2] to find differential `{r} rowNoun`s using the
negative binomial distribution to model counts, with IHW [@IHW2016]
for multiple-testing correction with greater power than
Benjamini-Hochberg, and ashr [@pkg_ashr] for effect-size shrinkage
to ensure reported fold-changes are more robust.


```{r}
#| label: differential

param$set("alpha", postpone=TRUE)
param$set("lfcThreshold", postpone=TRUE)
param$set("filterFun", postpone=TRUE)
param$set("LRT_effect")

rds_name <- file.path("data", paste0(basename(params$spec),"_x_", params$alignment, ".rds"))

if (interactive() && file.exists(rds_name)) {
  dds_model_comp <- decompress_dmc(readRDS(rds_name))
  # In an interactive session, you may want to skip the rest of this code chunk if you are
  # sure that the saved rds object represents the latest analysis.
}


## For each dataset, fit all its models
for (i in sub("\\.emmc$", "", ls("package:emmeans", pattern="*.emmc")))
  assign(paste0("my_", i,".emmc"), negate_emmc(paste0(i, ".emmc")))
dds_model_comp <- lapply(
  ddsList,
  function(dds) fit_models(
    dds=dds,
    param=param,
    minReplicatesForReplace = Inf
  )
)


## Now put results in each 3rd level mcols(dds)$results
dds_model_comp <- map_depth(
  dds_model_comp,
  .depth=3,
  function(dds) get_result(dds,
                         filterFun=param$get("filterFun"),
                         alpha=param$get("alpha"),
                         lfcThreshold=param$get("lfcThreshold"),
                         LRT_effect=param$get("LRT_effect", metadata(dds)$models)
                         )
)



local({
  dds_model_comp <- compress_dmc(dds_model_comp)
  saveRDS(dds_model_comp, file=rds_name)
  save(dds_model_comp, file=sub("rds$", "rda", rds_name))
})


xl_files <- write_results(dds_model_comp,
                         param,
                         params
                         )
iwalk(xl_files,
     ~cat("\n\n<a class=\"download-excel btn btn-primary\" href=\"", attr(.x, "link"), "\"> Open Spreadsheet '", .y,"'</a>", sep="")
     )

#write_all_results(dds_model_comp, dir=params$res_dir)

```


## Summary Tables 

Here we summarise the results of the differential testing. Each
sample-set gets its own table (flick between tabs to choose which),
within which the different models, and their null hypotheses, are
enumerated.

In the 'Significant' column we tally the number of significant
`{r} rowNoun`s (for pairwise comparisons, separated into up or down, where
A-B being labelled up means `{r} noun_to_readout(rowNoun)` is higher in A; for omnibus
comparisons, a broad categorisation of the most extreme groups, as
described above.) We also tally the total number of `{r} rowNoun`s
exhibiting that behaviour (ie not necessarily statistically
signficant) - these might not always add up to the same value, as
there is an independent filter of low-signal `{r} rowNoun`s whose effect
varies from comparison to comparison.

Our naming convention for a comparison often looks like `A-B|M` -
again, because this is an semi-automated report, this mightn't be
optimal, but the rationale is `A-B` refers to the two experimental
conditions being compared (a 'upward' fold change will be reported
if the `{r} noun_to_readout(rowNoun)` is higher in `A` than it is in `B`). The `|M` part
refers to the fact that we may have looked at that `A-B` comparison
conditional upon some other variable (which in this example takes the
value `M`):  anything before the `|` symbol identifies the
experimental conditions that are being contrasted, and the part after
the `|` identifies any stratification/conditioning of the data.

In @fig-summaryp-1 we present some diagnostic plots of the observed
p-values. Theoretically we want the pink part of the plot to have a
solid peak on the left, and then make a reasonable plateau across
higher p-values: gross deviations from this can indicate an absence of
differential behaviour (lack of left-peak) or a model
mis-specification that might need investigating. The blue tracks
indicate p-values corresponding to `{r} rowNoun`s with very low `{r} noun_to_readout(rowNoun)`, which
would typically struggle to provide evidence of differential
behaviour, and so show an example of what a 'bad' pink trace might
look like.

```{r}
#| label: tbl-summary
#| tbl-cap: Summaries

summaries <- dds_model_comp %>%
  map_dmc(summarise_results) %>%
  map_dmc(function(x) {bind_rows(x, .id="Comparison")}, depth="model") %>%
  map_dmc(function(x) {bind_rows(x, .id="Model")}, depth="dataset") %>%
  map(function(x) {
    levels(x$Group) <- sub(".*<(.+<.+<0)$", "\\1", levels(x$Group))
    levels(x$Group) <- sub("^(0<.+?<.+?)<.*$", "\\1", levels(x$Group))
    levels(x$Group) <- sub("(.*?<)(.*0.*)(<.*)", "\\10\\3", levels(x$Group))
    x
  }
  )

unclassed <- lapply(summaries, 
                    function(s) {
                      s %>%
                        subset(Group %in% c('Zero Count', 'Low Count', 'Outlier')) %>%
                        group_by(Model, Group) %>% summarise(N=paste(unique(Total), collapse=";")) %>%
                        ungroup() %>%
                        pivot_wider(id_cols=Model, names_from=Group, values_from=N)
                      })

pval_frame <- dds_model_comp %>%
  map_dmc(~as.data.frame(mcols(.x)$results)) %>%
  map_dmc(function(x) {bind_rows(x, .id="Comparison")}, depth="model")



## for (dataset in names(summaries)) {
##   #options(htmltools.preserve.raw=TRUE)
##   #cat("### ", dataset, " \n", sep="")
##   #caption=paste0("Size of gene-lists for ", dataset),
##   summaries[[dataset]] %>%
##     gt(groupname_col="Model") %>%
##     tab_header(title="Genelist summary",
##                subtitle=dataset) %>%
##     as_raw_html()
## }


for (dataset in names(summaries)) {
  res <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      paste0("#| label: tbl-summary-",dataset),
      paste0("#| tbl-cap: ",dataset),
      "summaries[[dataset]] %>%",
      "subset(!Group %in% c('Zero Count', 'Low Count', 'Outlier')) %>% ",
      "gt(groupname_col=\"Model\") %>%",
      "tab_style(",
      "  style = cell_fill(color = '#bee2e9'),",
      "  locations = cells_row_groups()",
      ") %>% ",
      paste0("tab_header(title=\"",RowNoun,"List summary\", subtitle=dataset)"),
      "```",
      "",
      ""
    ), envir = environment(), quiet = TRUE)
  cat(res, sep = '\n\n')

  if (nrow(unclassed[[dataset]])>0) {
    res <- knitr::knit_child(
      text = c(
        "```{r}",
        "#| echo: false",
        paste0("#| label: tbl-summaryfit-",dataset),
        paste0("#| tbl-cap: ",dataset, " model-fit stats"),
        "unclassed[[dataset]] %>%",
        "gt() %>%",
        "tab_header(title=\"Model-fit summary\", subtitle=dataset)",
        "```",
        "",
        ""
      ), envir = environment(), quiet = TRUE)
    cat(res, sep = '\n\n')
  }
}

```


```{r}
#| label: fig-summaryp

for (dataset in names(summaries)) {
  for (model in names(pval_frame[[dataset]])) {
    pl <- ggplot(pval_frame[[dataset]][[model]], aes(x=pvalue, fill=baseMean<1)) +
      geom_density(alpha=0.5) +
      facet_wrap(~Comparison) +
      theme_bw()
    do_plot(pl,
            label="summaryp",
            caption=paste("p-Diagnostic for", dataset, model)
            )
    cat("\n\n---\n\n")
  }
}
```

# Datasets and Models

As in the exploratory analysis, we may examine the samples in different combinations, and leave out
certain samples.  In the above table, the columns under the 'In
Subset' group indicate these combinations are listed, and the
samples' inclusions are indicated.

Each of those combinations may be analysed in
potentially several ways, as formulated here:


```{r}
#| label: formulae

for (dataset in names(ddsList)) {
  met <- metadata(ddsList[[dataset]])
  dmc_heading(ddsList, dataset=dataset, model=NULL, depth="##")
  mdls <- lapply(met$models, function(x) x$design)
  mdls <- mdls[sapply(met$models, function(x) "comparisons" %in% names(x))]
  for (mdl in names(mdls)) {
    dmc_heading(ddsList, dataset=dataset, model=mdl, depth="##")
    cat("`",
        Reduce(paste, sub("~", paste0(noun_to_readout(RowNoun)," ~ "), deparse(mdls[[mdl]]))),
        "`\n\n"
        )
  }
}

```


# Scatterplot of differential `{r} rowNoun`s

Here we plot the fold-change across all `{r} rowNoun`s on the vertical axis,
against their overall `{r} noun_to_readout(rowNoun)` on the horizontal. Statistically
significant `{r} rowNoun`s are highlighted in colour, and we attempt to
auto-label as many `{r} rowNoun` symbols as might be legible.

The fold-changes are regularised ('shrunk') to reflect their
reliability: `{r} rowNoun`s with large variability between replicates are
moved towards the horizontal axis, to de-emphasise them and to
provide a more reliable prediction of behaviour in future
replications.

Again, select the dataset you wish to examine via the tabs - there
are sub-tabs for different experimental designs, if any
(e.g. removing/ignoring batch effects).


```{r}
#| label: fig-differential-MA
first <- TRUE
for (dataset in names(dds_model_comp)) {
  dmc_heading(dds_model_comp, dataset=dataset, model=NULL, depth="##")
  met <- metadata(dds_model_comp[[dataset]][[1]][[1]])$dmc
  for (mdl in names(dds_model_comp[[dataset]])) {
    met <- metadata(dds_model_comp[[dataset]][[mdl]][[1]])$dmc
    dmc_heading(dds_model_comp, dataset=dataset, model=mdl, depth="###")
    this_mdl <- dds_model_comp[[dataset]][[mdl]]
    if (!is.null(metadata(this_mdl[[1]])$model$external_list)) {
      external_list <- read.csv(file.path(base_dir, metadata(this_mdl[[1]])$model$external_list))
      external_list <- subset(external_list, ID %in% row.names(this_mdl[[1]]))
    } else {
      external_list <- NULL
    }
    for (comp_id in names(this_mdl)) {
      res <- as.data.frame(mcols(this_mdl[[comp_id]])$results)
      md <- metadata(this_mdl[[comp_id]])
      pal <- RColorBrewer::brewer.pal(9, "Set1")
      res$class <- ifelse(grepl("\\*", res$class), "Sig", "Not")
      if (is_formula(md$comparison)) {
        yax <- "Largest fold change"
      } else {
        yax <- "Log fold change (shrunk)"
      }
      pl <- ggplot(res,
                  aes(x=baseMean,
                      y=shrunkLFC,
                      colour=ifelse(class == "Not", "#aaaaaa","#000000"),
                      size=ifelse(class == "Not", 0.5 , 2),
                      label=symbol
                      )
                  ) +
        geom_point() +
        scale_x_log10()  + scale_size_identity() +
        labs(x="Mean of normalised counts", y=yax)+ theme_bw()
      colour_scale <- list(labels = c("Not", "Significant"),
                           breaks = c("#aaaaaa", "#000000"))
      is_sig <- res$class != "Not"
      ind <- which(is_sig)[order(res$shrunkLFC[is_sig])]
      top_ind_ind <- 1:min(5,sum(is_sig))
      ind <- ind[unique(c(top_ind_ind, sum(is_sig)+1-top_ind_ind))]
      if (!is.null(external_list)) {
        dat <- res[external_list$ID,]
        if (length(unique(external_list$Comparison)) <= length(pal)) {
          col <- pal[match(external_list$Comparison, unique(external_list$Comparison))]
        } else {
          col <- rep(pal[1], nrow(dat))
        }
        colour_scale$labels <- c(colour_scale$labels, unique(external_list$Comparison))
        colour_scale$breaks <- c(colour_scale$breaks, pal[seq_along(unique(external_list$Comparison))])
        pl <- pl +
          ggrepel::geom_text_repel(aes(colour=col), size=3, data=res[external_list$ID,], max.overlaps=Inf, show_guide = FALSE) +
          geom_point(aes(colour=col), data=res[external_list$ID,])
        ind <- ind[!row.names(res)[ind] %in% external_list$ID]
      }
      pl <- pl+ggrepel::geom_text_repel(size=3, data=res[ind,], show_guide = FALSE) 
      pl <- pl +
        scale_colour_identity(labels=colour_scale$labels, breaks=colour_scale$breaks, guide = "legend") +
        labs(colour="Class")
      do_plot(pl,
              label="MA",
              caption=paste0("MA plot of differential ",rowNoun,"s ", sub(".*\\t", "", comp_id)),
              preview=first
              )
      first <- FALSE
      cat("\n\n---\n\n")
    } # next comparison
  }   # next model
}     # next dataset

```


# Differential Heatmaps

Here we present heatmaps of the differential `{r} RowNoun`Lists.  Note
that these will, by definition, divide the samples along lines
of the experimental groups, so should be interpreted differently
from the QC heatmaps which were blind to the experimental design.

```{r}
#| label: fig-differential-heatmap

param$set("clustering_distance_rows", postpone=TRUE)
param$set("clustering_distance_columns")
for (dataset in names(dds_model_comp)) {
  dmc_heading(dds_model_comp, dataset=dataset, model=NULL, depth="##")
  nice_legends <- separate_legend(ddsList[[dataset]])
  pl <- do.call(gridExtra::arrangeGrob,c(
    nice_legends,
    ncol=ceiling(sqrt(length(nice_legends)))))
  do_plot(pl,
          label= "legend-a",
          caption=paste("Metadata colour-codings for",
                        metadata(ddsList[[dataset]])$dmc$dataset_name %||% metadata(ddsList[[dataset]])$dmc$dataset
                        ))
  for (mdl_name in names(dds_model_comp[[dataset]])) {
    dmc_heading(dds_model_comp, dataset=dataset, model=mdl_name, depth="###")
    this_mdl <- dds_model_comp[[dataset]][[mdl_name]]
    first_done <- FALSE
    if (!is.null(metadata(this_mdl[[1]])$model$external_list)) {
      external_list <- read.csv(file.path(base_dir, metadata(this_mdl[[1]])$model$external_list))
      external_list <- subset(external_list, ID %in% row.names(this_mdl[[1]]))
    } else {
      external_list <- NULL
    }
    for (comp_id in c(names(this_mdl), unique(external_list$Comparison))) {
      this_comp <- this_mdl[[comp_id]]
      comp_name <- sub(".*\\t", "", comp_id)
      is_external <- !(comp_id %in% names(this_mdl))
      if (is_external) {
        ind <- row.names(this_mdl[[1]]) %in% external_list$ID[external_list$Comparison==comp_id]
        this_comp <- this_mdl[[1]]
        comp_type <- " "
      } else {
        ind <- grepl("\\*$", mcols(this_comp)$results$class)
        comp_type <- "-differential "
        if (!any(ind)) {
          next
        }
      }
      comp <- metadata(this_comp)$comparison
      fml <- metadata(this_comp)$model$design
      mdl <- metadata(this_comp)$model
      if ("mat" %in% names(mdl)) {
        mmat <- mdl$mat
      } else {
        mmat <- model.matrix(mdl$design, colData(this_comp))
      }
      var_roles <- list(all.vars(fml))
      if (!is_external && "baseline_contrast" %in% names(attributes(metadata(this_comp)$comparison))) {
        var_roles <- c(emmeans:::.parse.by.formula(attr(comp, "spec"))[c("by", "rhs")], var_roles)
        weights <- as.vector(attr(metadata(this_comp)$comparison, "baseline_contrast") %*% MASS::ginv(mmat))
        baseline_str <- paste("baseline", noun_to_readout(rowNoun))
      } else {
        weights <- rep(1/nrow(mmat), times=nrow(mmat))
        baseline_str <- paste("average", noun_to_readout(rowNoun))
      }
      tidied_data <- tidy_significant_dds(this_comp, ind, unique(unlist(var_roles)), weights=weights)
      pdat <- tidied_data$pdat
      pdat[sapply(pdat, is.character)] <- lapply(pdat[sapply(pdat, is.character)], 
                                                as.factor)
      colnames(tidied_data$mat) <- rownames(pdat)
      if (length(var_roles$by)==0) {
        col_split <- NULL
      } else {
        col_split <- apply(sapply(pdat[, var_roles$by, drop=FALSE], as.character), 1, paste, collapse=" ")
      }

      ha <- ComplexHeatmap::HeatmapAnnotation(
        df=pdat,
        col=metadata(colData(this_comp))$palette$Heatmap[names(pdat)],
        show_legend=FALSE)
      pl <- ComplexHeatmap::Heatmap(
        tidied_data$mat,
        col=sym_colour(na.omit(as.vector(tidied_data$mat))),
        heatmap_legend_param = list(direction = "horizontal" ),
        name=comp_name,
        cluster_columns = FALSE,
        show_column_names = TRUE,
        clustering_distance_rows=param$get("clustering_distance_rows"),
        column_split = col_split,
        top_annotation = ha,
        row_names_gp = gpar(fontsize = 6),
        show_row_names = nrow(tidied_data$mat)<100)
      do_plot(pl,
              label="Heatmap",
              caption=paste0("Heatmap on ", comp_name, comp_type, rowNoun,"s. White for ", baseline_str),
              height_mult=nrow(tidied_data$mat)/30,
              max_height=15, min_height=5)
      cat('\n\n---\n\n')
      if (do_residual_transform <- FALSE) {#length(all.vars(fml))>1) {
        part_resid <- partialise(tidied_data$mat, pdat, fml)
        term_names <- intersect(dimnames(part_resid$terms)[[3]], var_roles$rhs %||% unique(unlist(var_roles)))
        for (term_name in term_names) {
          tdat <- t(part_resid$terms[,,term_name, drop=TRUE]) +part_resid$const + t(part_resid$resid)
          pl <- ComplexHeatmap::Heatmap(
            tdat,
            col=sym_colour(tdat),
            heatmap_legend_param = list(direction = "horizontal" ),
            name=paste(comp_name,term_name),
            cluster_columns = FALSE,
            show_column_names = TRUE,
            column_split = col_split,
            clustering_distance_rows=param$get("clustering_distance_rows"),
            top_annotation = ha,
            row_names_gp = gpar(fontsize = 6),
            show_row_names = nrow(tidied_data$mat)<100)
          do_plot(pl,
                  label="focussed",
                  caption=paste0("Heatmap on differential ",rowNoun,"s ", comp_name, ", ", term_name, "-focussed")
                  )
          cat('\n\n---\n\n')
        }
      }
    } # next comparison
  } # next model
} #next dataset

```


# Differential Profiles

We can also plot summaries of clusters of differential `{r} rowNoun`s. Here we
split `{r} noun_to_readout(rowNoun)` data of the differential `{r} rowNoun`s into twelve clusters,
having first removed the participant-to-participant variability, and
stochastic noise. We aggregate `{r} noun_to_readout(rowNoun)` within each cluster, and
then plot the profiles per cluster - the clusters are ordered in
decreasing number of constituent `{r} rowNoun`s, and their cardinality is
encoded into the name of the cluster, e.g. `|1|=1000` means this is
the largest cluster, and represents 1000 `{r} rowNoun`s.


```{r}
#| label: fig-differential-curve

param$set("gene_clust")


exists_profile_plot <- FALSE
all_clusters <- list()
cluster_ind <- counter(post_fn=to_letter)

for (dataset_id in names(dds_model_comp)) {
  dmc_heading(dds_model_comp, dataset=dataset_id, model=NULL, depth="##")
  for (model_id in names(dds_model_comp[[dataset_id]])) {
    dmc_heading(dds_model_comp, dataset=dataset_id, model=model_id, depth="###")
    if (!is.null(metadata(this_mdl[[1]])$model$external_list)) {
      external_list <- read.csv(file.path(base_dir, metadata(this_mdl[[1]])$model$external_list))
      external_list$Cluster[is.na(external_list$Cluster)] <- external_list$ID[is.na(external_list$Cluster)]
    } else {
      external_list <- NULL
    }
    
    for (comp_id in c(names(dds_model_comp[[dataset_id]][[model_id]]), unique(external_list$Comparison))) {
      is_external <- !(comp_id %in% names(dds_model_comp[[dataset_id]][[model_id]]))
      if (is_external) {
        this_dds <- dds_model_comp[[dataset_id]][[model_id]][[1]]
        is_diff <- row.names(this_dds) %in% external_list$ID[external_list$Comparison==comp_id]
      } else {
        this_dds <- dds_model_comp[[dataset_id]][[model_id]][[comp_id]]
        is_diff <- grepl("^.*\\*", mcols(this_dds)$results$class)
      }
      plot_formulae <- section_chooser(metadata(this_dds)$model, "differential_profile_plots", "profile_plots")
      if (is_null(plot_formulae) || !any(is_diff)) next
      exists_profile_plot <- TRUE
      partials <- partialise.matrix(
        assay(this_dds[is_diff,,drop=FALSE], ifelse(inherits(this_dds, "DESeqDataSeq"),"vst", 1)),
        as.data.frame(colData(this_dds)),
        fml=metadata(this_dds)$model$design)
      for (plot_fml in plot_formulae) {
        cdata_df <- as.data.frame(colData(this_dds)[, all.vars(plot_fml), drop=FALSE])
        mat <- t(scale(t(
          assemble_partialised(partials, update(plot_fml, NULL ~ .), resids=TRUE)
        )))
        removed <- paste(
          translate_terms(dropped_terms(partials, update(plot_fml, NULL ~ .)),this_dds),
          sep=", ", collapse=", ")
        if (length(removed)==0 || removed=="") {
          removed <- ""
        } else {
          removed <- paste0(" having removed effect of ", removed)
        }
        hc <- param$get("gene_clust")
        if ("cut.params" %in% slotNames(hc) && "k" %in% names(slot(hc, "cut.params")) && slot(hc, "cut.params")$k >= nrow(mat)) {
          slot(hc, "cut.params") <- modifyList(slot(hc, "cut.params"), list(k=ceiling(nrow(mat)/3)))
        }
        if (is_external) {
          gene_clust <- list(clusters=setNames(
            factor(external_list$Cluster[match(row.names(mat), external_list$ID)]),
            row.names(mat)))
        } else {
          if (nrow(mat)>1) {
          gene_clust <- bluster::clusterRows(mat, hc, full=TRUE)
          } else {
            gene_clust <- list(
              clusters=setNames(factor(1), row.names(mat))
            )
          }
          this_clust <- cluster_ind()
          levels(gene_clust$clusters) <- paste0(this_clust, levels(gene_clust$clusters))
        }
        cc <- cluster_calc(mat, clusterID=gene_clust$clusters, relevel=!is_external)
        centroid_df <- cbind(cc, cdata_df) %>%
          pivot_longer(cols=any_of(names(cc)), names_to="babs.name", values_to="babs.value") %>%
          mutate(babs.name=factor(babs.name, levels=names(cc)))
        pl <- ggplot(centroid_df ,
                    eval(plot_fml[[2]])) +
          facet_wrap(~ babs.name) +
          theme_bw() +
          geom_point(aes(y= babs.value)) +
          my_alpha_scale(pl) +
          labs(y=noun_to_readout(RowNoun))
        if ("group" %in% names(plot_fml[[2]])) {
          pl <- pl+geom_line(aes(y=babs.value), stat="summary", fun=mean)
        }
        if ("ScaleDiscrete" %in% class(layer_scales(pl)$x)) {
          pl <- pl + scale_x_discrete(guide=guide_axis(n.dodge=2))
        }
        capt <- paste0("Cluster profile for ", comp_id, " ",rowNoun,"s  of ", mtag(model_id), " in ", dtag(dataset_id), removed)
        do_plot(pl,
                label= "diffClustProfile",
                caption=capt
                )
        if (!is_external) {
          all_clusters[[this_clust]] <- list(
            id=gene_clust$clusters,
            description=capt
          )
        }
        cat('\n\n---\n\n')
      }
    }
  }
}
if (!exists_profile_plot) {
  cat('\n\nNo profiles were generated, either because none were specified, or those that were corresponded to comparisons with no statistically significant differential ',rowNoun,'s.\n\n', sep="")
}

      

  ## ind <- which(hclustk=="|7|=927")
  ## mcols(this_comp)$results[ind,]
  ## ind <- "ENSG00000005483"
  ## plotFrame <- colData(this_comp)
  ## plotFrame$Y <- assay(this_comp, "vst")[ind,]
  ## ggplot(as.data.frame(plotFrame),
  ##        aes(x=Ct, y=Y, colour=eventual, group=Participant)) +
  ##   geom_line()

```


# Downloads


```{r}
#| label: downloads

iwalk(xl_files,
     ~cat("\n\n<a class=\"download-excel btn btn-primary\" href=\"", attr(.x, "link"), "\"> Differential ",rowNoun," for '", .y,"'</a>", sep="")
     )

fname <- write_clusters(all_clusters,
                          mcols(ddsList[[1]]),
                          p=params,
                          prefix="differential-clusters")

cat('\n\n<a class="download-excel btn btn-primary" href="',
    attr(fname,"link"),
    '">',RowNoun,'-Cluster mappings</a>', sep="")


```


# Terms Of Use

The Crick has a [publication
policy](https://intranet.crick.ac.uk/our-crick/library-information-services/pages/guidelines-publication)
and we expect to be included on publications, regardless of funding
arrangements. Any use of these results in publication must be
discussed with BABS regarding authorship. If not authorship then
the BABS analyst must receive a named acknowledgement. Please also
cite the following sources which have enabled the analysis to be
carried out.

# References

::: {#refs}
:::

