---
title: Differential analysis
description: Differential analysis
categories: [Differential]
author: 
  - name: Gavin Kelly
    affiliations: BABS - Francis Crick Institute
format:
  html:
    theme: cerulean
    fontsize: 1.1em
    linestretch: 1.7
code-fold: true
execute:
  output: asis
  echo: false
code-summary: "Show the code"
params:

---


# Preface

We load in all the necessary additional R packages, and set up some initial parameters.


```{r}
#| label: load-02
#| output: false
#| warning: false
#| error: false

library(tidyverse)       # Language
library(devtools)
library(openxlsx)        # IO
library(ggrepel)         # General Vis.
library(RColorBrewer)
library(ComplexHeatmap)
library(gt)
library(DESeq2)          # Assay-specific
library(PoiClaClu)
library(rtracklayer)
library(tximport)
library(clusterProfiler)
library(ReactomePA)
library(GO.db)
library(IHW)
library(equatiomatic)
library(DESdemonA)

fig_caption <- captioner()
set.seed(1)


knitr::opts_chunk$set(
  warning=FALSE, error=TRUE, message=FALSE,
  dev=c("ragg_png","pdf"), out.width="100%",
  fig.width=14, fig.height=10,
  results='asis', fig.cap=expression(fig_caption())
)
knitr::knit_hooks$set(
  error = function(x, options) {
    ERROR <<- x
    knitr::knit_exit()
  },
  chunk = function(x, options){
    if(exists('ERROR')) paste0(x,'```\n',ERROR,'```\n\n**Script stopped with error**')
    else x
  })


link2path <- function(lname) {  file.path(knitr::opts_knit$get("output.dir"),lname) }

```



```{r}
#| label: tbl-load2-02
#| tbl-cap: Context
if (!isTRUE(getOption('knitr.in.progress'))) {
  my_params_yml <- "resources/01_differential.yml"
  if (file.exists(my_params_yml)) {
    params <- yaml::read_yaml(my_params_yml)
    if (is.null(names(params))) {
      params <- params[[1]]$params
    }
  } else {
    params <- list()
  }
} 


```

```{r}
#| label: read-02
data(list=paste0("counts_", params$alignment))
data(list=paste0("pca_", params$alignment, "_", params$spec))

if (!is.null(metadata(dds)$organism$org)) {
  library(metadata(dds)$organism$org, character.only=TRUE)
}
metadata(dds)$template_git <- packageDescription("DESdemonA")$git_last_commit

specs   <- DESdemonA::load_specs(file=paste0(params$spec, ".spec"), context=dds)

param <- DESdemonA::ParamList$new(defaults=specs$settings)
# Calling the setter without a value will pick up the default (from the spec), _and_ alert it in the markdown, and return the default
set.seed(param$set("seed"))
param$set("title", "{{{project}}}")
param$set("script", file.path(getwd(),"01_differential.r"))
param$set("spec",  params$spec, "Using analysis plan '{}'.")
param$set("alignment", params$alignment, "Using alignment settings '{}'")
  
```

# Find differential genes

 We use DESeq2 [@pkg_DESeq2] to find differential genes using the
negative binomial distribution to model counts, with IHW [@pkg_IHW]
for multiple-testing correction with greater power than
Benjamini-Hochberg, and ashr [@pkg_ashr] for effect-size shrinkage
to ensure reported fold-changes are more robust.


```{r}
#| label: differential-02

param$set("alpha")
param$set("lfcThreshold")
param$set("filterFun")

## For each dataset, fit all its models
dds_model_comp <- map_des(
  ddsList,
  function(x) DESdemonA::fit_models(x,
                             minReplicatesForReplace = Inf)
)


## Now put results in each 3rd level mcols(dds)$results
dds_model_comp <- map_des(
  dds_model_comp,
  DESdemonA::get_result,
  filterFun=param$get("filterFun"),
  alpha=param$get("alpha"),
  lfcThreshold=param$get("lfcThreshold")
)



dds_name <- paste0(basename(params$spec),"_x_", params$alignment)
pre_save <- dds_model_comp
dds_model_comp <- lapply(dds_model_comp,
                  function(d) {
                    lapply(d,
                           function(m) {
                             list(dds=m[[1]],
                                  comps=lapply(m, function(comp) {
                                    res <- mcols(comp)$results
                                    metadata(res) <- metadata(comp)
                                    res
                                  })
                                  )
                           })
                    })
save(dds_model_comp,
     file=file.path("data", paste0(dds_name, ".rda")),
     eval.promises=FALSE
     )
saveRDS(dds_model_comp, file=file.path("data", paste0(dds_name, ".rds")))
dds_model_comp <- pre_save

xl_files <- DESdemonA::write_results(dds_model_comp, param, dir=params$res_dir)
iwalk(xl_files,
     ~cat("\n\n<a class=\"download-excel btn btn-primary\" href=\"", sub(paste0("^", params$res_dir, "/?"), "", .x), "\"> Open Spreadsheet '", .y,"'</a>", sep="")
     )

#DESdemonA::write_all_results(dds_model_comp, dir=params$res_dir)
```


## Summary Tables {.tabset}

Here we summarise the results of the differential testing. Each
sample-set gets its own table (flick between tabs to choose which),
within which the different models, and their null hypotheses, are
enumerated.

In the 'Significant' column we tally the number of significant
genes (for pairwise comparisons, separated into up or down, where
A-B being labelled up means expression is higher in A; for omnibus
comparisons, a broad categorisation of the most extreme groups, as
described above.) We also tally the total number of genes
exhibiting that behaviour (ie not necessarily statistically
signficant) - these might not always add up to the same value, as
there is an independent filter of low-signal genes whose effect
varies from comparison to comparison.

Our naming convention for a comparison often looks like `A-B|M` -
again, because this is an semi-automated report, this mightn't be
optimal, but the rationale is `A-B` refers to the two experimental
conditions being compared (a 'upward' fold change will be reported
if the expression is higher in `A` than it is in `B`). The `|M` part
refers to the fact that we may have looked at that `A-B` comparison
conditional upon some other variable (which in this example takes the
value `M`):  anything before the `|` symbol identifies the
experimental conditions that are being contrasted, and the part after
the `|` identifies any stratification/conditioning of the data.

In @fig-summary we present some diagnostic plots of the observed
p-values. Theoretically we want the pink part of the plot to have a
solid peak on the left, and then make a reasonable plateau across
higher p-values: gross deviations from this can indicate an absence of
differential behaviour (lack of left-peak) or a model
mis-specification that might need investigating. The blue tracks
indicate p-values corresponding to very lowly expressed genes, which
would typically struggle to provide evidence of differential
behaviour, and so show an example of what a 'bad' pink trace might
look like.

```{r}
#| label: tbl-summary-02
#| tbl-cap: Summaries

summaries <- dds_model_comp %>%
  map_des(DESdemonA::summarise_results) %>%
  map_des(function(x) {bind_rows(x, .id="Comparison")}, depth="model") %>%
  map_des(function(x) {bind_rows(x, .id="Model")}, depth="dataset") %>%
  map(function(x) {
    levels(x$Group) <- sub(".*<(.+<.+<0)$", "\\1", levels(x$Group))
    levels(x$Group) <- sub("^(0<.+?<.+?)<.*$", "\\1", levels(x$Group))
    levels(x$Group) <- sub("(.*?<)(.*0.*)(<.*)", "\\10\\3", levels(x$Group))
    droplevels(subset(x, Group!="Zero Count" & Group!="Low Count"))
  }
  )

pval_frame <- dds_model_comp %>%
  map_des(~as.data.frame(mcols(.x)$results)) %>%
  map_des(function(x) {bind_rows(x, .id="Comparison")}, depth="model")
                      
                      

## for (dataset in names(summaries)) {
##   #options(htmltools.preserve.raw=TRUE)
##   #cat("### ", dataset, " \n", sep="")
##   #caption=paste0("Size of gene-lists for ", dataset),
##   summaries[[dataset]] %>%
##     gt(groupname_col="Model") %>%
##     tab_header(title="Genelist summary",
##                subtitle=dataset) %>%
##     as_raw_html()
## }


for (dataset in names(summaries)) {
  res <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      paste0("#| label: tbl-summary-",dataset),
      paste0("#| tbl-cap: ",dataset),
      "summaries[[dataset]] %>%",
      "gt(groupname_col=\"Model\") %>%",
      "tab_header(title=\"Genelist summary\", subtitle=dataset)",
      "```",
      "",
      ""
    ), envir = environment(), quiet = TRUE)
  cat(res, sep = '\n')
}


```


```{r}
#| label: fig-summary-02
  
for (dataset in names(summaries)) {
  for (model in names(pval_frame[[dataset]])) {
    pl <- ggplot(pval_frame[[dataset]][[model]], aes(x=pvalue, fill=baseMean<1)) +
      geom_density(alpha=0.5) +
      facet_wrap(~Comparison) +
      theme_bw()
    print(pl)
    fig_caption(caption=paste0("p-Diagnostic for", dataset, model))
  }
}
```

# Scatterplot of differential genes {.tabset}

Here we plot the fold-change across all genes on the vertical axis,
against their overall expression on the horizontal. Statistically
significant genes are highlighted in colour, and we attempt to
auto-label as many gene symbols as might be legible.

The fold-changes are regularised ('shrunk') to reflect their
reliability: genes with large variability between replicates are
moved towards the horizontal axis, to de-emphasise them and to
provide a more reliable prediction of behaviour in future
replications.

Again, select the dataset you wish to examine via the tabs - there
are sub-tabs for different experimental designs, if any
(e.g. removing/ignoring batch effects).


```{r}
#| label: fig-differential-MA-02

for (dataset in names(dds_model_comp)) {
  cat("## ", dataset, " {.tabset} \n", sep="") 
  for (mdl in names(dds_model_comp[[dataset]])) {
    cat("### ", mdl, "\n", sep="") 
    differential_MA(dds_model_comp[[dataset]][[mdl]], caption=fig_caption)
  }
}
```


# Differential Heatmaps {.tabset}

Here we present heatmaps of the differential genelists.  Note
that these will, by definition, divide the samples along lines
of the experimental groups, so should be interpreted differently
from the QC heatmaps which were blind to the experimental design.

```{r}
#| label: fig-differential-heatmap-02

param$set("clustering_distance_rows")
param$set("clustering_distance_columns")
for (dataset in names(dds_model_comp)) {
  cat("## ", dataset, " {.tabset} \n", sep="") 
  for (mdl in names(dds_model_comp[[dataset]])) {
    cat("### ", mdl, "\n", sep="")
    DESdemonA::differential_heatmap(
      ddsList=dds_model_comp[[dataset]][[mdl]],
      param=param$publish(),
      caption=fig_caption
    )
  }
}

```

# Terms Of Use

The Crick has a [publication
policy](https://intranet.crick.ac.uk/our-crick/library-information-services/pages/guidelines-publication)
and we expect to be included on publications, regardless of funding
arrangements. Any use of these results in publication must be
discussed with BABS regarding authorship. If not authorship then
the BABS analyst must receive a named acknowledgement. Please also
cite the following sources which have enabled the analysis to be
carried out.

# Bibliography


