---
title: Differential analysis ${title}
description: Differential analysis of ${description}
categories: [Differential]
params:
  section: Differential analysis
---


# Preface

We load in all the necessary additional R packages, and set up some initial parameters.


```{r}
#| label: load-diff1
#| output: false
#| warning: false
#| error: false

library(tidyverse)       # Language
library(devtools)
library(openxlsx)        # IO
library(ggrepel)         # General Vis.
library(RColorBrewer)
library(ComplexHeatmap)
library(gt)
library(emmeans)
library(DESeq2)          # Assay-specific
library(PoiClaClu)
library(rtracklayer)
library(tximport)
library(clusterProfiler)
library(ReactomePA)
library(GO.db)
library(IHW)
library(ggiraph)
library(ggrastr)

source("R/plot.r", local = knitr::knit_global())
source("R/markdown.r", local = knitr::knit_global())
source("R/rnaseq.r", local = knitr::knit_global())
source("R/partial.r", local = knitr::knit_global())
source("R/output.r", local = knitr::knit_global())
source("R/helpers.r", local = knitr::knit_global())

for (i in sub("\\.emmc$", "", ls("package:emmeans", pattern="*.emmc")))
  assign(paste0("my_", i,".emmc"), negate_emmc(paste0(i, ".emmc")))

set.seed(1)

if (!isTRUE(getOption('knitr.in.progress'))) {
  params <- load_params(choose_staging("02_differential"))
}

align_params <- get_align_params(params$alignment)
rowNoun <- align_params$rowNoun %||% "gene"
RowNoun <- stringr::str_to_title(rowNoun)
if ("missing" %in% names(align_params)) {
  attr(rowNoun, "readout") <- "missingness"
  attr(RowNoun, "readout") <- "Missingness"
}

do_plot <- plot_tracker(params)

knitr::opts_chunk$set(
  warning=FALSE, error=TRUE, message=FALSE,
  dev=c("ragg_png","pdf"), out.width="100%",
  dev.args=list(ragg_png=list(scaling=1.5)),
  fig.width=14, fig.height=10, fig.process=rename_with_tag(params),
  results='asis'
)



```



```{r}
#| label: tbl-load-diff2
#| tbl-cap: Context
if (isTRUE(getOption('knitr.in.progress'))) {
  gt(data.frame(Setting=names(params),
                Value=unlist(params)))
}


```

```{r}
#| label: read-diff

base_dir <- getwd()
data(list=paste0("counts_", params$alignment))
data(list=paste0("ddsList_", params$alignment, "_", params$spec))

ddsList <- lapply(ddsList, function(dds) {
  imp <- metadata(dds)$impute
  if (!is.null(imp) && !is.null(imp$differential) && !imp$differential) {
    assays(dds) <- setFirstAssay(dds, setdiff(assayNames(dds), "imputed")[1])
  }
  dds
})
  
if (!is.null(metadata(dds)$organism$org)) {
  library(metadata(dds)$organism$org, character.only=TRUE)
}


specs <- load_specs(file=paste0(params$spec, ".spec"), context=dds)

param <- ParamList$new(defaults=specs$settings)
# Calling the setter without a value will pick up the default (from the spec), _and_ alert it in the markdown, and return the default
set.seed(param$set("seed", postpone=TRUE))
param$set("spec",  params$spec, "Using analysis plan '{}'.", postpone=TRUE)
param$set("alignment", params$alignment, "Using alignment settings '{}'")

dtag <- dmc_factory(ddsList, type="dataset")
mtag <- dmc_factory(ddsList, type="model")

```

# Find differential `{r} rowNoun`s

 We use DESeq2 [@pkg_DESeq2] to find differential `{r} rowNoun`s using the
negative binomial distribution to model counts, with IHW [@IHW2016]
for multiple-testing correction with greater power than
Benjamini-Hochberg, and ashr [@pkg_ashr] for effect-size shrinkage
to ensure reported fold-changes are more robust.


```{r}
#| label: differential

param$set("alpha", postpone=TRUE)
param$set("lfcThreshold", postpone=TRUE)
param$set("filterFun", postpone=TRUE)
param$set("LRT_effect", postpone=TRUE)
param$set("clustering_distance_rows", postpone=TRUE)
param$set("clustering_distance_columns", postpone=TRUE)
param$set("gene_clust")

rds_name <- file.path("data", paste0(basename(params$spec),"_x_", params$alignment, ".rds"))

if (interactive() && file.exists(rds_name)) {
  dds_model_comp <- decompress_dmc(readRDS(rds_name))
  # In an interactive session, you may want to skip the rest of this code chunk if you are
  # sure that the saved rds object represents the latest analysis.
}


## For each dataset, fit all its models

dds_model_comp <- lapply(
  ddsList,
  function(dds) fit_models(
    dds=dds,
    param=param,
    minReplicatesForReplace = Inf
  )
)


## Now put results in each 3rd level mcols(dds)$results
dds_model_comp <- map_depth(
  dds_model_comp,
  .depth=3,
  function(dds) get_result(dds,
                         filterFun=param$get("filterFun"),
                         alpha=param$get("alpha"),
                         lfcThreshold=param$get("lfcThreshold"),
                         LRT_effect=param$get("LRT_effect", metadata(dds)$models)
                         )
)



local({
  dds_model_comp <- compress_dmc(dds_model_comp)
  saveRDS(dds_model_comp, file=rds_name)
  save(dds_model_comp, file=sub("rds$", "rda", rds_name))
})


xl_files <- write_results(dds_model_comp,
                         param,
                         params
                         )
iwalk(xl_files,
     ~cat("\n\n<a class=\"download-excel btn btn-primary\" href=\"", attr(.x, "link"), "\"> Open Spreadsheet '", .y,"'</a>", sep="")
     )

#write_all_results(dds_model_comp, dir=params$res_dir)

```


## Summary Tables 

Here we summarise the results of the differential testing. Each
sample-set gets its own table (flick between tabs to choose which),
within which the different models, and their null hypotheses, are
enumerated.

In the 'Significant' column we tally the number of significant
`{r} rowNoun`s (for pairwise comparisons, separated into up or down, where
A-B being labelled up means `{r} noun_to_readout(rowNoun)` is higher in A; for omnibus
comparisons, a broad categorisation of the most extreme groups, as
described above.) We also tally the total number of `{r} rowNoun`s
exhibiting that behaviour (ie not necessarily statistically
signficant) - these might not always add up to the same value, as
there is an independent filter of low-signal `{r} rowNoun`s whose effect
varies from comparison to comparison.

Our naming convention for a comparison often looks like `A-B|M` -
again, because this is an semi-automated report, this mightn't be
optimal, but the rationale is `A-B` refers to the two experimental
conditions being compared (a 'upward' fold change will be reported
if the `{r} noun_to_readout(rowNoun)` is higher in `A` than it is in `B`). The `|M` part
refers to the fact that we may have looked at that `A-B` comparison
conditional upon some other variable (which in this example takes the
value `M`):  anything before the `|` symbol identifies the
experimental conditions that are being contrasted, and the part after
the `|` identifies any stratification/conditioning of the data.

In @fig-summaryp-1 we present some diagnostic plots of the observed
p-values. Theoretically we want the pink part of the plot to have a
solid peak on the left, and then make a reasonable plateau across
higher p-values: gross deviations from this can indicate an absence of
differential behaviour (lack of left-peak) or a model
mis-specification that might need investigating. The blue tracks
indicate p-values corresponding to `{r} rowNoun`s with very low `{r} noun_to_readout(rowNoun)`, which
would typically struggle to provide evidence of differential
behaviour, and so show an example of what a 'bad' pink trace might
look like.

```{r}
#| label: summary

summaries <- dds_model_comp %>%
  map_dmc(summarise_results) %>%
  map_dmc(function(x) {bind_rows(x, .id="Comparison")}, depth="model") %>%
  map_dmc(function(x) {bind_rows(x, .id="Model")}, depth="dataset") %>%
  map(function(x) {
    levels(x$Group) <- sub(".*<(.+<.+<0)$", "\\1", levels(x$Group))
    levels(x$Group) <- sub("^(0<.+?<.+?)<.*$", "\\1", levels(x$Group))
    levels(x$Group) <- sub("(.*?<)(.*0.*)(<.*)", "\\10\\3", levels(x$Group))
    x
  }
  )

unclassed <- lapply(summaries, 
                    function(s) {
                      s %>%
                        subset(Group %in% c('Zero Count', 'Low Count', 'Outlier')) %>%
                        group_by(Model, Group) %>% summarise(N=paste(unique(Total), collapse=";")) %>%
                        ungroup() %>%
                        pivot_wider(id_cols=Model, names_from=Group, values_from=N)
                      })

pval_frame <- dds_model_comp %>%
  map_dmc(~as.data.frame(mcols(.x)$results)) %>%
  map_dmc(function(x) {bind_rows(x, .id="Comparison")}, depth="model")

cat("\n\n")

for (dataset_id in names(summaries)) {
  res <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      paste0("#| label: tbl-summary-",dataset_id),
      paste0("#| tbl-cap: ",dataset_id),
      "summaries[[dataset_id]] %>%",
      "subset(!Group %in% c('Zero Count', 'Low Count', 'Outlier')) %>% ",
      "gt(groupname_col=\"Model\") %>%",
      "tab_style(",
      "  style = cell_fill(color = '#bee2e9'),",
      "  locations = cells_row_groups()",
      ") %>% ",
      paste0("tab_header(title=\"",RowNoun,"List summary\", subtitle=dataset_id)"),
      "```",
      "",
      ""
    ), envir = environment(), quiet = TRUE)
  cat(res, sep = '\n\n')
  
  if (nrow(unclassed[[dataset_id]])>0) {
    res <- knitr::knit_child(
      text = c(
        "```{r}",
        "#| echo: false",
        paste0("#| label: tbl-summaryfit-",dataset_id),
        paste0("#| tbl-cap: ",dataset_id, " model-fit stats"),
        "unclassed[[dataset_id]] %>%",
        "gt() %>%",
        "tab_header(title=\"Model-fit summary\", subtitle=dataset_id)",
        "```",
        "",
        ""
      ), envir = environment(), quiet = TRUE)
    cat(res, sep = '\n\n')
  }
  for (model_id in names(pval_frame[[dataset_id]])) {
    pl <- ggplot(pval_frame[[dataset_id]][[model_id]], aes(x=pvalue, fill=baseMean<1)) +
      geom_density(alpha=0.5) +
      facet_wrap(~Comparison) +
      theme_bw()
    do_plot(pl,
            label="summaryp",
            caption=paste("p-Diagnostic for", dataset_id, model_id),
            plot_meta=list(dataset=dataset_id, model=model_id)
            )
    cat("\n\n---\n\n")
  }
}

```

```{r}
#| label: results
all_clusters <- list()
for (dataset_id in names(dds_model_comp)) {
  section_hierarchy <- list(dataset=dataset_id)
  dataset_se <- ddsList[[dataset_id]]
  met <- metadata(dataset_se)
  dmc_heading(dds_model_comp, dataset=dataset_id, model=NULL, depth=strrep("#", length(section_hierarchy)))
  has_comparisons <- sapply(met$models, function(m) "comparisons" %in% names(m) && length(m$comparisons) > 0)
  src2bullets(met, c("subset", "extra_assays", "influential_samples"))
  for (model_id in names(dds_model_comp[[dataset_id]])) {
    section_hierarchy <- list(dataset=dataset_id, model=model_id)
    this_model <- dds_model_comp[[dataset_id]][[model_id]]
    dmc_heading(ddsList, dataset=dataset_id, model=model_id, depth=strrep("#", length(section_hierarchy)))
    src2bullets(met$models[[model_id]], c("subset", "extra_assays", "influential_samples"))
    cat("`",
        Reduce(paste, sub("~", paste0(noun_to_readout(RowNoun)," ~ "), deparse(design(this_model[[1]])))),
        "`\n\n"
        )
    first <- TRUE
    external_list <- metadata(this_model[[1]])$model$external_list
    if (!is.null(external_list)) {
      external_list <- read.csv(file.path(base_dir, external_list))
      external_list <- subset(external_list, ID %in% row.names(this_model[[1]]))
      external_list$Cluster[is.na(external_list$Cluster)] <- external_list$ID[is.na(external_list$Cluster)]
    }
    for (comparison_id in c(names(this_model), unique(external_list$Comparison))) {
      cluster_ind <- counter(post_fn=to_letter)      
      section_hierarchy <- list(dataset=dataset_id, model=model_id, comparison=comparison_id)
      dmc_heading(dds_model_comp, dataset=dataset_id, model=model_id, comparison=comparison_id, depth=strrep("#", length(section_hierarchy)))
      is_external <- !(comparison_id %in% names(this_model))
      if (is_external) {
        this_dds <- this_model[[1]]
        is_significant <- row.names(this_dds) %in% external_list$ID[external_list$Comparison==comparison_id]
        comp_type <- " "
      } else {
        this_dds <- this_model[[comparison_id]]
        is_significant <- grepl("\\*$", mcols(this_dds)$results$class)
        comp_type <- "-differential "
        if (!any(is_significant)) {
          next
        }
      }
      cat('\n\n::: {.panel-tabset .unnumbered group="diffplot"}\n\n')
      md <- metadata(this_dds)
      res <- as.data.frame(mcols(this_dds)$results)
      pal <- RColorBrewer::brewer.pal(9, "Set1")
      res$class <- ifelse(is_significant, "Sig", "Not")
      if (is_external) {
        cat("\n\n", strrep("#", length(section_hierarchy)+1), " MA\n\n", sep="")
        cat("For MA plots, external gene-lists are represented in specific comparisons, rather than getting there own plot.\n\n")
      } else {        
        if (is_formula(md$comparison)) {
          yax <- "Largest fold change"
        } else {
          yax <- "Log fold change (shrunk)"
        }
        pl <- ggplot(res,
                    aes(x=baseMean,
                        y=shrunkLFC,
                        colour=ifelse(class == "Not", "#aaaaaa","#000000"),
                        size=ifelse(class == "Not", 0.5 , 2),
                        label=symbol
                        )
                    ) +
          geom_point() +
          geom_point_interactive(aes(tooltip=symbol), data=res[top_true(is_significant, res$shrunkLFC, n=100),]) +
          scale_x_log10()  + scale_size_identity() +
          labs(x="Mean of normalised counts", y=yax)+ theme_bw()
        colour_scale <- list(labels = c("Not", "Significant"),
                            breaks = c("#aaaaaa", "#000000"))
        ind <- top_true(is_significant, res$shrunkLFC, n=5)
        if (!is.null(external_list)) {
          ## For MA plots, add external IDs 
          dat <- res[external_list$ID,]
          if (length(unique(external_list$Comparison)) <= length(pal)) {
            col <- pal[match(external_list$Comparison, unique(external_list$Comparison))]
          } else {
            col <- rep(pal[1], nrow(dat))
          }
          colour_scale$labels <- c(colour_scale$labels, unique(external_list$Comparison))
          colour_scale$breaks <- c(colour_scale$breaks, pal[seq_along(unique(external_list$Comparison))])
          pl <- pl +
            ggrepel::geom_text_repel(aes(colour=col), size=3, data=res[external_list$ID,], max.overlaps=Inf, show.legend = FALSE) +
            geom_point(aes(colour=col), data=res[external_list$ID,])
          ind <- ind[!row.names(res)[ind] %in% external_list$ID]
        }
        pl <- pl+ggrepel::geom_text_repel(size=3, data=res[ind,], show.legend = FALSE) 
        pl <- pl +
          scale_colour_identity(labels=colour_scale$labels, breaks=colour_scale$breaks, guide = "legend") +
          labs(colour="Class")
        cat("\n\n", strrep("#", length(section_hierarchy)+1), " MA\n\n", sep="")
        do_plot(pl,
                label="MA",
                caption=paste0("MA plot of ", tooltip(md, comparison_id), "-differential ", rowNoun,"s for ",mtag(model_id), " in ", dtag(dataset_id)),
                preview=first,
                plot_meta=section_hierarchy,
                dpi=150, interactive=FALSE
                )
        first <- FALSE
        cat("\n\n---\n\n")
      }
      nice_legends <- separate_legend(dataset_se)
      pl <- do.call(gridExtra::arrangeGrob,c(
        nice_legends,
        ncol=ceiling(sqrt(length(nice_legends)))))
      cat("\n\n", strrep("#", length(section_hierarchy)+1), " Heatmap\n\n", sep="")
      do_plot(pl,
              label= "legend-a",
              caption=paste("Metadata colour-codings for",
                            metadata(dataset_se)$dmc$dataset_name %||% metadata(dataset_se)$dmc$dataset
                            ),
              plot_meta=section_hierarchy
              )
      cat("\n\n---\n\n")
      plot_formulae <- section_chooser(md$model, "differential_profile_plots", "profile_plots")
      model_vars <- sort_vars(
        unique(c(all.vars(md$model$design),
                 unique(unlist(lapply(plot_formulae, all.vars)))
                 )),
        md$model$variables)
      model_vars <- intersect(model_vars, names(colData(this_dds)))
      if ("mat" %in% names(md$model)) {
        mmat <- md$model$mat
      } else {
        mmat <- model.matrix(md$model$design, colData(this_dds))
      }
      var_roles <- list(all.vars(md$model$design))
      if (!is_external && "baseline_contrast" %in% names(attributes(md$comparison))) {
        var_roles <- c(emmeans:::.parse.by.formula(attr(md$comparison, "spec"))[c("by", "rhs")], var_roles)
        weights <- t(as.vector(attr(md$comparison, "baseline_contrast")) %*% MASS::ginv(mmat))
        baseline_str <- paste("baseline", noun_to_readout(rowNoun))
      } else {
        weights <- rep(1/nrow(mmat), times=nrow(mmat))
        baseline_str <- paste("average", noun_to_readout(rowNoun))
      }
      zweights <- setNames(rep(0, ncol(dataset_se)), colnames(dataset_se))
      zweights[colnames(this_dds)] <- weights
      centroid_df <- list()
      for (plot_id in distinct_formulae(plot_formulae)) {
        plot_fml <- plot_formulae[[plot_id]]
        if ("y" %in% names(plot_fml[[2]])) {
          tidied_se <- tidy_significant_se(dataset_se, is_significant, weights=zweights, assay=as.character(as.character(plot_fml[[2]]$y)))
        } else {
          tidied_se <- tidy_significant_se(dataset_se, is_significant, weights=zweights)
        }
        tidied_se <- reorder_samples(tidied_se,model_vars)
        assay(tidied_se) <- assay(tidied_se) / rowSds(assay(tidied_se)[,colnames(this_dds),drop=FALSE])
        partials <- partialise(tidied_se, fml=md$model$design, influence=colnames(this_dds))
        assay(tidied_se) <- assemble_partialised(partials, update(plot_fml, NULL ~ .), resids=TRUE)
        tidied_se <- scale_se(tidied_se, ind=colnames(this_dds), centre=FALSE, scale=TRUE)
        if ("cols" %in% names(plot_fml[[2]])) {
          col_split <- rlang::eval_tidy(eval(plot_fml[[2]])$cols, colDF(tidied_se))
        } else {
          col_split <- NULL
        }
        ha <- ComplexHeatmap::HeatmapAnnotation(
          df=colDF(tidied_se),
          col=metadata(colData(this_dds))$palette$Heatmap[names(colData(tidied_se))],
          show_legend=FALSE)
        colname_annot <- HeatmapAnnotation(
          labels1 = anno_text(colnames(tidied_se)),
          height=unit(1, "inches")
        )
        mat_height <- height_scaler(n=nrow(tidied_se), small=10, big=20, N=1000)
        hc <- param$get("gene_clust")
        if ("cut.params" %in% slotNames(hc) && "k" %in% names(slot(hc, "cut.params")) && slot(hc, "cut.params")$k >= nrow(tidied_se)) {
          slot(hc, "cut.params") <- modifyList(slot(hc, "cut.params"), list(k=ceiling(nrow(tidied_se)/3)))
        }
        if (nrow(tidied_se)>1) {
          gene_clust <- bluster::clusterRows(assay(tidied_se)[,colnames(this_dds), drop=FALSE], hc, full=TRUE)
          if (is_external) {
            gene_clust$clusters <- match(names(gene_clust$clusters), external_list$ID)
          }
        } else {
          gene_clust <- list(
            clusters=setNames(factor(1), row.names(tidied_se)),
            objects=list(hclust=FALSE))
        }
        if (!is_external) {
          this_clust <- cluster_ind()
          gene_clust$clusters <- cluster_cardinality(gene_clust$clusters, this_clust)
        }
        removed <- paste(
          translate_terms(dropped_terms(partials, update(plot_fml, NULL ~ .)), dataset_se),
          sep=", ", collapse=", ")
        if (length(removed)==0 || removed=="") {
          removed <- ""
        } else {
          removed <- paste0(" having removed effect of ", removed)
        }
        capt <- paste0(tooltip(md, comparison_id), comp_type, rowNoun,"s  of ", mtag(model_id), " in ", dtag(dataset_id), removed)
        pl <- ComplexHeatmap::Heatmap(
          assay(tidied_se),
          col=sym_colour(na.omit(as.vector(assay(tidied_se)[,colnames(this_dds)]))),
          heatmap_legend_param = list(direction = "horizontal" ),
          name=comparison_id,
          cluster_columns = FALSE,
          show_column_names = TRUE,
          cluster_rows = gene_clust$objects$hclust,
          column_split = col_split,
          top_annotation = ha,
          row_names_gp = gpar(fontsize = 6),
          show_row_names = nrow(tidied_se)<100)
        do_plot(pl,
                label="Heatmap",
                caption=paste0("Heatmap on ", capt,  ". White for ", baseline_str),
                plot_meta=section_hierarchy,
                height_opt=paste0(
                  "#| fig-height: ",
                  mat_height 
                )
                )
        cc <- cluster_calc(assay(tidied_se), clusterID=gene_clust$clusters)
        centroid_df[[plot_id]] <- cbind(cc, colDF(tidied_se)) %>%
          pivot_longer(cols=any_of(names(cc)), names_to=".derived", values_to=".derived.value") %>%
          mutate(.derived=factor(.derived, levels=names(cc)))
        attr(centroid_df[[plot_id]], "caption") <- capt
        if (!is_external) {
          all_clusters[[this_clust]] <- list(
            id=gene_clust$clusters,
            description=capt
          )
        }
        cat('\n\n---\n\n')
      }
      if (is_null(plot_formulae)) {
        cat(':::\n\n')
        next
      }
      cat("\n\n", strrep("#", length(section_hierarchy)+1), " Profile\n\n", sep="")
      for (plot_id in names(plot_formulae)) {
        plot_fml <- plot_formulae[[plot_id]]
        if (! "name" %in% names(attributes(plot_fml))) attr(plot_fml, "name") <- plot_id
        section_hierarchy$plot_config <- attr(plot_fml, "name")
        dmc_heading(dds_model_comp, dataset=dataset_id, model=model_id, plot=plot_fml, depth=strrep("#", length(section_hierarchy)+2), numbered=FALSE)

        mapping <- modify_mapping(eval(plot_fml[[2]]))
        pl <- ggplot(centroid_df[[plot_id]], mapping) +
          facet_wrapper(mapping, default=aes(facets = .derived), scales="free_y") + 
          theme_bw() +
          geom_point_interactive(aes(y= .derived.value), size = 3, shape = 21) +
          labs(y=noun_to_readout(RowNoun))
        pl <- pl + my_alpha_scale(pl)
        if ("group" %in% names(mapping)) {
          pl <- pl+geom_line(aes(y=.derived.value), stat="summary", fun=mean)
        }
        if ("ScaleDiscrete" %in% class(layer_scales(pl)$x)) {
          pl <- pl + scale_x_discrete(guide=guide_axis(n.dodge=2))
        }
        pl <- pl + get_colour_scales(this_dds, mapping, all.vars(eval(plot_fml[[2]])$flag))
        do_plot(pl,
                label= "diffClustProfile",
                caption=paste0("Cluster profile for ", attr(centroid_df[[plot_id]], "caption")),
                plot_meta=section_hierarchy
                )
        cat('\n\n---\n\n')
      }
      cat(':::\n\n')
    }
  }
}

      

```


# Downloads


```{r}
#| label: downloads

iwalk(xl_files,
     ~cat("\n\n<a class=\"download-excel btn btn-primary\" href=\"", attr(.x, "link"), "\"> Differential ",rowNoun," for '", .y,"'</a>", sep="")
     )

fname <- write_clusters(all_clusters,
                          mcols(ddsList[[1]]),
                          p=params,
                          prefix="differential-clusters")

cat('\n\n<a class="download-excel btn btn-primary" href="',
    attr(fname,"link"),
    '">',RowNoun,'-Cluster mappings</a>', sep="")


```


# Terms Of Use

The Crick has a [publication
policy](https://intranet.crick.ac.uk/our-crick/library-information-services/pages/guidelines-publication)
and we expect to be included on publications, regardless of funding
arrangements. Any use of these results in publication must be
discussed with BABS regarding authorship. If not authorship then
the BABS analyst must receive a named acknowledgement. Please also
cite the following sources which have enabled the analysis to be
carried out.

# References

::: {#refs}
:::

