---
title: Enrichment analysis
description: Enrichment and over-representation analysis
categories: [Enrichment]
author: 
  - name: Gavin Kelly
    affiliations: BABS - Francis Crick Institute
format:
  html:
    theme: cerulean
    fontsize: 1.1em
    linestretch: 1.7
code-fold: true
execute:
  output: asis
  echo: false
code-summary: "Show the code"
params:

---


# Preface

We load in all the necessary additional R packages, and set up some initial parameters.


```{r}
#| label: load-enrich
#| output: false
#| warning: false
#| error: false

library(tidyverse)       # Language
library(devtools)
library(openxlsx)        # IO
library(ggrepel)         # General Vis.
library(RColorBrewer)
library(ComplexHeatmap)
library(gt)
library(DESeq2)          # Assay-specific
library(clusterProfiler)
library(DOSE)
library(ReactomePA)
library(GO.db)
library(IHW)
library(DESdemonA)

fig_caption <- captioner()
set.seed(1)


knitr::opts_chunk$set(
  warning=FALSE, error=TRUE, message=FALSE,
  dev=c("ragg_png","pdf"), out.width="100%",
  fig.width=14, fig.height=10,
  results='asis', fig.cap=expression(fig_caption())
)
knitr::knit_hooks$set(
  error = function(x, options) {
    ERROR <<- x
    knitr::knit_exit()
  },
  chunk = function(x, options){
    if(exists('ERROR')) paste0(x,'```\n',ERROR,'```\n\n**Script stopped with error**')
    else x
  })
link2path <- function(lname) {  file.path(knitr::opts_knit$get("output.dir"),lname) }

```


```{r}
#| label: load2-enrich2
#| tbl-cap: Context
if (!isTRUE(getOption('knitr.in.progress'))) {
  my_params_yml <- "resources/02_enrichment.yml"
  if (file.exists(my_params_yml)) {
    params <- yaml::read_yaml(my_params_yml)
    if (is.null(names(params))) {
      params <- params[[1]]$params
    }
  } else {
    params <- list()
  }
} else {
  data.frame(
  Option=names(params),
  Value=unlist(params)
) %>%
  gt() %>%
  as_raw_html()
}

```

```{r}
#| label: read-enrich
data(list=paste0("counts_", params$alignment))
specs   <- DESdemonA::load_specs(file=paste0(params$spec, ".spec"), context=dds)
param <- DESdemonA::ParamList$new(defaults=specs$settings)

# Calling the setter without a value will pick up the default (from the spec), _and_ alert it in the markdown, and return the default
set.seed(param$set("seed"))
param$set("script", params$script)
param$set("spec",  params$spec, "Using analysis plan '{}'.")
param$set("alignment", params$alignment, "Using alignment settings '{}'")
data(list=paste0(params$spec, "_x_", params$alignment))
dds_model_comp <-  lapply(dds_model_comp,
                    function(d) {
                    lapply(d,
                           function(m) {
                             lapply(m$comps, function(comp) {
                               dds <- m$dds
                               mcols(dds)$results <- comp
                               metadata(dds) <- metadata(comp)
                               dds
                             })
                           })
                    })

or_pipe <-  . %>%
  filter(p.adjust<0.05) %>%
  mutate(effect=log(parse_ratio(GeneRatio)/parse_ratio(BgRatio))) %>%
  slice_max(abs(effect), n=100)
en_pipe <-  . %>%
  filter(p.adjust<0.05) %>%
  mutate(effect=NES) %>%
  slice_max(abs(effect), n=100)


if (!"functional" %in% names(specs)) {
  funcs=list(
    "GO Enrichment"=list(method="enrichment", source="GO", group= ~ dataset + model, pipe=en_pipe),
    "GO Over-representation"=list(method="or", source="GO", group= ~ dataset + model, pipe=or_pipe),
    "Reactome Enrichment"=list(method="enrichment", source="Reactome", group= ~ dataset + model, pipe=en_pipe),
    "Reactome Over-representation"=list(method="or", source="Reactome", group= ~ dataset + model, pipe=or_pipe)
  )
} else {
  funcs <- specs$functional
}



tbl_child <- r"(```{r}
#| label: tbl-{{lbl}}
kable({{tbl}})
```)"

fig_child <- r"(```{r}
#| label: fig-{{lbl}}
#| fig-cap: {{capt}}
print({{pl}})
fig_caption("{{capt}}")
```)"


```


# Functional Analysis

Here we look at which [Reactome](https://reactome.org/) and/or [GO molecular functions](http://geneontology.org/)
are over-represented/enriched in the various genelists we have created.



```{r}
#| label: functional

param$set("showCategory")

dmc2df <- function(dmc) {
  dds <- do.call(c, lapply(dmc, function(d) {do.call(c, lapply(d, function(m) {m}))}))
  list2DF(
    list(dataset=sapply(dds, function(x) metadata(x)$dmc$dataset),
    model=sapply(dds, function(x) metadata(x)$dmc$model),
    comparison=sapply(dds, function(x) metadata(x)$dmc$comparison),
    dds=dds
    )
  )
}


dds_df <- tibble::as_tibble(dmc2df(dds_model_comp))


functional <- function (dds, method, source,  ...) {
  org <- metadata(dds)$organism$org
  res <- mcols(dds)$results
  ind <- grepl("\\*", res$class)
  if (method!="or") ind=T
  res <- mcols(dds)$results[ind & !is.na(res$entrez),]
  genes <- setNames(res$shrunkLFC, res$entrez)
  genes <- sort(genes, decreasing=TRUE)
  if (length(genes) < 1) return(NULL)
  reactome_species_list <- c(
    anopheles = "org.Ag.eg.db", arabidopsis = "org.At.tair.db", 
    bovine = "org.Bt.eg.db", canine = "org.Cf.eg.db", celegans = "org.Ce.eg.db", 
    chicken = "org.Gg.eg.db", chimp = "org.Pt.eg.db", coelicolor = "org.Sco.eg.db", 
    ecolik12 = "org.EcK12.eg.db", ecsakai = "org.EcSakai.eg.db", 
    fly = "org.Dm.eg.db", gondii = "org.Tgondii.eg.db", human = "org.Hs.eg.db", 
    malaria = "org.Pf.plasmo.db", mouse = "org.Mm.eg.db", pig = "org.Ss.eg.db", 
    rat = "org.Rn.eg.db", rhesus = "org.Mmu.eg.db", xenopus = "org.Xl.eg.db", 
    yeast = "org.Sc.sgd.db", zebrafish = "org.Dr.eg.db")
  reactome_species <- names(reactome_species_list)[reactome_species_list==org]
  yy <- switch(
    method,
    or=switch(
      source,
      GO=enrichGO(gene=names(genes), OrgDb = metadata(dds)$organism$org, ...),
      Reactome=enrichPathway(gene=names(genes), organism=reactome_species, ...)
    ),
    enrichment=switch(
      source,
      GO=gseGO(geneList=genes, OrgDb = metadata(dds)$organism$org, ...),
      Reactome=gsePathway(geneList=genes, organism=reactome_species, ...)
    )
  )
}
  

func_df <- cross_join(
  tibble(func_name=names(funcs), func_params=funcs),
  dds_df)
func_df <- rowwise(func_df) %>%
  mutate(func_res=list(do.call(functional, modifyList(func_params, list(dds=dds, group=NULL, pipe=NULL))))) %>%
  filter(!is.null(func_res) && nrow(func_res)!=0) %>%
  ungroup()

func_res_list <- lapply(
  unique(func_df$func_name) |> (function(x) setNames(x,x))(),
  function(f) {
    grp_vars <- all.vars(funcs[[f]]$group)
    within_vars <- setdiff(c("dataset","model","comparison"), grp_vars)
    func_df %>%
      filter(func_name==f) %>%
      unite("inner", all_of(within_vars)) %>%
      group_by(across(all_of(grp_vars))) %>%
      mutate(df=map2(func_res, inner, ~data.frame(as.data.frame(.x), inner=.y))) %>%
      summarise(df=list(map_dfr(df, identity)))
  })

sym_colour <- function(dat, lo="blue",zero="white", hi="red") {
  mx <- quantile(abs(dat), 0.9)
  circlize::colorRamp2(c(-mx, 0, mx), colors=c(lo, zero, hi))
}

for (i in names(func_res_list)) {
  cat("\n\n## ", i, "\n\n")
  grp <- unite(func_res_list[[i]], "grp", -df, sep=" ") %>% pull(grp)
  for (j in 1:nrow(func_res_list[[i]])) {
    df <-       func_res_list[[i]]$df[[j]] %>%
      funcs[[i]]$pipe() %>%
               complete(inner, ID)
    ord <- group_by(df,ID) %>%
      dplyr::select(inner, effect, ID) %>%
      mutate(na=is.na(effect) * 2^match(inner, unique(inner))) %>%
      summarise(na=sum(na), mean=mean(effect, na.rm=TRUE))
    ord <- order(ord$na, ord$mean)
    df$ID <- factor(df$ID, levels=df$ID[ord])
    capt <- paste(i, grp[j], sep=" - ")
    pl <- ggplot(
      df,
      aes(x=inner, y=ID, fill=effect)) +
      geom_tile() +
      labs(x="Comparison", y="Annotation", title=capt) +
      scale_fill_distiller(type="div", palette="RdYlBu") +
      scale_y_discrete(expand=c(0, 0))+
      scale_x_discrete(expand=c(0, 0)) +
      theme_grey() +
      theme(
        axis.ticks=element_line(size=0.4),
        plot.background=element_blank(),
        panel.border=element_blank()
      )
    cat(knitr::knit_child(text=knitr::knit_expand(text=fig_child, pl="pl", lbl=gsub("[^[:alnum:]]+", "-", capt), capt=capt)))
  }
}


```

# Terms Of Use

The Crick has a [publication
policy](https://intranet.crick.ac.uk/our-crick/library-information-services/pages/guidelines-publication)
and we expect to be included on publications, regardless of funding
arrangements. Any use of these results in publication must be
discussed with BABS regarding authorship. If not authorship then
the BABS analyst must receive a named acknowledgement. Please also
cite the following sources which have enabled the analysis to be
carried out.

# Bibliography


