---
title: Enrichment analysis
description: Enrichment and over-representation analysis
categories: [Enrichment]
author: 
  - name: Gavin Kelly
    affiliations: BABS - Francis Crick Institute
format:
  html:
    theme: cerulean
    fontsize: 1.1em
    linestretch: 1.7
code-fold: true
execute:
  output: asis
  echo: false
  error: true
code-summary: "Show the code"
params:

---


# Preface

We load in all the necessary additional R packages, and set up some initial parameters.


```{r}
#| label: load-enrich
#| output: false
#| warning: false
#| error: false

library(tidyverse)       # Language
library(devtools)
library(openxlsx)        # IO
library(ggrepel)         # General Vis.
library(RColorBrewer)
library(ComplexHeatmap)
library(gt)
library(DESeq2)          # Assay-specific
library(clusterProfiler)
library(DOSE)
library(ReactomePA)
library(GO.db)


source("R/plot.r", local = knitr::knit_global())
source("R/markdown.r", local = knitr::knit_global())
source("R/rnaseq.r", local = knitr::knit_global())
source("R/output.r", local = knitr::knit_global())
source("R/helpers.r", local = knitr::knit_global())
source("R/enrichment.r", local = knitr::knit_global())

fig_caption <- captioner()
fig_label <- counter()
tbl_label <- counter()

set.seed(1)


knitr::opts_chunk$set(
  warning=FALSE, error=TRUE, message=FALSE,
  dev=c("ragg_png","pdf"), out.width="100%",
  dev.args=list(ragg_png=list(scaling=1.5)),
  fig.width=14, fig.height=10,
  results='asis',
  fig.cap=expression(fig_caption())
)
link2path <- function(lname) {  file.path(knitr::opts_knit$get("output.dir"),lname) }

```


```{r}
#| label: load2-enrich2
if (!isTRUE(getOption('knitr.in.progress'))) {
  params <- load_params("resources/03_enrichment", extra=TRUE)
} else {
  gt(data.frame(Setting=names(params),
                Value=unlist(params)))
}

```

```{r}
#| label: read-enrich
data(list=paste0("counts_", params$alignment))
specs   <- load_specs(file=paste0(params$spec, ".spec"), context=dds)
param <- ParamList$new(defaults=specs$settings)

# Calling the setter without a value will pick up the default (from the spec), _and_ alert it in the markdown, and return the default
set.seed(param$set("seed"))
param$set("script", params$script)
param$set("spec",  params$spec, "Using analysis plan '{}'.")
param$set("alignment", params$alignment, "Using alignment settings '{}'")
data(list=paste0(params$spec, "_x_", params$alignment))
dds_model_comp <-  lapply(dds_model_comp,
                    function(d) {
                    lapply(d,
                           function(m) {
                             lapply(m$comps, function(comp) {
                               dds <- m$dds
                               mcols(dds)$results <- comp
                               metadata(dds) <- metadata(comp)
                               dds
                             })
                           })
                    })



if (!"functional_approaches" %in% names(specs)) {
  approaches=list(
    "GO Enrichment"=list(
      method="enrichment", # that, or `OR`
      source="GO",         # that, or `Reactome`
      ont="BP",            # can feed other args that get used by the relevant clusterProfiler call
      post_process=default_post_process(alpha=0.05, n=100, threshold=-Inf)), # args can be set explicitly (these are the defaults)
    "GO Over-representation"=list(
      method="OR",
      source="GO",
      ont="BP",
      post_process=default_post_process()), # or args can be left to default
    "Reactome Enrichment"=list(
      method="enrichment",
      source="Reactome"), # omitting will also use defaults. To do nothing, `post_process=identity`
    "Reactome Over-representation"=list(
      method="OR",
      source="Reactome")
  )
} else {
  approaches <- specs$functional_approaches
}



```


# Functional Analysis

Here we look at which [Reactome](https://reactome.org/) and/or [GO molecular functions](http://geneontology.org/)
are over-represented/enriched in the various genelists we have created.



```{r}
#| label: functional

#param$set("showCategory")

# Get a data.frame with the dds objects in a column
dds_df <- tibble::as_tibble(dmc2df(dds_model_comp))
  
# Enumerate all functional-approach x dds combos
approach_df <- cross_join(
  tibble(approach_name=names(approaches), approach_params=approaches),
  dds_df)

# Apply the functional approach to the dds
approach_df <- rowwise(approach_df) %>%
  mutate(
    approach_res=list(do.call(
      functional_api,
      modifyList(
        approach_params,
        list(dds=dds, post_process=NULL)) # don't pass post_process to the api
    )
    )) %>%
  filter(!is.null(approach_res) && nrow(approach_res)!=0) %>%
  ungroup()

# Gather results into common functional approaches, a list of
# dataframes where each element corresponds to an approach, which
# is grouped by the grouping structure (e.g. model within dataset),
# and the rows in a group are the comparisons.
grouper <- ~ dataset + model
approach_res_list <- lapply(
  unique(approach_df$approach_name) |> (function(x) setNames(x,x))(),
  function(f) {
    grp_vars <- all.vars(grouper)
    within_vars <- setdiff(c("dataset","model","comparison"), grp_vars)
    approach_df %>%
      filter(approach_name==f) %>%
      unite("inner", all_of(within_vars)) %>%
      group_by(across(all_of(grp_vars))) %>%
      mutate(df=map2(approach_res, inner, ~data.frame(as.data.frame(.x), inner=.y))) %>%
      summarise(df=list(map_dfr(df, identity)))
  })



## Speculative - recursively structure the document
if (do_approach_df_recurse <- FALSE) {
  struct <- list(method=list(source=list("dataset","model")))
  destruct(
    approach_df, struct,
    function(df, group) {
      approach <- unique(df$approach_name)
    })
}


## Put the results into the report
# For each approach produce a section,
# for each group of comparisons within an
# approach, produce a plot.

for (approach in names(approach_res_list)) {
  cat("\n\n## ", approach, "\n\n")
  groups <- approach_res_list[[approach]] %>%
    unite("grp", -df, sep=" ") %>%
    pull(grp)
  if ("post_process" %in% names(approaches[[approach]])) { # post_process allows filtering on a case-by-case basis
    if (is.list(approaches[[approach]]$post_process)) {# chose the method-specific post-processing
      my_post_process <- approaches[[approach]]$post_process[[approaches[[approach]]$method]]
    } else { # post_process can also be given as a single function
      my_post_process <- approaches[[approach]]$post_process
    }
  } else { # or drop back to the default if not specified.
    my_post_process <- default_post_process(method=approaches[[approach]]$method)
  }
  biggest_effect <- max(abs(sapply(
    approach_res_list[[approach]]$df,
    function(x) range(my_post_process(x)$effect, na.rm=TRUE))
    ))
  effect_label <- list(OR="Relative effect", enrichment="NES")[[approaches[[approach]]$method]]
  for (grp in 1:nrow(approach_res_list[[approach]])) {
    df <- approach_res_list[[approach]]$df[[grp]] %>%
      my_post_process() %>%
      tidyr::complete(inner, ID)
    df$ID <- factor(df$ID, levels=df$ID[order_enrichments(df)])
    capt <- paste(approach, groups[grp], sep=" - ")
    size <- NULL#fall-back if post-processing doesn't define a 'size'
    pl <- ggplot(df, aes(x=inner, y=Description, colour=effect, size=size)) +
      geom_point() +
      labs(x="Comparison", y="Annotation", title=capt, colour=effect_label, size="Proportion of differential") +
      scale_colour_distiller(type="div", palette="RdYlBu", limits=c(-1,1)* biggest_effect) +
      scale_y_discrete(labels = function(x) stringr::str_wrap(x, width = 40))+
      scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 10)) +
      theme_bw() +
      theme(
        axis.ticks=element_line(size=0.4),
        plot.background=element_blank(),
        panel.border=element_blank()
      ) +
      theme(legend.position="bottom", legend.box = "horizontal")
    do_plot(pl,
            label=fig_label("functional"),
            caption=capt,
            height_mult=nlevels(df$ID)/30
            )
    tbl <- df %>%
      rename(Comparison=inner) %>%
      filter(!is.na(Description)) %>%
      mutate(core_enrichment=NULL) %>%
      gt() %>%
      data_color(
        columns=effect,
        reverse=TRUE,
        palette="RdYlBu",
        domain=c(-1,1) * biggest_effect
      ) %>%
    tab_header(title=capt) %>%
    opt_interactive()
    do_tbl(tbl,
           label=tbl_label("functional"),
           caption=capt,
           path=params$staging_dir)
  }
}

```

# Terms Of Use

The Crick has a [publication
policy](https://intranet.crick.ac.uk/our-crick/library-information-services/pages/guidelines-publication)
and we expect to be included on publications, regardless of funding
arrangements. Any use of these results in publication must be
discussed with BABS regarding authorship. If not authorship then
the BABS analyst must receive a named acknowledgement. Please also
cite the following sources which have enabled the analysis to be
carried out.

# Bibliography



