---
title: Gene-level plots
description: Plots of expression profiles
categories: [Differential]
author: 
  - name: Gavin Kelly
    affiliations: BABS - Francis Crick Institute
format:
  html:
    theme: cerulean
    fontsize: 1.1em
    linestretch: 1.7
code-fold: true
execute:
  output: asis
  echo: false
code-summary: "Show the code"
params:
  section: Gene plots

---


# Preface

We load in all the necessary additional R packages, and set up some initial parameters.


```{r}
#| label: load-diff1
#| output: false
#| warning: false
#| error: false

library(tidyverse)       # Language
library(devtools)
library(openxlsx)        # IO
library(ggrepel)         # General Vis.
library(RColorBrewer)
library(ComplexHeatmap)
library(gt)
library(DESeq2)          # Assay-specific
library(PoiClaClu)
library(rtracklayer)
library(tximport)
library(clusterProfiler)
library(ReactomePA)
library(GO.db)
library(IHW)

source("R/plot.r", local = knitr::knit_global())
source("R/markdown.r", local = knitr::knit_global())
source("R/rnaseq.r", local = knitr::knit_global())
source("R/partial.r", local = knitr::knit_global())
source("R/output.r", local = knitr::knit_global())
source("R/helpers.r", local = knitr::knit_global())

set.seed(1)

if (!isTRUE(getOption('knitr.in.progress'))) {
  params <- load_params(choose_staging("04_gene_plots"))
}

do_plot <- plot_tracker(params)

knitr::opts_chunk$set(
  warning=FALSE, error=TRUE, message=FALSE,
  dev=c("ragg_png","pdf"), out.width="100%",
  dev.args=list(ragg_png=list(scaling=1.5)),
  fig.width=14, fig.height=10, fig.process=rename_with_tag(params),
  results='asis'
)


link2path <- function(lname) {  file.path(knitr::opts_knit$get("output.dir"),lname) }

```




```{r}
#| label: read-diff
data(list=paste0("counts_", params$alignment))
data(list=paste0("ddsList_", params$alignment, "_", params$spec))

if (!is.null(metadata(dds)$organism$org)) {
  library(metadata(dds)$organism$org, character.only=TRUE)
}


specs <- load_specs(file=paste0(params$spec, ".spec"), context=dds)

param <- ParamList$new(defaults=specs$settings)
# Calling the setter without a value will pick up the default (from the spec), _and_ alert it in the markdown, and return the default
set.seed(param$set("seed", postpone=TRUE))
param$set("spec",  params$spec, "Using analysis plan '{}'.", postpone=TRUE)
param$set("alignment", params$alignment, "Using alignment settings '{}'")

```


# Find differential genes

 We use DESeq2 [@pkg_DESeq2] to find differential genes using the
negative binomial distribution to model counts, with IHW [@IHW2016]
for multiple-testing correction with greater power than
Benjamini-Hochberg, and ashr [@pkg_ashr] for effect-size shrinkage
to ensure reported fold-changes are more robust.


```{r}
#| label: differential

param$set("alpha", postpone=TRUE)
param$set("lfcThreshold", postpone=TRUE)
param$set("filterFun", postpone=TRUE)
param$set("baseline_heuristic", postpone=TRUE)
param$set("LRT_effect")

rds_name <- file.path("data", paste0(basename(params$spec),"_x_", params$alignment, ".rds"))


dds_model_comp <- decompress_dmc(readRDS(rds_name))

```



```{r}
#| label: fig-differential-MA
first <- TRUE



for (dataset in names(dds_model_comp)) {
  dmc_heading(dds_model_comp, dataset=dataset, model=NULL, depth="##")
  met <- metadata(dds_model_comp[[dataset]][[1]][[1]])$dmc
  for (mdl in names(dds_model_comp[[dataset]])) {
    met <- metadata(dds_model_comp[[dataset]][[mdl]][[1]])$dmc
    dmc_heading(dds_model_comp, dataset=dataset, model=mdl, depth="###")
    this_mdl <- dds_model_comp[[dataset]][[mdl]]
    for (comp_name in names(this_mdl)) {
      dds <- this_mdl[[comp_name]]
      res <- as.data.frame(mcols(this_mdl[[comp_name]])$results)
      md <- metadata(this_mdl[[comp_name]])
      filt <- expression(grepl('.*\\*$', class))
      genes <- row.names(res)[eval(filt,res)][1:5]
      plotFrames <- lapply(genes, function(i) cbind(as.data.frame(colData(dds)), Y=assay(dds, "vst")[i,], .gene=i))
      plotFrame <- do.call(rbind, plotFrames)
      my_aes <- aes(x=days, col=eventual, group=interaction(eventual, Participant)) ~ 1
      pl <- ggplot(plotFrame,
                  eval(my_aes[[2]])) +
        geom_point(aes(y=Y)) + facet_wrap(~.gene) +
        theme_bw()
      if ("group" %in% names(my_aes[[2]])) {
        pl <- pl + geom_line(aes(y=Y))
      }
      do_plot(pl,
              label="02-MA",
              caption=paste0("MA plot of differential genes ", sub(".*\\t", "", comp_name)),
              preview=first
              )
      first <- FALSE
      cat("\n\n---\n\n")
    } # next comparison
  }   # next model
}     # next dataset

```


# Terms Of Use

The Crick has a [publication
policy](https://intranet.crick.ac.uk/our-crick/library-information-services/pages/guidelines-publication)
and we expect to be included on publications, regardless of funding
arrangements. Any use of these results in publication must be
discussed with BABS regarding authorship. If not authorship then
the BABS analyst must receive a named acknowledgement. Please also
cite the following sources which have enabled the analysis to be
carried out.

# References

::: {#refs}
:::

